### 2021-05-12 paper/main.tex
Initial template for the paper

\documentclass[sigplan,screen]{acmart}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2021}
\acmYear{2021}
\acmDOI{tbd}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Haskell '21]{the 14th ACM SIGPLAN International Haskell Symposium}{August 26--27, 2021}{Virtual Event}
%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%  June 03--05, 2018, Woodstock, NY}
%\acmPrice{15.00}
%\acmISBN{978-1-4503-XXXX-X/18/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Complexity theory for all! - Using LiquidHaskell to prove complexity of data structures (Functional Pearl)}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Jan van Brügge}
\email{jan.van-bruegge@tum.de}
%\orcid{1234-5678-9012}
\affiliation{%
  \institution{Technical University Munich}
  \streetaddress{Arcisstraße 21}
  \city{Munich}
  \state{Bavaria}
  \country{Germany}
  \postcode{80333}
}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Trovato and Tobin, et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}

\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003809.10010031</concept_id>
       <concept_desc>Theory of computation~Data structures design and analysis</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
       <concept_desc>Software and its engineering~Functional languages</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10010940.10010992.10010998.10010999</concept_id>
       <concept_desc>Software and its engineering~Software verification</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Data structures design and analysis}
\ccsdesc[500]{Software and its engineering~Functional languages}
\ccsdesc[500]{Software and its engineering~Software verification}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{datastructures, LiquidHaskell, amortized complexity, finger trees, theorem proving, functional pearl}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}

\section{Amortized complexity}

\section{LiquidHaskell for theorem proving}

\section{Case studies}

\subsection{Stack with Multipop}

\subsection{Finger Tree}

\section{Limitations of this approach}

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
%\begin{acks}
%\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\end{document}
\endinput

### 2021-05-15 paper/main.tex
Add outlines of all sections

@@
 %% complete the rights form.
-\setcopyright{acmcopyright}
+\setcopyright{acmlicensed}
 \copyrightyear{2021}
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Complexity theory for all! - Using LiquidHaskell to prove complexity of data structures (Functional Pearl)}
+\title{Liquid Amortization - Proving amortized complexity with LiquidHaskell (Functional Pearl)}
 
@@
 \begin{abstract}
-
+TODO
 \end{abstract}
@@
 
+Motivation:
+\begin{itemize}
+\item{Formal proofs for properties of data structures are getting more common}
+\item{Most of the time proof is done in other language}
+\item{Needs a translation step, either manual or programmatic}
+\item{Might have errors in translation}
+\item{Might not support all features of implementation language}
+\item{Thus needs encoding}
+\item{Hard to reconnect proofs back to the actual implementation}
+\end{itemize}
+
+Solution:
+\begin{itemize}
+\item{LiquidHaskell \cite{liquidhaskell} is a refinement type \cite{refinement_types} checker for Haskell}
+\item{Can be used as a theorem prover \cite{tpfa}}
+\item{Directly provide proofs in Haskell, skipping the translation step}
+\end{itemize}
+
+Contributions:
+\begin{itemize}
+\item{Explain how amortized complexity works}
+\item{How to prove amortized complexity}
+\item{Show with case studies of core Haskell data structures how to apply this}
+\end{itemize}
+
 \section{Amortized complexity}
+
+\begin{itemize}
+\item{Worst case is often too pessimistic}
+\item{Idea: spread out expensive operations over multiple cheap ones}
+\item{Usually only talks about sequence of operations without sharing}
+\item{Formalization: Physicist's method}
+\end{itemize}
 
@@
 
+\begin{itemize}
+\item{Uses refinement types}
+\item{Refinments are solved by external SMT solver (z3)}
+\item{With the refinement one can define proof combinators}
+\item{Use equational reasoning and induction to prove interesing properties}
+\end{itemize}
+
 \section{Case studies}
+
+\begin{itemize}
+\item{encode requirements of potential as type synonym}
+\end{itemize}
 
@@
 
+\begin{itemize}
+\item{definition of data structure and operations}
+\item{worst case of push and multipop is O(1) and O(n)}
+\item{but every item we pop was pushed at some point}
+\item{define potential as height of stack}
+\item{refine all functions to the logic level \cite{reflection}}
+\item{guess constant factor}
+\item{prove base case and adjust constant}
+\item{use induction for complex case}
+\end{itemize}
+
 \subsection{Finger Tree}
 
+\begin{itemize}
+\item{originally described in \cite{fingertrees}}
+\item{definition of finger tree}
+\item{code is copy pased from the containers package}
+\item{in contrast to \cite{fingertrees_new} four constructors and different append function}
+\item{slightly modified because LiquidHaskell does not work great with type classes}
+\item{mention polymorphic recursion that is not supported in Lean \cite{lean} and Isabelle/HOL \cite{isabelle}}
+\item{potential of a finger tree}
+\end{itemize}
+
+\subsubsection{Cons and Snoc}
+
+\begin{itemize}
+\item{cons and snoc are very similar, see snoc code in appendix}
+\item{definition of the cons function}
+\item{definition of the consT function}
+\item{step by step proof of amortized complexity}
+\end{itemize}
+
+\subsubsection{Append}\label{sec:append}
+
+\begin{itemize}
+\item{same as worst case, but potential may differ}
+\item{definition of append as mutually recursive functions}
+\item{no automatic termination metric, track termination across recursion}
+\item{generates a lot of constraints > 3000, takes long to typecheck}
+\item{defining the timing function would generate over 8000 constraints}
+\item{proving with logarithms is not easy, but potentiation makes it easier}
+\item{sadly not really possible to do as every step takes forever to typecheck}
+\end{itemize}
+
 \section{Limitations of this approach}
+
+\subsection{Definition of the timing function}
+
+\begin{itemize}
+\item{without changing the system under test, we have to manually create a timing function}
+\item{manual approach might lead to errors}
+\item{Idea 1: factor out recursion and use same recursion for implementation and timing}
+\item{Idea 2: use template Haskell to generate timing function}
+\end{itemize}
+
+\subsection{LiquidHaskell for verification}
+
+\begin{itemize}
+\item{support for type classes is not great at the moment}
+\item{as seen in section~\ref{sec:append}, more complicated pattern matches lead to combinatorial explosion}
+\item{might be possible to be smarter when generating constraints}
+\item{no direct support for rationals make non-linear proofs involving division harder}
+\item{several performance related options like --no-termination, --diff, --fast, did not make a difference for type checking}
+\end{itemize}
+
+\section{Conclusion}
+
+\begin{itemize}
+\item{Get more confidence that your implementation matches the theoretical expectations}
+\item{Easy to apply for simple complexities like O(1) or O(n)}
+\item{Harder with O(log n) because the SMT prover/LiquidHaskell is not quite there yet}
+\end{itemize}
 

### 2021-05-17 paper/main.tex
Write most of the paper

--- 2021-05-12 paper/main.tex
@@
 \documentclass[sigplan,screen]{acmart}
+
+\usepackage{xcolor}
+\definecolor{codegreen}{rgb}{0,0.6,0}
+
+\usepackage{listings}
+\lstset{%
+    language=Haskell,
+    columns=fullflexible,
+    basicstyle=\ttfamily,
+    commentstyle=\color{codegreen}
+}
+
+% shorthand for big O
+\renewcommand\O[1]{$\mathcal{O}(#1)$}
 
@@
 
-\section{Amortized complexity}
+Formally proving certain properties of production code is getting more and more common.
+For either safety critical code that is not allowed to "go wrong" or code that is fundamental and reused very often, bugs can have a widespread impact.
+Proving that the code behaves according to the specification can give an extra layer of confidence in such cases.
+Fundamental data structures are one of the most important parts of a language ecosystem and a lot of production code depends on them.
+In the Haskell ecosystem, this job is done by the \textit{containers} package~\cite{containers} which has over 5300 reverse dependencies.
+A lot of code is depending on the assumptions that the data structures perform well and adhere to the stated computational complexities from the documentation.
+Thus, proving those assumptions correct benefits a huge number of people.
+
+Usually these proofs are done in an external theorem prover, requiring an explicit translation step from the production language to the language of the theorem prover.
+This translation step has a few disadvantages.
+First, the semantics of the target language may differ from the semantics of the production language.
+This is especially true for Haskell with its pervasive laziness, compared with most theorem provers that operate with a strict language.
+Second, such a translation step may introduce subtle differences or bugs, either by accident in the manual case or through a bug in the automated case.
+Lastly, the target language may not support certain features present in the source language such as polymorphic recursion in data types.
+In those cases, the data structure needs to be encoded in the target language, again opening the doors for small and subtle errors.
+Because the code that is actually used to conduct the proofs might me far removed from the source code that is running in production, it is often hard to reconnect the proofs back to the source code.
+
+A different approach comes in the form of LiquidHaskell~\cite{liquidhaskell}.
+It is a refinement type~\cite{refinement_types} checker for Haskell that since 2020 is implemented as a type checker plugin for GHC, the main Haskell compiler used in the industry.
+It allows to specify additional constraints on types like "integers greater than zero" instead of just "integers".
+These constraints are propagated through the code and checked by an external SMT prover.
+With a handful of proof combinators defined by LiquidHaskell, it is possible to use Haskell directly as a theorem prover, enabling intuitive equational reasoning within the language~\cite{tpfa}.
+This avoids any translation step and makes it easy to keep proof and implementation in sync, as any failure to do so will result in an immediate compiler error.
+Another benefit is that the reader or writer of a proof does not need to learn a new language, the production language is the proof language.
+
+This paper thus:
+\begin{itemize}
+\item{Assumes no prior knowledge of amortized complexity by recapping the theoretical background}
+\item{Shows the reasoning infrastructure that is needed to prove amortized complexity with LiquidHaskell}
+\item{Applies this technique to the finger tree, the data structure behind \texttt{Data.Seq} from \textit{containers}, as a case study}
+\item{Discusses the usefulness of LiquidHaskell for such proofs compared with other theorem provers}
+\end{itemize}
+
+\section{Amortized complexity}\label{sec:complexity}
+
+\begin{itemize}
+\item{Worst case is often too pessimistic}
+\item{Idea: spread out expensive operations over multiple cheap ones}
+\item{Usually only talks about sequence of operations without sharing}
+\item{Formalization: Physicist's method}
+\end{itemize}
 
 \section{LiquidHaskell for theorem proving}
+
+As described by Vazou et. al.~\cite{tpfa}, one can use LiquidHaskell to do complex equational reasoning that is machine checked by the compiler.
+
+\begin{itemize}
+\item{Uses refinement types}
+\item{Refinments are solved by external SMT solver (z3)}
+\item{With the refinement one can define proof combinators}
+\item{Use equational reasoning and induction to prove interesing properties}
+\end{itemize}
 
@@
 
+One of the simplest data structures that have an amortized complexity is a stack with two operations: Push and Multipop.
+$push(e)$ adds a singular element $e$ on top of the stack, while $multipop(k)$ removes the top $k$ elements.
+This data structure and its operations could be defined like this in Haskell.
+We already add some refinements to assert that $k$ is not negative.
+
+\begin{lstlisting}
+data Stack a = Empty | Elem a (Stack a)
+
+push :: a -> Stack a -> Stack a
+push = Elem
+
+{-@ multipop :: Nat -> Stack a -> Stack a @-}
+multipop :: Int -> Stack a -> Stack a
+multipop _ Empty = Empty
+multipop 0 s = s
+multipop n (Elem _ s) = multipop (n - 1) s
+\end{lstlisting}
+
+If we look at this definition we can easily see that \texttt{push} takes constant time.
+No matter how large the rest of the stack is, it will only allocate a new \texttt{Pushed} cell that contains a reference to the rest of the stack.
+Thus this operation has complexity \O{1}.
+
+\texttt{multipop} however may pop all of the elements in the stack if $k$ is greater or equal to $n$, where $n$ is the height of the stack.
+So in the worst case this operation takes $n$ steps and thus it is in \O{n}.
+
+This worst case however is a bit too pessimistic.
+If we pop all elements, we will not be able to do it again, as the stack is empty after the first operation.
+In fact, in order for an element to be popped, it first has to be pushed.
+So we can spread the cost for the expensive $multipop$ operation over all the constant time pushes, making \texttt{multipop} amortized \O{1}.
+With the banker's method this would mean for a push we could pay one time unit actual cost and one time unit extra for the bank.
+Then the bank has always as much time units as there are elements in the stack.
+Thus a \texttt{multipop} will not be able to make the bank account negative.
+
+To formally prove this with the physicist's method, we first need to define a potential $\Phi$ and a timing function for each of our operations.
+The intuition for the potential is what property of the data structure makes \texttt{multipop} more expensive.
+Because a stack is such a simple data structure, there is basically only one property: its height.
+So we can define a function that calculates the height of a stack and use it as our potential.
+To define the timing functions, we copy the code for the operation, replace constant time steps with $1$ and adjust the recursion.
+
+\begin{lstlisting}
+{-@ phi :: Stack a -> Nat @-}
+phi :: Stack a -> Int
+phi Empty = 0
+phi (Elem _ s) = 1 + phi s
+
+{-@ pushT :: a -> Stack a -> Nat @-}
+pushT :: a -> Stack a -> Int
+pushT _ _ = 1
+
+{-@ multipopT :: Nat -> Stack a -> Nat @-}
+multipopT :: Int -> Stack a -> Int
+multipopT _ Empty = 1
+multipopT 0 _ = 1
+multipopT n (Elem _ s) = 1 + multipopT (n - 1) s
+\end{lstlisting}
+
+To use those functions in a LiquidHaskell proof, we also need to reflect them into the logic level~\cite{reflection} by adding \texttt{reflect} annotations.
+
+\begin{lstlisting}
+{-@ reflect push @-}
+{-@ reflect pushT @-}
+{-@ reflect multipop @-}
+{-@ reflect multipopT @-}
+{-@ reflect phi @-}
+\end{lstlisting}
+
+First we will prove the complexity of \texttt{push}.
+While the amortized cost is the same as the actual cost, we need to make sure that our potential function does not cause those two to diverge.
+As LiquidHaskell does not allow quantifiers to ensure termination of the SMT solver, we need to guess an upper limit that our amortized runtime will have.
+As seen earlier, we can pay two time units for a push to amortize the pops, so we will use $2$ as our guess.
+To specify our proof goal, we need to name the parameters in the LiquidHaskell annotation.
+After than we can use equational reasoning to proof our goal.
+
+\begin{lstlisting}
+import Language.Haskell.Liquid.ProofCombinators
+
+{-@ pushP :: x:a -> s:Stack a ->
+  { pushT x s + phi (push x s) - phi s <= 2 }
+@-}
+pushP :: a -> Stack a -> Proof
+pushP x s = pushT x s + phi (push x s) - phi s
+-- Use definition of push and pushT
+  === 1 + phi (Elem x s) - phi s
+-- Use definition of phi (Elem _ _)
+  === 1 + (1 + phi s) - phi s
+-- phi s gets canceled, 1 + 1 <= 2
+  *** QED
+\end{lstlisting}
+
+To proof complexity of \texttt{multipop}, we will need to make a case distinction between the two base cases and the recursive case.
+With LiquidHaskell this is just a normal pattern match.
+Leaving the recursive case undefined for now, proving the base cases is very similar to \texttt{push}.
+
+\begin{lstlisting}
+{-@ multipopP :: k:Nat -> s:Stack a ->
+  { multipopT k s + phi (multipop k s)
+      - phi s <= 2 }
+@-}
+multipopP :: Int -> Stack a -> Proof
+multipopP 0 s = multipopT 0 s
+    + phi (multipop 0 s) - phi s
+-- Use definition of multipop/T 0 _
+  === 1 + phi s - phi s
+-- phi s cancels, 1 <= 2
+  *** QED
+multipopP k Empty = multipopT k Empty
+    + phi (multipop k Empty) - phi Empty
+-- Use definition of multipop/T _ Empty
+  === 1 + phi Empty - phi Empty
+-- phi Empty cancels, 1 <= 2
+  *** QED
+\end{lstlisting}
+
+To avoid cluttering the proof of the recursive case, we use an \textit{as pattern} to alias \texttt{Elem x s} to \texttt{xs}.
+The first steps again are applying the definitions of the used functions and simplifying the result.
+In the end we use the \texttt{?} operator to add the induction hypothesis to the proof step and thus complete the proof.
+
+\begin{lstlisting}
+multipopP k xs@(Elem x s) = multipopT k xs
+    + phi (multipop k xs) - phi xs
+-- Use definition of multipopT for recusive case
+  === (1 + multipopT (k - 1) s)
+    + phi (multipop k xs) - phi xs
+-- Use definition of multipop for recusive case
+  === (1 + multipopT (k - 1) s)
+    + phi (multipop (k - 1) s) - phi xs
+-- Use definition of phi (Elem _ _)
+  === (1 + multipopT (k - 1) s)
+    + phi (multipop (k - 1) s) - (1 + phi s)
+-- Remove parenthesis
+  === 1 + multipopT (k - 1) s
+    + phi (multipop (k - 1) s) - 1 - phi s
+-- 1s cancel out
+  === multipopT (k - 1) s
+    + phi (multipop (k - 1) s) - phi s
+-- use induction
+  ? multipopP (k - 1) s
+  *** QED
+\end{lstlisting}
+
 \subsection{Finger Tree}
 
-\section{Limitations of this approach}
+To show that this approach scales also to more complicated data structures that are widely used in the industry, we will prove the complexity of a finger tree, the data structure behind \texttt{Data.Seq} from \textit{containers}~\cite{containers}.
+It is originally described by Hinze and Paterson~\cite{fingertrees}, which the implementation from \textit{containers} closely follows.
+The following definition is taken directly from \textit{containers}, but slightly modified to remove the size parameter.
+LiquidHaskell caused problems in the proofs when using type classes, so because the functions we will use do not use the size we omitted it.
+We will discuss such limitations more in depth in section~\ref{sec:limitations}.
+
+\begin{lstlisting}
+data FingerTree a = Empty
+  | Single a
+  | Deep (Digit a) (FingerTree (Node a)) (Digit a)
+
+data Node a = Node2 a a | Node3 a a a
+
+data Digit a = One a
+  | Two a a
+  | Three a a a
+  | Four a a a a
+\end{lstlisting}
+
+The finger tree has between one and four \textit{digits} on either side to enable cheap \texttt{cons} and \texttt{snoc}.
+As shown by Claessen~\cite{fingertrees_new}, this would also work with a maximum of just three digits, but we will go with the definition from \textit{containers}.
+The most interesting point about the data structure is the polymorphic recursion that makes the data structure type safe as it guarantees equal nesting of \texttt{Node} types on both sides of the spine.
+It is also what makes it difficult to verify in other languages, because for example the Isabelle/HOL~\cite{isabelle} and the Lean~\cite{lean} theorem provers both do not directly support polymorphic recursion.
+In these languages, it is necessary to define the type such that it would allow uneven nesting of \texttt{Node}s and then "carve out" only the valid finger trees.
+Other provers like Coq~\cite{coq} support polymorphic recursion making a more direct one to one translation possible.
+
+\subsubsection{Cons and Snoc}
+
+As the finger tree is a symmetric tree, \texttt{cons} and \texttt{snoc} are very similar.
+For this reason we will show an explicit proof only for \texttt{cons} and provide the proof for \texttt{snoc} in the appendix.
+First we again copy definition of \texttt{cons} from the \textit{containers} package, adjusting it slightly to remove the size information that would require the type class.
+Also, to later use the function in a proof we add a \texttt{reflect} pragma.
+
+\begin{lstlisting}
+{-@ reflect cons @-}
+cons :: a -> FingerTree a -> FingerTree a
+cons a EmptyT = Single a
+cons a (Single b) =
+  Deep (One a) EmptyT (One b)
+cons a (Deep (One b) m sf) =
+  Deep (Two a b) m sf
+cons a (Deep (Two b c) m sf) =
+  Deep (Three a b c) m sf
+cons a (Deep (Three b c d) m sf) =
+  Deep (Four a b c d) m sf
+cons a (Deep (Four b c d e) m sf) =
+  Deep (Two a b) (cons (Node3 c d e) m) sf
+\end{lstlisting}
+
+The only complicated case is if we have already four digits at the front, as in all other cases we can simply increase the amount by one.
+In the case that we can not do so, we put the new data in front and push the "overflow" in to the recursion.
+The next step is to think of a valid potential function $\Phi$.
+Again, the intuition is which state of the data structure makes \texttt{cons} expensive.
+As seen above it is when the digit in the front is already full.
+So we can define the \textit{danger} of a digit as every state that will make the operation expensive.
+To also support and proof \texttt{uncons}, the inverse operation, we not only mark a full digit as dangerous, but also one that has only one element inside (as this would make \texttt{uncons} go into the recursion).
+Our potential is then the sum of the danger in the tree.
+
+\begin{lstlisting}
+{-@ reflect dang @-}
+{-@ dang :: Digit a -> Nat @-}
+dang :: Digit a -> Int
+dang One{} = 1
+dang Two{} = 0
+dang Three{} = 0
+dang Four{} = 1
+
+{-@ reflect phi @-}
+{-@ phi :: FingerTree a -> Nat @-}
+phi :: FingerTree a -> Int
+phi EmptyT = 0
+phi Single{} = 0
+phi (Deep pr m sf) =
+  dang pr + phi m + dang sf
+\end{lstlisting}
+
+The timing function for cons is similar to the timing functions in the stack just a copy of the original function with all non-recursive cases return $1$.
+
+\begin{lstlisting}
+{-@ reflect consT @-}
+{-@ consT :: a -> FingerTree a -> Nat @-}
+consT :: a -> FingerTree a -> Int
+consT _ EmptyT = 1
+consT _ Single{} = 1
+consT _ (Deep One{} _ _) = 1
+consT _ (Deep Two{} _ _) = 1
+consT _ (Deep Three{} _ _) = 1
+consT _ (Deep (Four _ c d e) m _) =
+  1 + consT (Node3 c d e) m
+\end{lstlisting}
+
+With all the preliminaries out of the way we can start proving the amortized complexity.
+Like with the stack we need to guess a constant that is an upper bound on our amortized run time.
+As we do not have a simple mental model like with the stack, we will delay this choice until we have proven the base cases.
+We will also use \textit{as patterns} again to avoid repetition.
+
+\begin{lstlisting}
+{-@ consP :: x:a -> t:FingerTree a ->
+  { consT x t + phi (cons x t) - phi t <= c }
+@-}
+consP :: a -> FingerTree a -> Proof
+consP x t@EmptyT =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of cons/T _ EmptyT
+  === 1 + phi (Single x) - phi EmptyT
+-- Use definition of phi
+  === 1 + 0 - 0
+-- 1 must be <= c, so c >= 1
+  *** QED
+consP x t@(Single a) =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of cons/T _ (Single a)
+  === 1 + phi (Deep (One x) EmptyT (One a))
+        - phi (Single a)
+-- Use definition of phi
+  === 1 + dang (One x) + phi EmptyT
+        + dang (One a) - 0
+-- Use definition of phi and dang
+  === 1 + 1 + 0 + 1 - 0
+-- 3 must be <= c, so c >= 3
+  *** QED
+\end{lstlisting}
+
+We omit the proof cases for one to three digits, as they work exactly as the other two base cases.
+The interesting case is when we have four digits.
+In this case we need to use induction to proof our goal.
+
+\begin{lstlisting}
+consP x t@(Deep (Four a b c d) m sf) =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of consT
+  === (1 + consT (Node3 b c d) m)
+  + phi (cons x t) - phi t
+-- Use definition of cons
+  === (1 + consT (Node3 b c d) m)
+  + phi (Deep (Two x a) (cons (Node3 b c d) m) sf)
+  - phi t
+-- Use definition of phi
+  === (1 + consT (Node3 b c d) m)
+  + dang (Two x a) + phi (cons (Node3 b c d) m)
+  + dang sf
+  - (dang (Four a b c d) + phi m + dang sf)
+-- Remove parenthesis
+  === 1 + consT (Node3 b c d) m
+  + dang (Two x a) + phi (cons (Node3 b c d) m)
+  + dang sf
+  - dang (Four a b c d) - phi m - dang sf
+-- dang sf cancels, use definition of dang
+  === 1 + consT (Node3 b c d) m
+  + 0 + phi (cons (Node3 b c d) m) - 1 - phi m
+-- 1s cancel
+  === consT (Node3 b c d) m
+  + phi (cons (Node3 b c d) m) - phi m
+-- Use induction
+  ? consP (Node3 b c d) m
+  *** QED
+\end{lstlisting}
+
+\subsubsection{Append}\label{sec:append}
+
+Similar to \texttt{push} in the stack, even though \texttt{append} has the same amortized complexity as its normal complexity, we need to show that it never increases the potential more than a logarithmic amount.
+Contrary to the simplified definition of Claessen~\cite{fingertrees_new}, the \textit{containers} package uses two sets of code generated, mutually recursive functions to implement \texttt{append}.
+One set of functions called \texttt{appendTreeX} handles the base cases where one of the two trees is either empty or consists of a singular element by using \texttt{cons} and \texttt{snoc}.
+In the recursive case it calls the \texttt{addDigitsX} functions that group arguments as \texttt{Node}s and itself call the \texttt{appendTreeX} functions.
+Because the \texttt{addDigitsX} functions do pattern match on their tree arguments at all, LiquidHaskell can not find a termination metric by itself.
+For this, we need to track the arguments across functions calls because \texttt{appendTreeX} has a strictly decreasing argument.
+We can do so by adding a ghost parameter to order the functions for LiquidHaskell.
+This order is lower for all \texttt{appendTreeX} functions than the \texttt{addDigitsX} functions.
+Additionally we need to set the case expansion to a high number such that all pattern matches are evaluation, else the termination proof fails.
+See the following snippet as an example with most cases and function omitted for brevity.
+
+\begin{lstlisting}
+{-@ LIQUID --max-case-expand=1000 @-}
+
+{-@ appendTree1 :: a:FingerTree (Node a) ->
+  Node a -> FingerTree (Node a) ->
+  FingerTree (Node a) / [size a, 1] @-}
+{-@ reflect appendTree1 @-}
+appendTree1 :: FingerTree (Node a) -> Node a ->
+    FingerTree (Node a) -> FingerTree (Node a)
+appendTree1 EmptyT !a xs = a `consTree` xs
+appendTree1 xs !a EmptyT = xs `snocTree` a
+appendTree1 (Single x) !a xs =
+    x `consTree` a `consTree` xs
+appendTree1 xs !a (Single x) =
+    xs `snocTree` a `snocTree` x
+appendTree1 (Deep pr1 m1 sf1) a (Deep pr2 m2 sf2) =
+    Deep pr1 (addDigits1 m1 sf1 a pr2 m2) sf2
+
+{-@ addDigits1 :: a:FingerTree (Node (Node a)) ->
+  Digit (Node a) -> Node a -> Digit (Node a) ->
+  FingerTree (Node (Node a)) ->
+  FingerTree (Node (Node a)) / [size a, 5] @-}
+{-@ reflect addDigits1 @-}
+addDigits1 :: FingerTree (Node (Node a)) ->
+  Digit (Node a) -> Node a -> Digit (Node a) ->
+  FingerTree (Node (Node a)) ->
+  FingerTree (Node (Node a))
+addDigits1 m1 (One a) b (One c) m2 =
+  appendTree1 m1 (Node3 a b c) m2
+addDigits1 m1 (One a) b (Two c d) m2 =
+  appendTree2 m1 (Node2 a b) (Node2 c d) m2
+-- further cases omitted
+\end{lstlisting}
+
+This case expansion results in an explosion of constraints for the SMT solver to check.
+While the whole proof for \texttt{cons} including the defined functions generates less than 100 constraints, this termination proof generates over 3000.
+This leads to extremely high compile times.
+This is made worse by the need to define a timing function that needs to follow this mutually recursive structure and generates an additional 8000 constraints such that loading the file takes several minutes.
+Another problem is the logarithmic runtime itself.
+SMT provers are great with linear arithmetic, but are significantly slower with multiplication.
+Division is usually left as an unevaluated function.
+While one can transform the proof goal to turn logarithms into powers, it still requires assuming basic arithmetic rules on the logic level.
+
+For those reasons, using LiquidHaskell to proof the complexity with the physicist's method is not feasible (see section~\ref{sec:limitations} for a discussion of those limitations).
+Claessen~\cite{fingertrees_new} also does not formally proof this in his functional pearl, but provides a different reasoning to why \texttt{append} still is logarithmic.
+The potential only depends on the spine of the tree, so a new layer in the tree only adds at most a constant amount to the potential ($2$ if both digits are dangerous) making it grow linearly.
+Because the nodes in the digits always at least double (in the case of two \texttt{Node2}) the overall size of the tree grows exponentially.
+Thus the potential grows logarithmic compared to the size of the tree.
+
+\section{Limitations of this approach}\label{sec:limitations}
+
+\subsection{Definition of the timing function}
+
+One of the points where the analysis may lead to wrong results is if the timing function does not directly mirror the original function.
+So when the implementation changes and the timing function does not, it might be possible for the proofs to still pass.
+While this is unlikely, given how verbose proofs with LiquidHaskell are, it is not negligible.
+
+If one is in control of the source code and can change the implementation, it would be possible to factor out the recursion and share it between the implementation and the timing function.
+This however would clutter the implementation and most likely decrease performance as it is often the case with recursion schemes.
+
+A better solution would be to derive the timing function using meta programming in the form of TemplateHaskell.
+This meta function would take the name of the function in the implementation and create a new one with a configurable suffix.
+To do so it would check if the right hand sides contain a recursive call to the given name.
+If not, it replaces the right hand side with $1$, otherwise it changes the recursion to use the name with suffix and add $1$ at the end.
+The resulting function could then be spliced back into the source code.
+
+\begin{lstlisting}
+cons :: a -> FingerTree a -> FingerTree a
+cons a EmptyT = Single a
+cons a (Deep (Four b c d e) m sf) =
+    Deep (Two a b) (cons (Node3 c d e) m) sf
+-- other cases omitted
+
+$(mkTimingFn "T" ''cons)
+-- generates this; variable names omitted:
+-- consT :: a -> FingerTree a -> Int
+-- consT _ EmptyT = 1
+-- consT _ (Deep (Four _ c d e) m _) =
+--    consT (Node3 c d e) m + 1
+-- other cases omitted
+\end{lstlisting}
+
+\subsection{LiquidHaskell for verification}
+
+As shown in earlier sections, LiquidHaskell has some limitations at the moment with regards to proving more complex functions.
+First and foremost the support for type classes or at least the support for instances involving Haskell \texttt{newtype}s is not great.
+Currently, LiquidHaskell only supports measure classes, which need to be defined by recursion over a single data type similar to a normal measure.
+For this reason it was not possible to use it for the \texttt{Sized} type class from \textit{containers}.
+There are several open issues about the type class support on the LiquidHaskell bug tracker, so we expect this to get better in the future.
+
+Another pain point was the combinatorial explosion when generating the constraints for the termination of \texttt{append} (see section~\ref{sec:append}).
+It might be possible to be smarter when generating the constraints here and thus creating less work for the SMT solver.
+LiquidHaskell also provides an option to disable the termination checker, but enabling this option surprisingly did not improve compile time.
+In fact, several other options that are supposed to improve performance lik \texttt{{-}{-}diff} which enables incremental type checking or \texttt{{-}{-}fast} which disables inference of polymorphic refinement, did either not affect performance at all or (in the case of \texttt{{-}{-}fast}) made it even worse.
+
+\section{Conclusion}
+
+Despite the current limitations of LiquidHaskell preventing more complicated functions from being proved formally, there is still a lot of value in this approach.
+Proving complexity of core data structures immediately pays back in a huge gain in confidence across the many applications that depend on them.
+While at the moment, amortized complexity of \O{1} seems to be the most feasable to proof, we expect the tooling around LiquidHaskell to improve a lot in the future, be it by incremental changes like better support for type classes or by more complicated means like a deeper embedding into a future dependently typed Haskell core.
+
+Aside from the industry application, we think this approach also has a big didactic advantage.
+To learn more about the complexity of data structures, a learner does not need to learn a new language on top of new concepts, but instead can use the language they already know to apply their new knowledge.
+The recent improvements in the usability of LiquidHaskell --- turning the previously standalone executable into a GHC plugin~\cite{lh_plugin} --- play a big role here as it drastically reduces the barrier of entry.
 

### 2021-05-21 paper/main.tex
Finish first paper draft

@@
 \definecolor{codegreen}{rgb}{0,0.6,0}
+
+\usepackage{cancel}
 
@@
 \begin{abstract}
-TODO
+Proofs of complexity of production code is usually done in external theorem provers.
+This makes it necessary to translate the data structures to be proved from the production language into the language of the prover.
+Such a translation step may introduce errors by a mismatch in features in the two languages.
+We show how to prove amortized complexity of data structures directly in Haskell using LiquidHaskell~\cite{liquidhaskell}.
+Besides skipping the translation step, our approach also has a didactic advantage as new learners do not need to learn a new language for their proofs and can focus on the new concepts only.
+We do not assume prior knowledge of amortized complexity as we explain the concepts and apply them in our first case study, a simple stack with multipop.
+The usefulness of our approach in a production setting is verified by the second case study, the finger tree which powers \texttt{Data.Seq} from the \textit{containers} package which is widely depended on.
+Finally we discuss the current limitations of LiquidHaskell in this area and how they might be lifted.
 \end{abstract}
@@
 
-\begin{itemize}
-\item{Worst case is often too pessimistic}
-\item{Idea: spread out expensive operations over multiple cheap ones}
-\item{Usually only talks about sequence of operations without sharing}
-\item{Formalization: Physicist's method}
-\end{itemize}
+It is common to check how many time units an operation on a data structure takes in relation to the number of elements in the data structure.
+As the performance can vary wildly for the same operation depending on the state of data structure --- a linear search takes $1$ time unit if the searched element is at the front, while it takes $n$ time units if it is the last element --- usually one uses the worst case for an operation as a metric and uses Landau notation to specify it.
+With our linear search example we would say that it is in \O{n} because it will not perform worse than $n$ time units.
+
+But often this worst case is too pessimistic.
+When an operation changes the data structure, for example by removing elements, consecutive uses of operation may be cheaper.
+Or to use an expensive operation it might be necessary to have a certain amount of cheap operations.
+In this case one uses \textit{amortized complexity} which does not only look at a single operation in isolation, but on chains of operations.
+In particular, it assumes that that the intermediate data structures are not shared, i.e.
+they are not persisted.
+There are multiple methods to proof amortized complexity.
+
+As a running example we will use a stack with \textit{push} and \textit{multipop} operations.
+The former allocates a new element and a references to the rest of the stack and thus taking a constant amount of time.
+The latter removes the first $k$ elements from the stack.
+In the worst case, we remove all the elements, making \texttt{multipop} \O{n}.
+One can already see that this is too pessimistic, as after removing all elements from the stack, the next \texttt{multipop} will do nothing.
+In fact, every element we remove needs to be added to the stack at some point, so we can spread out the cost of \texttt{multipop} over many pushes.
+There are multiple techniques to prove amortized complexity, we present the two most popular ones.
+
+\subsection{Banker's method}\label{sec:banker}
+
+The banker's method works by having an bank account that can store time units.
+Operations have to pay their actual cost and may additionally pay extra into the account.
+Expensive operations can then use the stored time units in the bank to pay for themselves.
+In our example a \texttt{push} has to pay one actual time unit and it will also pay one extra time unit into the bank account.
+This means that the bank account always contains as many time units as there are elements on the stack.
+That way, any call to \texttt{multipop} can use the bank account to pay for itself and the account will never become negative.
+
+\subsection{Physicist's method}\label{sec:physicist}
+
+The banker's method is hard to use for a formal proof because the bank account acts as state between the operations.
+A simpler method is the physicist's method that does not need such an account.
+The idea is to define a potential $\Phi$ ("phi") for the data structure.
+This potential should be $0$ for an empty data structure and $\ge 0$ for all other states.
+Intuitively the potential will declare how much time needs to be saved up to pay for expensive operations.
+We then define the amortized cost of an operation as $c + \Phi(h') - \Phi(h)$ where $c$ is the actual cost for the operation, $h$ is the data structure before the operation and $h'$ is the data structure afterwards.
+Looking at a chain of operations $1, 2, ..., n$, the sum of this amortized time is an overestimation of the actual cost by $\Phi(h_n)$ --- because $\Phi(h_0)$ is by definition $0$ --- and thus an upper bound on the actual cost (see equation~\ref{eq:sum}).
+
+\begin{gather}
+\nonumber (c_1 + \cancel{\Phi(h_1)} - \Phi(h_0)) + (c_2 + \cancel{\Phi(h_2)} - \cancel{\Phi(h_1)}) \\
+\nonumber    + ... + (c_n + \Phi(h_n) - \cancel{\Phi(h_{n - 1})}) = \\
+  c_1 + c_2 + ... + c_n + \Phi(h_n) - \Phi(h_0) \label{eq:sum}
+\end{gather}
+
+For our stack we can define the potential to be the height of the stack.
+This fulfils the requirements as the empty stack has a potential of zero and all other states have a potential greater or equal to zero.
+For the actual proof of this example see the first case study in section~\ref{sec:stack}.
 
@@
 
-As described by Vazou et. al.~\cite{tpfa}, one can use LiquidHaskell to do complex equational reasoning that is machine checked by the compiler.
+As described by Vazou et.
+al.~\cite{tpfa}, one can use LiquidHaskell to do complex equational reasoning that is machine checked by the compiler.
+It uses refinement types~\cite{refinement_types} that add additional constraints on top of normal Haskell data types which are checked by an external SMT solver.
+By default LiquidHaskell uses the z3~\cite{z3} solver, but it also supports cvc4~\cite{cvc4}.
+For example we can declare the type of integers greater or equal to zero with a type synonym.
+LiquidHaskell annotions use comments with an @ sign, so that the code still compiles even without LiquidHaskell.
+It also allows to name arguments, so they can be referred to later in the refinement.
+At the moment to define functions with refinement, one has to copy the Haskell type signature into a LiquidHaskell comment and add the refinements there.
+A future version that is more deeply integrated into the Haskell compiler may not need this code duplication.
 
-\begin{itemize}
-\item{Uses refinement types}
-\item{Refinments are solved by external SMT solver (z3)}
-\item{With the refinement one can define proof combinators}
-\item{Use equational reasoning and induction to prove interesing properties}
-\end{itemize}
+\begin{lstlisting}
+{-@ type Nat = { x:Int | x >= 0 } @-}
+\end{lstlisting}
+
+In LiquidHaskell, all functions need to terminate.
+Usually a suitable termination metric is automatically guessed, but in some more complicated cases, we can also manually supply this metric (see section~\ref{sec:append} for an example of this).
+A simple recursive function can be marked as \textit{measure} which allows the use of the function in those termination metrics as well as in the refinement itself.
+With the termination, LiquidHaskell is able to generate an induction principle for the function that can be used in a proof.
+Because LiquidHaskell is just Haskell, this induction amounts to a recursive call to the proof and is also used when defining ordinary Haskell functions with refinement on their types.
+For example the fact that the length of a list is never negative.
+
+\begin{lstlisting}
+{-@ measure length @-}
+{-@ length :: [a] -> Nat @-}
+length :: [a] -> Int
+length [] = 0 -- trivial, 0 >= 0
+length (x:xs) = 1 + length xs
+    -- by induction: length xs >= 0
+    -- so 1 + length xs >= 0
+\end{lstlisting}
+
+LiquidHaskell also provides a set of proof combinators and a \texttt{Proof} type.
+This type is just a type alias for the normal Haskell unit type.
+The reasoning comes from the \texttt{===} operator that asserts that both sides of the equality are truly equal.
+As a simple example we can proof that the length of an \texttt{append} function is the sum of the lengths of the two original lists.
+As append is not a \textit{measure} (it does not return a natural number), we have to use \textit{reflection}~\cite{reflection} to lift it to the logic level of LiquidHaskell.
+The proof itself is done by case distinction on the first argument.
+This distinction is done by pattern matching on it like one would in a normal Haskell function.
+We start with stating the left hand side of the equation how it is in the refinement type and try to get to the right hand side of the equality by simplification and induction.
+
+\begin{lstlisting}
+{-@ reflect append @-}
+append :: [a] -> [a] -> [a]
+append [] ys = ys
+append (x:xs) ys = x:(append xs ys)
+
+{-@ lengthP :: xs:[a] -> ys:[a] ->
+  { length (append xs ys) ==
+      length xs + length ys }
+@-}
+lengthP :: [a] -> [a] -> Proof
+
+-- base case
+lengthP [] ys = length (append [] ys)
+  -- Use definition of append [] _
+    === length ys
+    === 0 + length ys
+  -- use definition of length []
+    === length [] + length ys
+    *** QED
+
+-- recursive case
+lengthP (x:xs) ys =
+    length (append (x:xs) ys)
+  -- Use definition of append
+    === length (x:(append xs ys))
+  -- Use definition of length
+    === 1 + length (append xs ys)
+  -- Use induction with the ? operator
+    ? lengthP xs ys
+    === 1 + length xs + length ys
+  -- Use definition of length
+    === length (x:xs) + length ys
+    *** QED
+\end{lstlisting}
+
+The triple star operator (\texttt{***}) and the \texttt{QED} type at the end are just to cast our proof to Haskell's unit type and to make the proof look nicer.
 
@@
 
-\subsection{Stack with Multipop}
+\subsection{Stack with Multipop}\label{sec:stack}
 
-One of the simplest data structures that have an amortized complexity is a stack with two operations: Push and Multipop.
-$push(e)$ adds a singular element $e$ on top of the stack, while $multipop(k)$ removes the top $k$ elements.
+To continue with the example from section~\ref{sec:complexity}, we want to prove that $multipop$ has an amortized complexity of \O{1}.
 This data structure and its operations could be defined like this in Haskell.
@@
 
-If we look at this definition we can easily see that \texttt{push} takes constant time.
-No matter how large the rest of the stack is, it will only allocate a new \texttt{Pushed} cell that contains a reference to the rest of the stack.
-Thus this operation has complexity \O{1}.
-
-\texttt{multipop} however may pop all of the elements in the stack if $k$ is greater or equal to $n$, where $n$ is the height of the stack.
-So in the worst case this operation takes $n$ steps and thus it is in \O{n}.
-
-This worst case however is a bit too pessimistic.
-If we pop all elements, we will not be able to do it again, as the stack is empty after the first operation.
-In fact, in order for an element to be popped, it first has to be pushed.
-So we can spread the cost for the expensive $multipop$ operation over all the constant time pushes, making \texttt{multipop} amortized \O{1}.
-With the banker's method this would mean for a push we could pay one time unit actual cost and one time unit extra for the bank.
-Then the bank has always as much time units as there are elements in the stack.
-Thus a \texttt{multipop} will not be able to make the bank account negative.
-
-To formally prove this with the physicist's method, we first need to define a potential $\Phi$ and a timing function for each of our operations.
-The intuition for the potential is what property of the data structure makes \texttt{multipop} more expensive.
-Because a stack is such a simple data structure, there is basically only one property: its height.
-So we can define a function that calculates the height of a stack and use it as our potential.
+To formally prove this with the physicist's method, we first need to define a potential $\Phi$ and a timing function for each of our operations that gives us the actual cost of that operation.
+As seen in section~\ref{sec:physicist}, we use the height of the stack as our potential.
+Thus we need to define a function that calculates the height of a stack.
 To define the timing functions, we copy the code for the operation, replace constant time steps with $1$ and adjust the recursion.
@@
 As LiquidHaskell does not allow quantifiers to ensure termination of the SMT solver, we need to guess an upper limit that our amortized runtime will have.
-As seen earlier, we can pay two time units for a push to amortize the pops, so we will use $2$ as our guess.
+As seen earlier in section~\ref{sec:banker}, we can pay two time units for a push to amortize the pops, so we will use $2$ as our guess.
 To specify our proof goal, we need to name the parameters in the LiquidHaskell annotation.
@@
 We omit the proof cases for one to three digits, as they work exactly as the other two base cases.
+The constant however will not change in the other base cases, so $3$ really is our upper bound to use here.
 The interesting case is when we have four digits.
@@
 
-\section{Limitations of this approach}\label{sec:limitations}
-
 \subsection{Definition of the timing function}
@@
 
+\section{Full proof code for the stack}
+
+\begin{lstlisting}
+import Language.Haskell.Liquid.ProofCombinators
+
+data Stack a = Empty | Elem a (Stack a)
+
+{-@ reflect push @-}
+push :: a -> Stack a -> Stack a
+push = Elem
+
+{-@ multipop :: Nat -> Stack a -> Stack a @-}
+multipop :: Int -> Stack a -> Stack a
+multipop _ Empty = Empty
+multipop 0 s = s
+multipop n (Elem _ s) = multipop (n - 1) s
+
+{-@ reflect phi @-}
+{-@ phi :: Stack a -> Nat @-}
+phi :: Stack a -> Int
+phi Empty = 0
+phi (Elem _ s) = 1 + phi s
+
+{-@ reflect pushT @-}
+{-@ pushT :: a -> Stack a -> Nat @-}
+pushT :: a -> Stack a -> Int
+pushT _ _ = 1
+
+{-@ reflect multipopT @-}
+{-@ multipopT :: Nat -> Stack a -> Nat @-}
+multipopT :: Int -> Stack a -> Int
+multipopT _ Empty = 1
+multipopT 0 _ = 1
+multipopT n (Elem _ s) = 1 + multipopT (n - 1) s
+
+{-@ pushP :: x:a -> s:Stack a ->
+  { pushT x s + phi (push x s) - phi s <= 2 }
+@-}
+pushP :: a -> Stack a -> Proof
+pushP x s = pushT x s + phi (push x s) - phi s
+-- Use definition of push and pushT
+  === 1 + phi (Elem x s) - phi s
+-- Use definition of phi (Elem _ _)
+  === 1 + (1 + phi s) - phi s
+-- phi s gets canceled, 1 + 1 <= 2
+  *** QED
+
+{-@ multipopP :: k:Nat -> s:Stack a ->
+  { multipopT k s + phi (multipop k s)
+      - phi s <= 2 }
+@-}
+multipopP :: Int -> Stack a -> Proof
+multipopP 0 s = multipopT 0 s
+    + phi (multipop 0 s) - phi s
+-- Use definition of multipop/T 0 _
+  === 1 + phi s - phi s
+-- phi s cancels, 1 <= 2
+  *** QED
+multipopP k Empty = multipopT k Empty
+    + phi (multipop k Empty) - phi Empty
+-- Use definition of multipop/T _ Empty
+  === 1 + phi Empty - phi Empty
+-- phi Empty cancels, 1 <= 2
+  *** QED
+multipopP k xs@(Elem x s) = multipopT k xs
+    + phi (multipop k xs) - phi xs
+-- Use definition of multipopT for recusive case
+  === (1 + multipopT (k - 1) s)
+    + phi (multipop k xs) - phi xs
+-- Use definition of multipop for recusive case
+  === (1 + multipopT (k - 1) s)
+    + phi (multipop (k - 1) s) - phi xs
+-- Use definition of phi (Elem _ _)
+  === (1 + multipopT (k - 1) s)
+    + phi (multipop (k - 1) s) - (1 + phi s)
+-- Remove parenthesis
+  === 1 + multipopT (k - 1) s
+    + phi (multipop (k - 1) s) - 1 - phi s
+-- 1s cancel out
+  === multipopT (k - 1) s
+    + phi (multipop (k - 1) s) - phi s
+-- use induction
+  ? multipopP (k - 1) s
+  *** QED
+\end{lstlisting}
+
+\section{Full proof code of the finger tree}\label{ap:fingertree}
+
+\begin{lstlisting}
+data FingerTree a = Empty
+  | Single a
+  | Deep (Digit a) (FingerTree (Node a)) (Digit a)
+
+data Node a = Node2 a a | Node3 a a a
+
+data Digit a = One a
+  | Two a a
+  | Three a a a
+  | Four a a a a
+
+{-@ reflect cons @-}
+cons :: a -> FingerTree a -> FingerTree a
+cons a EmptyT = Single a
+cons a (Single b) =
+  Deep (One a) EmptyT (One b)
+cons a (Deep (One b) m sf) =
+  Deep (Two a b) m sf
+cons a (Deep (Two b c) m sf) =
+  Deep (Three a b c) m sf
+cons a (Deep (Three b c d) m sf) =
+  Deep (Four a b c d) m sf
+cons a (Deep (Four b c d e) m sf) =
+  Deep (Two a b) (cons (Node3 c d e) m) sf
+
+{-@ reflect snoc @-}
+snoc :: FingerTree a -> a -> FingerTree a
+snoc EmptyT x = Single x
+snoc (Single a) x =
+  Deep (One a) EmptyT (One x)
+snoc (Deep pr m (One a)) x =
+  Deep pr m (Two a x)
+snoc (Deep pr m (Two a b)) x =
+  Deep pr m (Three a b x)
+snoc (Deep pr m (Three a b c)) x =
+  Deep pr m (Four a b c x)
+snoc (Deep pr m (Four a b c d)) x =
+  Deep pr (snoc m (Node3 a b c)) (Two d x)
+
+{-@ reflect dang @-}
+{-@ dang :: Digit a -> Nat @-}
+dang :: Digit a -> Int
+dang One{} = 1
+dang Two{} = 0
+dang Three{} = 0
+dang Four{} = 1
+
+{-@ reflect phi @-}
+{-@ phi :: FingerTree a -> Nat @-}
+phi :: FingerTree a -> Int
+phi EmptyT = 0
+phi Single{} = 0
+phi (Deep pr m sf) =
+  dang pr + phi m + dang sf
+
+{-@ reflect consT @-}
+{-@ consT :: a -> FingerTree a -> Nat @-}
+consT :: a -> FingerTree a -> Int
+consT _ EmptyT = 1
+consT _ Single{} = 1
+consT _ (Deep One{} _ _) = 1
+consT _ (Deep Two{} _ _) = 1
+consT _ (Deep Three{} _ _) = 1
+consT _ (Deep (Four _ c d e) m _) =
+  1 + consT (Node3 c d e) m
+
+{-@ reflect snocT @-}
+{-@ snocT :: FingerTree a -> a -> Nat @-}
+snocT :: FingerTree a -> a -> Int
+snocT EmptyT _ = 1
+snocT Single{} _ = 1
+snocT (Deep _ _ One{}) _ = 1
+snocT (Deep _ _ Two{}) _ = 1
+snocT (Deep _ _ Three{}) _ = 1
+snocT (Deep _ m (Four a b c _)) _ =
+    1 + snocT m (Node3 a b c)
+
+{-@ consP :: x:a -> t:FingerTree a ->
+  { consT x t + phi (cons x t) - phi t <= 3 }
+@-}
+consP :: a -> FingerTree a -> Proof
+consP x t@EmptyT =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of cons/T _ EmptyT
+  === 1 + phi (Single x) - phi EmptyT
+-- Use definition of phi
+  === 1 + 0 - 0 -- trivial, 1 <= 3
+  *** QED
+consP x t@(Single a) =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of cons/T _ (Single a)
+  === 1 + phi (Deep (One x) EmptyT (One a))
+        - phi (Single a)
+-- Use definition of phi
+  === 1 + dang (One x) + phi EmptyT
+        + dang (One a) - 0
+-- Use definition of phi and dang
+  === 1 + 1 + 0 + 1 - 0 -- trivial, 3 <= 3
+  *** QED
+consP x t@(Deep (One a) m sf) =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of cons/T
+  === 1 + phi (Deep (Two x a) m sf) - phi t
+-- Use definition of phi
+  === 1 + dang (Two x a) + phi m + dang sf
+    - (dang (One a) + phi m + dang sf)
+-- Remove parenthesis, use definition of dang
+  === 1 + 0 + phi m + dang sf
+    - 1 - phi m - dang sf
+-- everything cancels
+  === 0 -- trivial, 0 <= 3
+  *** QED
+consP x t@(Deep (Two a b) m sf) =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of cons/T
+  === 1 + phi (Deep (Three x a b) m sf) - phi t
+-- Use definition of phi
+  === 1 + dang (Three x a b) + phi m + dang sf
+    - (dang (Two a b) + phi m + dang sf)
+-- Remove parenthesis, use definition of dang
+  === 1 + 0 + phi m + dang sf
+    - 0 - phi m - dang sf
+-- phi and dang cancels
+  === 1 -- trivial, 1 <= 3
+  *** QED
+consP x t@(Deep (Three a b c) m sf) =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of cons/T
+  === 1 + phi (Deep (Four x a b c) m sf) - phi t
+-- Use definition of phi
+  === 1 + dang (Four x a b c) + phi m + dang sf
+    - (dang (Three a b c) + phi m + dang sf)
+-- Remove parenthesis, use definition of dang
+  === 1 + 1 + phi m + dang sf
+    - 0 - phi m - dang sf
+-- phi and dang cancels
+  === 2 -- trivial, 2 <= 3
+  *** QED
+consP x t@(Deep (Four a b c d) m sf) =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of consT
+  === (1 + consT (Node3 b c d) m)
+  + phi (cons x t) - phi t
+-- Use definition of cons
+  === (1 + consT (Node3 b c d) m)
+  + phi (Deep (Two x a) (cons (Node3 b c d) m) sf)
+  - phi t
+-- Use definition of phi
+  === (1 + consT (Node3 b c d) m)
+  + dang (Two x a) + phi (cons (Node3 b c d) m)
+  + dang sf
+  - (dang (Four a b c d) + phi m + dang sf)
+-- Remove parenthesis
+  === 1 + consT (Node3 b c d) m
+  + dang (Two x a) + phi (cons (Node3 b c d) m)
+  + dang sf
+  - dang (Four a b c d) - phi m - dang sf
+-- dang sf cancels, use definition of dang
+  === 1 + consT (Node3 b c d) m
+  + 0 + phi (cons (Node3 b c d) m) - 1 - phi m
+-- 1s cancel
+  === consT (Node3 b c d) m
+  + phi (cons (Node3 b c d) m) - phi m
+-- Use induction
+  ? consP (Node3 b c d) m
+  *** QED
+
+{-@ snocP :: t:FingerTree a -> x:a ->
+  { snocT t x + phi (snoc t x) - phi t <= 3 }
+@-}
+snocP :: a -> FingerTree a -> Proof
+snocP t@EmptyT x =
+    snocT t x + phi (snoc t x) - phi t
+-- Use definition of snoc/T EmptyT _
+  === 1 + phi (Single x) - phi EmptyT
+-- Use definition of phi
+  === 1 + 0 - 0 -- trivial, 1 <= 3
+  *** QED
+snocP t@(Single a) x =
+    snocT t x + phi (snoc t x) - phi t
+-- Use definition of snoc/T (Single a) _
+  === 1 + phi (Deep (One a) EmptyT (One x))
+        - phi (Single a)
+-- Use definition of phi
+  === 1 + dang (One a) + phi EmptyT
+        + dang (One x) - 0
+-- Use definition of phi and dang
+  === 1 + 1 + 0 + 1 - 0 -- trivial, 3 <= 3
+  *** QED
+snocP t@(Deep pr m (One a)) x =
+    snocT t x + phi (snoc t x) - phi t
+-- Use definition of snoc/T
+  === 1 + phi (Deep pr m (Two a x)) - phi t
+-- Use definition of phi
+  === 1 + dang pr + phi m + dang (Two a x)
+    - (dang pr + phi m + dang (One a))
+-- Remove parenthesis, use definition of dang
+  === 1 + dang pr + phi m + 0
+    - dang pr - phi m - 1
+-- everything cancels
+  === 0 -- trivial, 0 <= 3
+  *** QED
+snocP t@(Deep pr m (Two a b)) x =
+    snocT t x + phi (snoc t x) - phi t
+-- Use definition of snoc/T
+  === 1 + phi (Deep pr m (Three a b x)) - phi t
+-- Use definition of phi
+  === 1 + dang pr + phi m + dang (Three a b x)
+    - (dang pr + phi m + dang (Two a b))
+-- Remove parenthesis, use definition of dang
+  === 1 + dang pr + phi m + 0
+    - dang pr - phi m - 0
+-- phi and dang cancels
+  === 1 -- trivial, 1 <= 3
+  *** QED
+snocP t@(Deep pr m (Three a b c)) x =
+    snocT t x + phi (snoc t x) - phi t
+-- Use definition of snoc/T
+  === 1 + phi (Deep pr m (Four a b c x)) - phi t
+-- Use definition of phi
+  === 1 + dang pr + phi m + dang (Four a b c x)
+    - (dang pr + phi m + dang (Three a b c))
+-- Remove parenthesis, use definition of dang
+  === 1 + dang pr + phi m + 1
+    - dang pr - phi m - 0
+-- phi and dang cancels
+  === 2 -- trivial, 2 <= 3
+  *** QED
+snocP t@(Deep pr m (Four a b c d)) x =
+    snocT t x + phi (snoc t x) - phi t
+-- Use definition of snocT
+  === (1 + snocT m (Node3 a b c))
+  + phi (snoc t x) - phi t
+-- Use definition of snoc
+  === (1 + snocT m (Node3 a b c))
+  + phi (Deep pr (snoc m (Node3 a b c)) (Two x a))
+  - phi t
+-- Use definition of phi
+  === (1 + snocT m (Node3 a b c))
+  + dang pr + phi (snoc m (Node3 a b c))
+  + dang (Two x a)
+  - (dang pr + phi m + dang (Four a b c d))
+-- Remove parenthesis
+  === 1 + snocT m (Node3 a b c)
+  + dang pr + phi (snoc m (Node3 a b c))
+  + dang (Two x a)
+  - dang pr - phi m - dang (Four a b c d))
+-- dang pr cancels, use definition of dang
+  === 1 + snocT m (Node3 a b c)
+  + phi (snoc m (Node3 a b c)) + 0 - phi m - 1
+-- 1s cancel
+  === snocT m (Node3 a b c)
+  + phi (snoc m (Node3 a b c)) - phi m
+-- Use induction
+  ? snocP m (Node3 a b c)
+  *** QED
+
+\end{lstlisting}
+
 \end{document}

### 2021-05-22 paper/main.tex
Incorporate feedback

@@
-\documentclass[sigplan,screen]{acmart}
+%\documentclass[sigplan,screen]{acmart}
+\documentclass[sigplan,screen,review,anonymous]{acmart}
 
@@
 \author{Jan van Brügge}
-\email{jan.van-bruegge@tum.de}
-%\orcid{1234-5678-9012}
+\orcid{0000-0003-1560-7326}
 \affiliation{%
-  \institution{Technical University Munich}
+  \institution{Technical University of Munich}
   \streetaddress{Arcisstraße 21}
@@
 }
+\email{jan.van-bruegge@tum.de}
 
@@
 \begin{abstract}
-Proofs of complexity of production code is usually done in external theorem provers.
-This makes it necessary to translate the data structures to be proved from the production language into the language of the prover.
-Such a translation step may introduce errors by a mismatch in features in the two languages.
+Proofs of complexity of production code are usually done in external theorem provers.
+This makes it necessary to translate the data structures to be verified from the production language into the language of the prover.
+Such a translation step could introduce errors, for example due to a mismatch in features between the two languages.
 We show how to prove amortized complexity of data structures directly in Haskell using LiquidHaskell~\cite{liquidhaskell}.
-Besides skipping the translation step, our approach also has a didactic advantage as new learners do not need to learn a new language for their proofs and can focus on the new concepts only.
+Besides skipping the translation step, our approach also provides a didactic advantage as learners do not have to learn an additional language for their proofs and can focus on the new concepts only.
 We do not assume prior knowledge of amortized complexity as we explain the concepts and apply them in our first case study, a simple stack with multipop.
-The usefulness of our approach in a production setting is verified by the second case study, the finger tree which powers \texttt{Data.Seq} from the \textit{containers} package which is widely depended on.
-Finally we discuss the current limitations of LiquidHaskell in this area and how they might be lifted.
+The practicality of our approach in a production setting is demonstrated by our second case study: the finger tree.
+It powers \texttt{Data.Seq} from the \textit{containers} package~\cite{containers} which is widely depended on.
+Finally we discuss the current limitations of LiquidHaskell in this area and how it might be possible to improve the situation.
 \end{abstract}
@@
 
-Formally proving certain properties of production code is getting more and more common.
-For either safety critical code that is not allowed to "go wrong" or code that is fundamental and reused very often, bugs can have a widespread impact.
-Proving that the code behaves according to the specification can give an extra layer of confidence in such cases.
-Fundamental data structures are one of the most important parts of a language ecosystem and a lot of production code depends on them.
-In the Haskell ecosystem, this job is done by the \textit{containers} package~\cite{containers} which has over 5300 reverse dependencies.
-A lot of code is depending on the assumptions that the data structures perform well and adhere to the stated computational complexities from the documentation.
+Formally proving certain properties of production code is becoming more and more common.
+For safety critical code that is not allowed to "go wrong" as well as code that is fundamental and reused very often like core libraries, bugs can have a widespread impact.
+Proving that the code behaves according to the specification can provide an extra layer of confidence.
+Fundamental data structures are one of the most important parts of a programming language ecosystem and a lot of production code depends on them.
+In the Haskell ecosystem, this job is done by the \textit{containers} package which has over 5300 reverse dependencies\footnote{reverse dependecies are other packages that depend on \textit{containers}} on Hackage~\cite{hackage}, the main repository for Haskell packages.
+A lot of code depends on the assumptions that the data structures are fast and adhere to computational complexities stated in their documentation.
 Thus, proving those assumptions correct benefits a huge number of people.
 
-Usually these proofs are done in an external theorem prover, requiring an explicit translation step from the production language to the language of the theorem prover.
+Usually these proofs are done in an external theorem prover (e.g.
+Coq~\cite{complexity_coq} or Isabelle/HOL~\cite{complexity_isabelle}), requiring an explicit translation step from the production language to the language the theorem prover uses.
 This translation step has a few disadvantages.
-First, the semantics of the target language may differ from the semantics of the production language.
-This is especially true for Haskell with its pervasive laziness, compared with most theorem provers that operate with a strict language.
-Second, such a translation step may introduce subtle differences or bugs, either by accident in the manual case or through a bug in the automated case.
-Lastly, the target language may not support certain features present in the source language such as polymorphic recursion in data types.
-In those cases, the data structure needs to be encoded in the target language, again opening the doors for small and subtle errors.
-Because the code that is actually used to conduct the proofs might me far removed from the source code that is running in production, it is often hard to reconnect the proofs back to the source code.
+First, the semantics of the target language could differ from the semantics of the production language.
+This is especially true for Haskell with its pervasive laziness, compared to most theorem provers that operate on a strict language.
+Second, such a translation step could introduce subtle differences or bugs, either by accident when doing a manual translation or through a bug when an automated translation tool is used.
+Lastly, the target language may not support certain features present in the source language (e.g.
+polymorphic recursion in data types, see section~\ref{sec:fingertree}).
+In those cases, the missing features have to be emulated in the target language, again opening the door for small and subtle errors.
+Because the code that is actually used in the formal proofs differs from the source code that is running in production, it is often hard to reconnect the proofs back to the original production code.
 
 A different approach comes in the form of LiquidHaskell~\cite{liquidhaskell}.
-It is a refinement type~\cite{refinement_types} checker for Haskell that since 2020 is implemented as a type checker plugin for GHC, the main Haskell compiler used in the industry.
-It allows to specify additional constraints on types like "integers greater than zero" instead of just "integers".
-These constraints are propagated through the code and checked by an external SMT prover.
+It is a refinement type~\cite{refinement_types} checker plugin\footnote{before 2020, LiquidHaskell was implemented as a standalone executable, it now being a GHC plugin~\cite{lh_plugin} drastically improves usability} for GHC, the main Haskell compiler used in industry.
+It allows specifing additional constraints on types, e.g.
+"integers greater than zero" instead of just "integers".
+These constraints are propagated through the code and checked by an external SMT solver.
 With a handful of proof combinators defined by LiquidHaskell, it is possible to use Haskell directly as a theorem prover, enabling intuitive equational reasoning within the language~\cite{tpfa}.
-This avoids any translation step and makes it easy to keep proof and implementation in sync, as any failure to do so will result in an immediate compiler error.
-Another benefit is that the reader or writer of a proof does not need to learn a new language, the production language is the proof language.
-
-This paper thus:
+This avoids any translation steps and makes it easy to keep implementation and its respective proof in sync, as any failure to do so will result in a compiler error.
+Another benefit is that the reader or author of a proof does not need to learn an additional language, the production language is the proof language.
+
+In this paper, we:
 \begin{itemize}
-\item{Assumes no prior knowledge of amortized complexity by recapping the theoretical background}
-\item{Shows the reasoning infrastructure that is needed to prove amortized complexity with LiquidHaskell}
-\item{Applies this technique to the finger tree, the data structure behind \texttt{Data.Seq} from \textit{containers}, as a case study}
-\item{Discusses the usefulness of LiquidHaskell for such proofs compared with other theorem provers}
+\item{Assume no prior knowledge of amortized complexity as we recap the theoretical background (section~\ref{sec:complexity})}
+\item{Show the reasoning infrastructure that is needed to prove amortized complexity using LiquidHaskell (sections~\ref{sec:liquidhaskell} and~\ref{sec:stack})}
+\item{Apply this technique to \textit{finger trees}, the data structure behind \texttt{Data.Seq} from \textit{containers} (see our second case study in section~\ref{sec:fingertree})}
+\item{Discuss the usefulness of LiquidHaskell for such proofs compared with other theorem provers (sections~\ref{sec:limitations} and~\ref{sec:conclusion})}
 \end{itemize}
@@
 
-It is common to check how many time units an operation on a data structure takes in relation to the number of elements in the data structure.
-As the performance can vary wildly for the same operation depending on the state of data structure --- a linear search takes $1$ time unit if the searched element is at the front, while it takes $n$ time units if it is the last element --- usually one uses the worst case for an operation as a metric and uses Landau notation to specify it.
-With our linear search example we would say that it is in \O{n} because it will not perform worse than $n$ time units.
+It is common practise to check how long an operation on a data structure takes in relation to the number of elements in the data structure.
+As the performance can vary wildly for the same operation depending on the state of data structure, usually the worst case for each operation is used as a metric and specified using \textit{Landau notation} ("big O").
+For example, a linear search takes $1$ unit of time if the searched element is at the front, while it takes $n$ units of time if it is the last element.
+Thus we would say that a linear search is in the complexity class of \O{n} because it will not take longer than $c \cdot n$ units of time for some fixed constant $c$.
 
 But often this worst case is too pessimistic.
-When an operation changes the data structure, for example by removing elements, consecutive uses of operation may be cheaper.
+When an operation changes the data structure, for example by removing elements, consecutive operations might be cheaper.
 Or to use an expensive operation it might be necessary to have a certain amount of cheap operations.
-In this case one uses \textit{amortized complexity} which does not only look at a single operation in isolation, but on chains of operations.
-In particular, it assumes that that the intermediate data structures are not shared, i.e.
-they are not persisted.
-There are multiple methods to proof amortized complexity.
+In such cases one uses \textit{amortized complexity}, which does not only look at a single operation in isolation, but at chains of operations.
 
 As a running example we will use a stack with \textit{push} and \textit{multipop} operations.
-The former allocates a new element and a references to the rest of the stack and thus taking a constant amount of time.
+The former allocates a new element and a reference to the rest of the stack and thus takes a constant amount of time.
 The latter removes the first $k$ elements from the stack.
-In the worst case, we remove all the elements, making \texttt{multipop} \O{n}.
-One can already see that this is too pessimistic, as after removing all elements from the stack, the next \texttt{multipop} will do nothing.
-In fact, every element we remove needs to be added to the stack at some point, so we can spread out the cost of \texttt{multipop} over many pushes.
+In the worst case, it removes all the elements, making \texttt{multipop} \O{n}.
+One can already see that this is too pessimistic, because after removing all elements from the stack, the next \texttt{multipop} will do nothing.
+In fact, to be able to remove an element, it has had to be pushed to the stack at some point, so we can spread out the cost of \texttt{multipop} over those pushes.
 There are multiple techniques to prove amortized complexity, we present the two most popular ones.
@@
 
-The banker's method works by having an bank account that can store time units.
-Operations have to pay their actual cost and may additionally pay extra into the account.
-Expensive operations can then use the stored time units in the bank to pay for themselves.
+The banker's method works by having a bank account that can store time units.
+Operations have to pay their actual runtime cost and may additionally deposit into or withdraw from the account.
+The sum of the actual cost and the deposit/withdrawal is the \textit{amortized cost} of the operation.
+The idea is that cheap operations pay extra time units, making their amortized cost higher than their actual cost.
+The goal is to keep the difference between the two costs to a constant factor because that means the operation still retains their (cheap) complexity class.
+Expensive operations on the other hand can use the stored time to pay for their actual cost, hopefully putting them in a cheaper complexity class.
 In our example a \texttt{push} has to pay one actual time unit and it will also pay one extra time unit into the bank account.
-This means that the bank account always contains as many time units as there are elements on the stack.
-That way, any call to \texttt{multipop} can use the bank account to pay for itself and the account will never become negative.
+The difference between amortized cost and actual cost is constant, so \texttt{push} is still in \O{1}.
+Because for every element pushed we deposit one extra time unit into the bank, the account always contains as many time units as there are elements on the stack.
+As \texttt{multipop} can not remove more elements than there are on the stack, withdrawing $k$ time units to pay for the actual cost will never make the account become negative.
+Because withdrawing from the bank account is enough to pay for the complete actual cost of \texttt{multipop} regardless of the choice of $k$, its \textit{amortized} runtime is in \O{1}.
 
@@
 
-The banker's method is hard to use for a formal proof because the bank account acts as state between the operations.
-A simpler method is the physicist's method that does not need such an account.
+While the banker's method is easy to visualize, it is hard to use in a formal proof because the bank account acts as state between the operations.
+A simpler method is the physicist's method, which does not need such an account.
 The idea is to define a potential $\Phi$ ("phi") for the data structure.
-This potential should be $0$ for an empty data structure and $\ge 0$ for all other states.
-Intuitively the potential will declare how much time needs to be saved up to pay for expensive operations.
-We then define the amortized cost of an operation as $c + \Phi(h') - \Phi(h)$ where $c$ is the actual cost for the operation, $h$ is the data structure before the operation and $h'$ is the data structure afterwards.
-Looking at a chain of operations $1, 2, ..., n$, the sum of this amortized time is an overestimation of the actual cost by $\Phi(h_n)$ --- because $\Phi(h_0)$ is by definition $0$ --- and thus an upper bound on the actual cost (see equation~\ref{eq:sum}).
+This potential is defined to be $0$ for an empty data structure and $\ge 0$ for all other states.
+Intuitively, the potential represents how much time needs to be saved up to pay for expensive operations.
+We then define the amortized cost of an operation as $c + \Phi(h') - \Phi(h)$ where $c$ is the actual time cost of the operation, $h$ is the data structure before the operation and $h'$ is the data structure afterwards.
+Looking at a chain of operations $1, 2, ..., n$, the sum of the amortized time is an overestimation of the actual cost by $\Phi(h_n)$ and thus an upper bound on the actual cost (see equation~\ref{eq:sum}, remember that by definition $\Phi(h_0) = 0$).
 
@@
 
-For our stack we can define the potential to be the height of the stack.
+For our stack we define the potential to be the height of the stack.
 This fulfils the requirements as the empty stack has a potential of zero and all other states have a potential greater or equal to zero.
@@
 
-\section{LiquidHaskell for theorem proving}
+\section{LiquidHaskell for theorem proving}\label{sec:liquidhaskell}
 
 As described by Vazou et.
-al.~\cite{tpfa}, one can use LiquidHaskell to do complex equational reasoning that is machine checked by the compiler.
+al.~\cite{tpfa}, one can use LiquidHaskell to do complex equational reasoning that is machine checked.
 It uses refinement types~\cite{refinement_types} that add additional constraints on top of normal Haskell data types which are checked by an external SMT solver.
 By default LiquidHaskell uses the z3~\cite{z3} solver, but it also supports cvc4~\cite{cvc4}.
-For example we can declare the type of integers greater or equal to zero with a type synonym.
-LiquidHaskell annotions use comments with an @ sign, so that the code still compiles even without LiquidHaskell.
-It also allows to name arguments, so they can be referred to later in the refinement.
-At the moment to define functions with refinement, one has to copy the Haskell type signature into a LiquidHaskell comment and add the refinements there.
-A future version that is more deeply integrated into the Haskell compiler may not need this code duplication.
+For example, we can declare the type of integers greater or equal to zero using a type synonym:
 
@@
 
+LiquidHaskell annotations are multiline Haskell comments with "at" signs, such that the code still compiles even without LiquidHaskell.
+It also allows the user to name argument types, so they can be referred to later.
+At the time of writing one has to copy the Haskell type signature into a LiquidHaskell comment and add the refinements there.
+A future version that is integrated deeper into the Haskell compiler might not need this code duplication.
+
 In LiquidHaskell, all functions need to terminate.
-Usually a suitable termination metric is automatically guessed, but in some more complicated cases, we can also manually supply this metric (see section~\ref{sec:append} for an example of this).
-A simple recursive function can be marked as \textit{measure} which allows the use of the function in those termination metrics as well as in the refinement itself.
+Usually a suitable termination metric is automatically deduced, but in some more complex cases, we can also manually supply this metric (see section~\ref{sec:append} for an example).
 With the termination, LiquidHaskell is able to generate an induction principle for the function that can be used in a proof.
 Because LiquidHaskell is just Haskell, this induction amounts to a recursive call to the proof and is also used when defining ordinary Haskell functions with refinement on their types.
-For example the fact that the length of a list is never negative.
+For example the fact that the length of a list is never negative:
 
@@
     -- by induction: length xs >= 0
-    -- so 1 + length xs >= 0
-\end{lstlisting}
-
-LiquidHaskell also provides a set of proof combinators and a \texttt{Proof} type.
+    -- thus 1 + length xs >= 0
+\end{lstlisting}
+
+To use functions in the refinements, one has to \textit{reflect}~\cite{reflection} them to the logic level.
+For a limited class of functions called \textit{measures}, this can be done while retaining full automatic reasoning.
+A measure is a function that takes only one argument which is an algebraic data type, has one equation per constructor of this data type and which only uses arithmetic functions and other measures in their right hand sides.
+Other function still can be used in refinements when annotated with \texttt{reflect}, but the SMT solver is not able to check those refinement automatically.
+The user always has to provide the proofs themselves.
+
+To do so, LiquidHaskell provides a set of proof combinators and a \texttt{Proof} type.
 This type is just a type alias for the normal Haskell unit type.
 The reasoning comes from the \texttt{===} operator that asserts that both sides of the equality are truly equal.
-As a simple example we can proof that the length of an \texttt{append} function is the sum of the lengths of the two original lists.
-As append is not a \textit{measure} (it does not return a natural number), we have to use \textit{reflection}~\cite{reflection} to lift it to the logic level of LiquidHaskell.
-The proof itself is done by case distinction on the first argument.
-This distinction is done by pattern matching on it like one would in a normal Haskell function.
-We start with stating the left hand side of the equation how it is in the refinement type and try to get to the right hand side of the equality by simplification and induction.
+For a simple example we prove that the length of \texttt{append} applied to two lists is the sum of the lengths of the two lists.
+Because \texttt{append} takes more than one argument, it is not a measure and thus we need to do the proof manually.
+
+First, we need to do a case distinction on the first argument.
+This is done by pattern matching on it like in a normal Haskell function.
+The rest of the proof is done by applying the definitions of the used functions, arithmetic equalities and induction:
 
@@
 
-The triple star operator (\texttt{***}) and the \texttt{QED} type at the end are just to cast our proof to Haskell's unit type and to make the proof look nicer.
+The triple star operator (\texttt{***}) and the \texttt{QED} type at the end are just there to cast our proof to Haskell's unit type and improve readability.
 
@@
 
-To continue with the example from section~\ref{sec:complexity}, we want to prove that $multipop$ has an amortized complexity of \O{1}.
-This data structure and its operations could be defined like this in Haskell.
+To continue with the example from section~\ref{sec:complexity}, we prove that $multipop$ has an amortized complexity of \O{1}.
+This data structure and its operations are defined like the following in Haskell.
 We already add some refinements to assert that $k$ is not negative.
+Note that normally, \texttt{multipop} would return the removed elements in addition to the new stack, but as this is irrelevant to the proof and only makes it more verbose we omit this in our definition.
 
@@
 
+{-@ reflect push @-}
 push :: a -> Stack a -> Stack a
-push = Elem
-
+push x s = Elem x s
+
+{-@ reflect multipop @-}
 {-@ multipop :: Nat -> Stack a -> Stack a @-}
@@
 
-To formally prove this with the physicist's method, we first need to define a potential $\Phi$ and a timing function for each of our operations that gives us the actual cost of that operation.
+To formally prove the amortized complexity with the physicist's method, we first need to define a potential $\Phi$ and a timing function for each of our operations that gives us the actual time cost of that operation.
 As seen in section~\ref{sec:physicist}, we use the height of the stack as our potential.
 Thus we need to define a function that calculates the height of a stack.
-To define the timing functions, we copy the code for the operation, replace constant time steps with $1$ and adjust the recursion.
-
-\begin{lstlisting}
+To define the timing functions, we copy the code for the operation, replace constant time steps with $1$ and adjust the recursion:
+
+\begin{lstlisting}
+{-@ reflect phi @-}
 {-@ phi :: Stack a -> Nat @-}
@@
 
+{-@ reflect pushT @-}
 {-@ pushT :: a -> Stack a -> Nat @-}
@@
 
+{-@ reflect multipopT @-}
 {-@ multipopT :: Nat -> Stack a -> Nat @-}
@@
 
-To use those functions in a LiquidHaskell proof, we also need to reflect them into the logic level~\cite{reflection} by adding \texttt{reflect} annotations.
-
-\begin{lstlisting}
-{-@ reflect push @-}
-{-@ reflect pushT @-}
-{-@ reflect multipop @-}
-{-@ reflect multipopT @-}
-{-@ reflect phi @-}
-\end{lstlisting}
-
 First we will prove the complexity of \texttt{push}.
-While the amortized cost is the same as the actual cost, we need to make sure that our potential function does not cause those two to diverge.
-As LiquidHaskell does not allow quantifiers to ensure termination of the SMT solver, we need to guess an upper limit that our amortized runtime will have.
-As seen earlier in section~\ref{sec:banker}, we can pay two time units for a push to amortize the pops, so we will use $2$ as our guess.
-To specify our proof goal, we need to name the parameters in the LiquidHaskell annotation.
-After than we can use equational reasoning to proof our goal.
+While the amortized cost is in the same complexity class as the actual cost, we need to make sure that our potential function does not cause those two to diverge.
+LiquidHaskell does not allow quantifiers in refinement, because for functions other than measures, instanciating quantifiers in the SMT solver has unpredictable performance~\cite{lh_quantifiers}.
+This means we have to guess an upper limit for our amortized runtime.
+As seen earlier in section~\ref{sec:banker}, we can pay two time units for a push to amortize the pops, so we will use $2$ as our guess:
 
@@
 
-To proof complexity of \texttt{multipop}, we will need to make a case distinction between the two base cases and the recursive case.
+To prove the complexity of \texttt{multipop}, we will need to distinquish between the two base cases and the recursive case.
 With LiquidHaskell this is just a normal pattern match.
-Leaving the recursive case undefined for now, proving the base cases is very similar to \texttt{push}.
+Leaving the recursive case undefined for now, proving the base cases is very similar to \texttt{push}:
 
@@
 To avoid cluttering the proof of the recursive case, we use an \textit{as pattern} to alias \texttt{Elem x s} to \texttt{xs}.
-The first steps again are applying the definitions of the used functions and simplifying the result.
-In the end we use the \texttt{?} operator to add the induction hypothesis to the proof step and thus complete the proof.
-
-\begin{lstlisting}
+The first steps are again applying the definitions of the used functions and simplifying the result.
+In the end we use the \texttt{?} operator to add the induction hypothesis to the proof step and thus complete the proof:
+
+\begin{lstlisting}
+multipopP k xs@(Elem x s) = multipopT k xs
+    + phi (multipop k xs) - phi xs
+-- Use definition of multipopT for recursive case
+  === (1 + multipopT (k - 1) s)
+    + phi (multipop k xs) - phi xs
+-- Use definition of multipop for recursive case
+  === (1 + multipopT (k - 1) s)
+    + phi (multipop (k - 1) s) - phi xs
+-- Use definition of phi (Elem _ _)
+  === (1 + multipopT (k - 1) s)
+    + phi (multipop (k - 1) s) - (1 + phi s)
+-- Remove parentheses
+  === 1 + multipopT (k - 1) s
+    + phi (multipop (k - 1) s) - 1 - phi s
+-- 1s cancel out
+  === multipopT (k - 1) s
+    + phi (multipop (k - 1) s) - phi s
+-- use induction hypothesis
+  ? multipopP (k - 1) s
+  *** QED
+\end{lstlisting}
+
+\subsection{Finger Tree}\label{sec:fingertree}
+
+To show that this approach scales to more complicated data structures that are widely used in the industry, we will prove the complexity of finger trees, the data structure behind \texttt{Data.Seq} from \textit{containers}~\cite{containers}.
+It was originally described by Hinze and Paterson~\cite{fingertrees}, the implementation from \textit{containers} closely follows this description.
+The following definition is taken directly from \textit{containers}, but slightly modified to remove the size parameter.
+LiquidHaskell caused problems in the proofs when using type classes because at the time of writing it is not possible to replace a call to a polymorphic function with the implementation in its instance.
+None of the function we use in the proofs use this size parameter, so we omit it.
+We will discuss such limitations more in depth in section~\ref{sec:limitations}.
+
+\begin{lstlisting}
+data FingerTree a = Empty
+  | Single a
+  | Deep (Digit a) (FingerTree (Node a)) (Digit a)
+
+data Node a = Node2 a a | Node3 a a a
+
+data Digit a = One a
+  | Two a a
+  | Three a a a
+  | Four a a a a
+\end{lstlisting}
+
+The finger tree has between one and four \textit{digits} on either side to enable cheap \texttt{cons} and \texttt{snoc}.
+As shown by Claessen~\cite{fingertrees_new}, this would also work with a maximum of just three digits, but we will follow the definition from \textit{containers}.
+The most interesting property of the data structure is the polymorphic recursion.
+It makes the data structure type safe as it guarantees equal nesting of \texttt{Node} types on both sides of the spine.
+It is also what makes it difficult to verify in other languages, because for example the both Isabelle/HOL~\cite{isabelle} and the Lean~\cite{lean} theorem provers do not directly support polymorphic recursion.
+In these languages, it is necessary to define the type in such a way that it would allow uneven nesting of \texttt{Node}s and then "carve out" the subset of valid finger trees.
+Other provers like Coq~\cite{coq} support polymorphic recursion, allowing a more direct one to one translation.
+
+\subsubsection{Cons and Snoc}
+
+Because the finger tree is a symmetric tree, \texttt{cons} and \texttt{snoc} are very similar.
+For this reason we show an explicit proof only for \texttt{cons} and provide the proof for \texttt{snoc} in the appendix.
+First we again copy the definition of \texttt{cons} from the \textit{containers} package, adjusting it slightly to remove the size information that would require the type class.
+In order to later use the function in a proof we add a \texttt{reflect} pragma.
+
+\begin{lstlisting}
+{-@ reflect cons @-}
+cons :: a -> FingerTree a -> FingerTree a
+cons a EmptyT = Single a
+cons a (Single b) =
+  Deep (One a) EmptyT (One b)
+cons a (Deep (One b) m sf) =
+  Deep (Two a b) m sf
+cons a (Deep (Two b c) m sf) =
+  Deep (Three a b c) m sf
+cons a (Deep (Three b c d) m sf) =
+  Deep (Four a b c d) m sf
+cons a (Deep (Four b c d e) m sf) =
+  Deep (Two a b) (cons (Node3 c d e) m) sf
+\end{lstlisting}
+
+The only complex case is if we already have four digits at the front, as in all other cases we can simply grow the digit in size.
+In the complex case, we put the new data in front and push the "overflow" into the recursion.
+The next step is to find a valid potential function $\Phi$.
+For this, we check which states of the data structure make \texttt{cons} expensive.
+As evident by the definition above, this is the case when the digit in the front is already full.
+So we can define the \textit{danger} of a digit as every state that will make the operation expensive.
+To also support \texttt{uncons}, the inverse operation, we not only mark a full digits as dangerous, but also one that has only one element inside (as this would make \texttt{uncons} go into the recursion).
+We then define the potential to be the sum of the danger in the tree.
+
+\begin{lstlisting}
+{-@ reflect dang @-}
+{-@ dang :: Digit a -> Nat @-}
+dang :: Digit a -> Int
+dang One{} = 1
+dang Two{} = 0
+dang Three{} = 0
+dang Four{} = 1
+
+{-@ reflect phi @-}
+{-@ phi :: FingerTree a -> Nat @-}
+phi :: FingerTree a -> Int
+phi EmptyT = 0
+phi Single{} = 0
+phi (Deep pr m sf) =
+  dang pr + phi m + dang sf
+\end{lstlisting}
+
+The timing function for \texttt{cons} is similar to the timing functions for the stack just a copy of the original function where all non-recursive cases return $1$.
+
+\begin{lstlisting}
+{-@ reflect consT @-}
+{-@ consT :: a -> FingerTree a -> Nat @-}
+consT :: a -> FingerTree a -> Int
+consT _ EmptyT = 1
+consT _ Single{} = 1
+consT _ (Deep One{} _ _) = 1
+consT _ (Deep Two{} _ _) = 1
+consT _ (Deep Three{} _ _) = 1
+consT _ (Deep (Four _ c d e) m _) =
+  1 + consT (Node3 c d e) m
+\end{lstlisting}
+
+With all the required preliminaries defined, we start proving the amortized complexity.
+Just like in the stack example (see section~\ref{sec:stack}) we need to guess a constant that is an upper bound of our amortized run time.
+Because we do not have a simple mental model like with the stack, we will delay this choice until we have proven the base cases.
+We will also use \textit{as patterns} again to avoid repetition.
+
+\begin{lstlisting}
+{-@ consP :: x:a -> t:FingerTree a ->
+  { consT x t + phi (cons x t) - phi t <= c }
+@-}
+consP :: a -> FingerTree a -> Proof
+consP x t@EmptyT =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of cons/T _ EmptyT
+  === 1 + phi (Single x) - phi EmptyT
+-- Use definition of phi
+  === 1 + 0 - 0
+-- 1 must be <= c, so c >= 1
+  *** QED
+consP x t@(Single a) =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of cons/T _ (Single a)
+  === 1 + phi (Deep (One x) EmptyT (One a))
+        - phi (Single a)
+-- Use definition of phi
+  === 1 + dang (One x) + phi EmptyT
+        + dang (One a) - 0
+-- Use definition of phi and dang
+  === 1 + 1 + 0 + 1 - 0
+-- 3 must be <= c, so c >= 3
+  *** QED
+\end{lstlisting}
+
+We omit the proof cases for one to three digits, as they work exactly as the other two base cases.
+The constant however will not change in the other base cases, so $3$ really is our upper bound to use here.
+The interesting case is when there are four digits.
+In this case we need to use induction to prove our goal:
+
+\begin{lstlisting}
+consP x t@(Deep (Four a b c d) m sf) =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of consT
+  === (1 + consT (Node3 b c d) m)
+  + phi (cons x t) - phi t
+-- Use definition of cons
+  === (1 + consT (Node3 b c d) m)
+  + phi (Deep (Two x a) (cons (Node3 b c d) m) sf)
+  - phi t
+-- Use definition of phi
+  === (1 + consT (Node3 b c d) m)
+  + dang (Two x a) + phi (cons (Node3 b c d) m)
+  + dang sf
+  - (dang (Four a b c d) + phi m + dang sf)
+-- Remove parenthesis
+  === 1 + consT (Node3 b c d) m
+  + dang (Two x a) + phi (cons (Node3 b c d) m)
+  + dang sf
+  - dang (Four a b c d) - phi m - dang sf
+-- dang sf cancels, use definition of dang
+  === 1 + consT (Node3 b c d) m
+  + 0 + phi (cons (Node3 b c d) m) - 1 - phi m
+-- 1s cancel
+  === consT (Node3 b c d) m
+  + phi (cons (Node3 b c d) m) - phi m
+-- Use induction
+  ? consP (Node3 b c d) m
+  *** QED
+\end{lstlisting}
+
+\subsubsection{Append}\label{sec:append}
+
+Similar to the \texttt{push} operation on the stack, even though \texttt{append} has the same amortized complexity as its normal complexity, we need to show that it never increases the potential more than a logarithmic amount.
+Contrary to the simplified definition of Claessen~\cite{fingertrees_new}, the \textit{containers} package uses two sets of mutually recursive functions to implement \texttt{append}.
+One set of functions called \texttt{appendTreeX} handles the base cases where one of the two trees is either empty, or consists of a singular element by using \texttt{cons} and \texttt{snoc}.
+In the recursive case it calls the \texttt{addDigitsX} functions that group arguments as \texttt{Node}s and in turn call the \texttt{appendTreeX} functions.
+Because the \texttt{addDigitsX} functions do not pattern match on their tree arguments at all, LiquidHaskell can not find a termination metric automatically.
+For this, we need to track the arguments across functions calls because \texttt{appendTreeX} has a strictly decreasing argument.
+We can do so by adding a ghost parameter to order the functions for LiquidHaskell.
+This order is defined to be smaller for all \texttt{appendTreeX} functions than the \texttt{addDigitsX} functions.
+Additionally we need to set the case expansion to a high number such that all pattern matches are evaluated, else the termination proof would fail.
+The following snippet provides an example with most cases and function omitted for readability.
+
+\begin{lstlisting}
+{-@ LIQUID --max-case-expand=1000 @-}
+
+{-@ appendTree1 :: a:FingerTree (Node a) ->
+  Node a -> FingerTree (Node a) ->
+  FingerTree (Node a) / [size a, 1] @-}
+{-@ reflect appendTree1 @-}
+appendTree1 :: FingerTree (Node a) -> Node a ->
+    FingerTree (Node a) -> FingerTree (Node a)
+appendTree1 EmptyT !a xs = a `consTree` xs
+appendTree1 xs !a EmptyT = xs `snocTree` a
+appendTree1 (Single x) !a xs =
+    x `consTree` a `consTree` xs
+appendTree1 xs !a (Single x) =
+    xs `snocTree` a `snocTree` x
+appendTree1 (Deep pr1 m1 sf1) a (Deep pr2 m2 sf2) =
+    Deep pr1 (addDigits1 m1 sf1 a pr2 m2) sf2
+
+{-@ addDigits1 :: a:FingerTree (Node (Node a)) ->
+  Digit (Node a) -> Node a -> Digit (Node a) ->
+  FingerTree (Node (Node a)) ->
+  FingerTree (Node (Node a)) / [size a, 5] @-}
+{-@ reflect addDigits1 @-}
+addDigits1 :: FingerTree (Node (Node a)) ->
+  Digit (Node a) -> Node a -> Digit (Node a) ->
+  FingerTree (Node (Node a)) ->
+  FingerTree (Node (Node a))
+addDigits1 m1 (One a) b (One c) m2 =
+  appendTree1 m1 (Node3 a b c) m2
+addDigits1 m1 (One a) b (Two c d) m2 =
+  appendTree2 m1 (Node2 a b) (Node2 c d) m2
+-- further cases omitted
+\end{lstlisting}
+
+This case expansion results in an explosion of constraints for the SMT solver to check.
+While the whole proof for \texttt{cons} including the defined functions generates fewer than 100 constraints, this termination proof generates over 3000.
+This leads to extremely high compile times which is further complicated by the need to define a timing function.
+The function needs to follow this mutually recursive structure and generates an additional 8000 constraints, such that loading the file takes several minutes.
+Another problem is the logarithmic runtime itself.
+SMT solvers are great with linear arithmetic, but are significantly slower when tasked with multiplication.
+Division is usually left as an unevaluated function.
+While it is possible to transform the proof goal to convert logarithms into powers, it still requires assumptions about basic arithmetic rules on the logic level.
+
+For those reasons, using LiquidHaskell to proof the complexity with the physicist's method is not feasible (see section~\ref{sec:limitations} for a discussion of those limitations).
+Claessen~\cite{fingertrees_new} does not formally prove this in his functional pearl as well, but provides a different reasoning to why \texttt{append} is logarithmic.
+The potential only depends on the spine of the tree, so a new layer in the tree only adds at most a constant amount to the potential ($2$ if both digits are dangerous) making it grow linearly.
+Because the nodes in the digits always at least double (in the case of two \texttt{Node2}) the overall size of the tree grows exponentially.
+Thus the potential grows logarithmic in terms of the size of the tree.
+
+\section{Limitations of our approach}\label{sec:limitations}
+
+\subsection{Definition of the timing function}
+
+The analysis could lead to wrong results is if the timing function does not directly mirror the original function.
+When the implementation changes and the timing function does not, it might be possible for the proofs to still pass despite being invalid.
+While this is unlikely, given how verbose proofs with LiquidHaskell are, it is not negligible.
+
+If it is possible to change the implementation, it would be possible to factor out the recursion and share it between the implementation and the timing function.
+This however would clutter the implementation and most likely decrease performance as it is often the case with recursion schemes.
+
+A better solution is to derive the timing function using meta programming in the form of \textit{TemplateHaskell}.
+Given the name of the function in the implementation, TemplateHaskell would allow to create a new function with a configurable suffix.
+This can be done by checking if the right hand sides of the original function contain recursive calls to the given name.
+If not, replace the right hand side with $1$, otherwise change the recursion to use the name with the suffix and add $1$ at the end.
+The resulting timing function is then be spliced back into the source code:
+
+\begin{lstlisting}
+cons :: a -> FingerTree a -> FingerTree a
+cons a EmptyT = Single a
+cons a (Deep (Four b c d e) m sf) =
+    Deep (Two a b) (cons (Node3 c d e) m) sf
+-- other cases omitted
+
+$(mkTimingFn "T" ''cons)
+-- generates this; variable names omitted:
+-- consT :: a -> FingerTree a -> Int
+-- consT _ EmptyT = 1
+-- consT _ (Deep (Four _ c d e) m _) =
+--    consT (Node3 c d e) m + 1
+-- other cases omitted
+\end{lstlisting}
+
+\subsection{LiquidHaskell for verification}
+
+As shown in earlier sections, at the time of writing LiquidHaskell has some limitations with regards to proving more complex functions.
+First and foremost the support for type classes and Haskell \texttt{newtype}s is insufficient.
+Currently, LiquidHaskell only supports measure classes, which need to be defined by recursion over a single data type similar to a normal measure.
+For this reason it is not possible to use them for the size parameter in finger trees (see section~\ref{sec:fingertree}).
+There are several open tickets concerning the type class support on the LiquidHaskell bug tracker, so we expect this to improve in the future.
+
+Another road block was the combinatorial explosion when generating the constraints for the termination of \texttt{append} (see section~\ref{sec:append}).
+It might be possible to improve the constraint generation and thus create less work for the SMT solver.
+LiquidHaskell also provides an option to disable the termination checker, but with this option the compile time surprisingly did not change.
+Similarly, several other options that are supposed to improve performance like \texttt{{-}{-}diff}, which enables incremental type checking or \texttt{{-}{-}fast}, which disables inference of polymorphic refinement, did either not affect performance at all or (in the case of \texttt{{-}{-}fast}) made it even worse.
+
+\section{Conclusion}\label{sec:conclusion}
+
+Despite the current limitations of LiquidHaskell preventing more complex functions from being formally proven, it drastically improves the readability of proofs compared to other theorem provers.
+The explicit equational reasoning of LiquidHaskell makes following the proof a lot easier than in tactic based provers like Coq and Lean.
+While Isabelle/HOL also features equational reasoning with Isar~\cite{isar}, it does not support polymorphic recursion which makes the definition of our second example a lot harder.
+
+While at the moment amortized complexity of \O{1} seems to be the most feasable to prove, we expect the tooling around LiquidHaskell to improve a lot in the future, be it by incremental changes like better support for type classes, or by more complicated means like a deeper embedding into a future dependently typed Haskell core.
+The recent improvements in the usability of LiquidHaskell --- that were achieved by turning the previously standalone executable into a GHC plugin~\cite{lh_plugin} --- are already a big step into this future by drastically reducing the barrier of entry.
+
+%%
+%% The acknowledgments section is defined using the "acks" environment
+%% (and NOT an unnumbered section). This ensures the proper
+%% identification of the section in the article metadata, and the
+%% consistent spelling of the heading.
+%\begin{acks}
+%\end{acks}
+
+%%
+%% The next two lines define the bibliography style to be used, and
+%% the bibliography file.
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{references}
+
+%%
+%% If your work has an appendix, this is the place to put it.
+\appendix
+
+\section{Full proof code for the stack}
+
+\begin{lstlisting}
+import Language.Haskell.Liquid.ProofCombinators
+
+data Stack a = Empty | Elem a (Stack a)
+
+{-@ reflect push @-}
+push :: a -> Stack a -> Stack a
+push = Elem
+
+{-@ multipop :: Nat -> Stack a -> Stack a @-}
+multipop :: Int -> Stack a -> Stack a
+multipop _ Empty = Empty
+multipop 0 s = s
+multipop n (Elem _ s) = multipop (n - 1) s
+
+{-@ reflect phi @-}
+{-@ phi :: Stack a -> Nat @-}
+phi :: Stack a -> Int
+phi Empty = 0
+phi (Elem _ s) = 1 + phi s
+
+{-@ reflect pushT @-}
+{-@ pushT :: a -> Stack a -> Nat @-}
+pushT :: a -> Stack a -> Int
+pushT _ _ = 1
+
+{-@ reflect multipopT @-}
+{-@ multipopT :: Nat -> Stack a -> Nat @-}
+multipopT :: Int -> Stack a -> Int
+multipopT _ Empty = 1
+multipopT 0 _ = 1
+multipopT n (Elem _ s) = 1 + multipopT (n - 1) s
+
+{-@ pushP :: x:a -> s:Stack a ->
+  { pushT x s + phi (push x s) - phi s <= 2 }
+@-}
+pushP :: a -> Stack a -> Proof
+pushP x s = pushT x s + phi (push x s) - phi s
+-- Use definition of push and pushT
+  === 1 + phi (Elem x s) - phi s
+-- Use definition of phi (Elem _ _)
+  === 1 + (1 + phi s) - phi s
+-- phi s gets canceled, 1 + 1 <= 2
+  *** QED
+
+{-@ multipopP :: k:Nat -> s:Stack a ->
+  { multipopT k s + phi (multipop k s)
+      - phi s <= 2 }
+@-}
+multipopP :: Int -> Stack a -> Proof
+multipopP 0 s = multipopT 0 s
+    + phi (multipop 0 s) - phi s
+-- Use definition of multipop/T 0 _
+  === 1 + phi s - phi s
+-- phi s cancels, 1 <= 2
+  *** QED
+multipopP k Empty = multipopT k Empty
+    + phi (multipop k Empty) - phi Empty
+-- Use definition of multipop/T _ Empty
+  === 1 + phi Empty - phi Empty
+-- phi Empty cancels, 1 <= 2
+  *** QED
 multipopP k xs@(Elem x s) = multipopT k xs
@@
 
-\subsection{Finger Tree}
-
-To show that this approach scales also to more complicated data structures that are widely used in the industry, we will prove the complexity of a finger tree, the data structure behind \texttt{Data.Seq} from \textit{containers}~\cite{containers}.
-It is originally described by Hinze and Paterson~\cite{fingertrees}, which the implementation from \textit{containers} closely follows.
-The following definition is taken directly from \textit{containers}, but slightly modified to remove the size parameter.
-LiquidHaskell caused problems in the proofs when using type classes, so because the functions we will use do not use the size we omitted it.
-We will discuss such limitations more in depth in section~\ref{sec:limitations}.
+\section{Full proof code of the finger tree}\label{ap:fingertree}
 
@@
   | Four a a a a
-\end{lstlisting}
-
-The finger tree has between one and four \textit{digits} on either side to enable cheap \texttt{cons} and \texttt{snoc}.
-As shown by Claessen~\cite{fingertrees_new}, this would also work with a maximum of just three digits, but we will go with the definition from \textit{containers}.
-The most interesting point about the data structure is the polymorphic recursion that makes the data structure type safe as it guarantees equal nesting of \texttt{Node} types on both sides of the spine.
-It is also what makes it difficult to verify in other languages, because for example the Isabelle/HOL~\cite{isabelle} and the Lean~\cite{lean} theorem provers both do not directly support polymorphic recursion.
-In these languages, it is necessary to define the type such that it would allow uneven nesting of \texttt{Node}s and then "carve out" only the valid finger trees.
-Other provers like Coq~\cite{coq} support polymorphic recursion making a more direct one to one translation possible.
-
-\subsubsection{Cons and Snoc}
-
-As the finger tree is a symmetric tree, \texttt{cons} and \texttt{snoc} are very similar.
-For this reason we will show an explicit proof only for \texttt{cons} and provide the proof for \texttt{snoc} in the appendix.
-First we again copy definition of \texttt{cons} from the \textit{containers} package, adjusting it slightly to remove the size information that would require the type class.
-Also, to later use the function in a proof we add a \texttt{reflect} pragma.
-
-\begin{lstlisting}
+
 {-@ reflect cons @-}
@@
   Deep (Two a b) (cons (Node3 c d e) m) sf
-\end{lstlisting}
-
-The only complicated case is if we have already four digits at the front, as in all other cases we can simply increase the amount by one.
-In the case that we can not do so, we put the new data in front and push the "overflow" in to the recursion.
-The next step is to think of a valid potential function $\Phi$.
-Again, the intuition is which state of the data structure makes \texttt{cons} expensive.
-As seen above it is when the digit in the front is already full.
-So we can define the \textit{danger} of a digit as every state that will make the operation expensive.
-To also support and proof \texttt{uncons}, the inverse operation, we not only mark a full digit as dangerous, but also one that has only one element inside (as this would make \texttt{uncons} go into the recursion).
-Our potential is then the sum of the danger in the tree.
-
-\begin{lstlisting}
+
+{-@ reflect snoc @-}
+snoc :: FingerTree a -> a -> FingerTree a
+snoc EmptyT x = Single x
+snoc (Single a) x =
+  Deep (One a) EmptyT (One x)
+snoc (Deep pr m (One a)) x =
+  Deep pr m (Two a x)
+snoc (Deep pr m (Two a b)) x =
+  Deep pr m (Three a b x)
+snoc (Deep pr m (Three a b c)) x =
+  Deep pr m (Four a b c x)
+snoc (Deep pr m (Four a b c d)) x =
+  Deep pr (snoc m (Node3 a b c)) (Two d x)
+
 {-@ reflect dang @-}
@@
   dang pr + phi m + dang sf
-\end{lstlisting}
-
-The timing function for cons is similar to the timing functions in the stack just a copy of the original function with all non-recursive cases return $1$.
-
-\begin{lstlisting}
+
 {-@ reflect consT @-}
@@
   1 + consT (Node3 c d e) m
-\end{lstlisting}
-
-With all the preliminaries out of the way we can start proving the amortized complexity.
-Like with the stack we need to guess a constant that is an upper bound on our amortized run time.
-As we do not have a simple mental model like with the stack, we will delay this choice until we have proven the base cases.
-We will also use \textit{as patterns} again to avoid repetition.
-
-\begin{lstlisting}
+
+{-@ reflect snocT @-}
+{-@ snocT :: FingerTree a -> a -> Nat @-}
+snocT :: FingerTree a -> a -> Int
+snocT EmptyT _ = 1
+snocT Single{} _ = 1
+snocT (Deep _ _ One{}) _ = 1
+snocT (Deep _ _ Two{}) _ = 1
+snocT (Deep _ _ Three{}) _ = 1
+snocT (Deep _ m (Four a b c _)) _ =
+    1 + snocT m (Node3 a b c)
+
 {-@ consP :: x:a -> t:FingerTree a ->
-  { consT x t + phi (cons x t) - phi t <= c }
+  { consT x t + phi (cons x t) - phi t <= 3 }
 @-}
@@
 -- Use definition of phi
-  === 1 + 0 - 0
--- 1 must be <= c, so c >= 1
+  === 1 + 0 - 0 -- trivial, 1 <= 3
   *** QED
@@
 -- Use definition of phi and dang
-  === 1 + 1 + 0 + 1 - 0
--- 3 must be <= c, so c >= 3
-  *** QED
-\end{lstlisting}
-
-We omit the proof cases for one to three digits, as they work exactly as the other two base cases.
-The constant however will not change in the other base cases, so $3$ really is our upper bound to use here.
-The interesting case is when we have four digits.
-In this case we need to use induction to proof our goal.
-
-\begin{lstlisting}
+  === 1 + 1 + 0 + 1 - 0 -- trivial, 3 <= 3
+  *** QED
+consP x t@(Deep (One a) m sf) =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of cons/T
+  === 1 + phi (Deep (Two x a) m sf) - phi t
+-- Use definition of phi
+  === 1 + dang (Two x a) + phi m + dang sf
+    - (dang (One a) + phi m + dang sf)
+-- Remove parenthesis, use definition of dang
+  === 1 + 0 + phi m + dang sf
+    - 1 - phi m - dang sf
+-- everything cancels
+  === 0 -- trivial, 0 <= 3
+  *** QED
+consP x t@(Deep (Two a b) m sf) =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of cons/T
+  === 1 + phi (Deep (Three x a b) m sf) - phi t
+-- Use definition of phi
+  === 1 + dang (Three x a b) + phi m + dang sf
+    - (dang (Two a b) + phi m + dang sf)
+-- Remove parenthesis, use definition of dang
+  === 1 + 0 + phi m + dang sf
+    - 0 - phi m - dang sf
+-- phi and dang cancels
+  === 1 -- trivial, 1 <= 3
+  *** QED
+consP x t@(Deep (Three a b c) m sf) =
+    consT x t + phi (cons x t) - phi t
+-- Use definition of cons/T
+  === 1 + phi (Deep (Four x a b c) m sf) - phi t
+-- Use definition of phi
+  === 1 + dang (Four x a b c) + phi m + dang sf
+    - (dang (Three a b c) + phi m + dang sf)
+-- Remove parenthesis, use definition of dang
+  === 1 + 1 + phi m + dang sf
+    - 0 - phi m - dang sf
+-- phi and dang cancels
+  === 2 -- trivial, 2 <= 3
+  *** QED
 consP x t@(Deep (Four a b c d) m sf) =
@@
   *** QED
-\end{lstlisting}
-
-\subsubsection{Append}\label{sec:append}
-
-Similar to \texttt{push} in the stack, even though \texttt{append} has the same amortized complexity as its normal complexity, we need to show that it never increases the potential more than a logarithmic amount.
-Contrary to the simplified definition of Claessen~\cite{fingertrees_new}, the \textit{containers} package uses two sets of code generated, mutually recursive functions to implement \texttt{append}.
-One set of functions called \texttt{appendTreeX} handles the base cases where one of the two trees is either empty or consists of a singular element by using \texttt{cons} and \texttt{snoc}.
-In the recursive case it calls the \texttt{addDigitsX} functions that group arguments as \texttt{Node}s and itself call the \texttt{appendTreeX} functions.
-Because the \texttt{addDigitsX} functions do pattern match on their tree arguments at all, LiquidHaskell can not find a termination metric by itself.
-For this, we need to track the arguments across functions calls because \texttt{appendTreeX} has a strictly decreasing argument.
-We can do so by adding a ghost parameter to order the functions for LiquidHaskell.
-This order is lower for all \texttt{appendTreeX} functions than the \texttt{addDigitsX} functions.
-Additionally we need to set the case expansion to a high number such that all pattern matches are evaluation, else the termination proof fails.
-See the following snippet as an example with most cases and function omitted for brevity.
-
-\begin{lstlisting}
-{-@ LIQUID --max-case-expand=1000 @-}
-
-{-@ appendTree1 :: a:FingerTree (Node a) ->
-  Node a -> FingerTree (Node a) ->
-  FingerTree (Node a) / [size a, 1] @-}
-{-@ reflect appendTree1 @-}
-appendTree1 :: FingerTree (Node a) -> Node a ->
-    FingerTree (Node a) -> FingerTree (Node a)
-appendTree1 EmptyT !a xs = a `consTree` xs
-appendTree1 xs !a EmptyT = xs `snocTree` a
-appendTree1 (Single x) !a xs =
-    x `consTree` a `consTree` xs
-appendTree1 xs !a (Single x) =
-    xs `snocTree` a `snocTree` x
-appendTree1 (Deep pr1 m1 sf1) a (Deep pr2 m2 sf2) =
-    Deep pr1 (addDigits1 m1 sf1 a pr2 m2) sf2
-
-{-@ addDigits1 :: a:FingerTree (Node (Node a)) ->
-  Digit (Node a) -> Node a -> Digit (Node a) ->
-  FingerTree (Node (Node a)) ->
-  FingerTree (Node (Node a)) / [size a, 5] @-}
-{-@ reflect addDigits1 @-}
-addDigits1 :: FingerTree (Node (Node a)) ->
-  Digit (Node a) -> Node a -> Digit (Node a) ->
-  FingerTree (Node (Node a)) ->
-  FingerTree (Node (Node a))
-addDigits1 m1 (One a) b (One c) m2 =
-  appendTree1 m1 (Node3 a b c) m2
-addDigits1 m1 (One a) b (Two c d) m2 =
-  appendTree2 m1 (Node2 a b) (Node2 c d) m2
--- further cases omitted
-\end{lstlisting}
-
-This case expansion results in an explosion of constraints for the SMT solver to check.
-While the whole proof for \texttt{cons} including the defined functions generates less than 100 constraints, this termination proof generates over 3000.
-This leads to extremely high compile times.
-This is made worse by the need to define a timing function that needs to follow this mutually recursive structure and generates an additional 8000 constraints such that loading the file takes several minutes.
-Another problem is the logarithmic runtime itself.
-SMT provers are great with linear arithmetic, but are significantly slower with multiplication.
-Division is usually left as an unevaluated function.
-While one can transform the proof goal to turn logarithms into powers, it still requires assuming basic arithmetic rules on the logic level.
-
-For those reasons, using LiquidHaskell to proof the complexity with the physicist's method is not feasible (see section~\ref{sec:limitations} for a discussion of those limitations).
-Claessen~\cite{fingertrees_new} also does not formally proof this in his functional pearl, but provides a different reasoning to why \texttt{append} still is logarithmic.
-The potential only depends on the spine of the tree, so a new layer in the tree only adds at most a constant amount to the potential ($2$ if both digits are dangerous) making it grow linearly.
-Because the nodes in the digits always at least double (in the case of two \texttt{Node2}) the overall size of the tree grows exponentially.
-Thus the potential grows logarithmic compared to the size of the tree.
-
-\subsection{Definition of the timing function}
-
-One of the points where the analysis may lead to wrong results is if the timing function does not directly mirror the original function.
-So when the implementation changes and the timing function does not, it might be possible for the proofs to still pass.
-While this is unlikely, given how verbose proofs with LiquidHaskell are, it is not negligible.
-
-If one is in control of the source code and can change the implementation, it would be possible to factor out the recursion and share it between the implementation and the timing function.
-This however would clutter the implementation and most likely decrease performance as it is often the case with recursion schemes.
-
-A better solution would be to derive the timing function using meta programming in the form of TemplateHaskell.
-This meta function would take the name of the function in the implementation and create a new one with a configurable suffix.
-To do so it would check if the right hand sides contain a recursive call to the given name.
-If not, it replaces the right hand side with $1$, otherwise it changes the recursion to use the name with suffix and add $1$ at the end.
-The resulting function could then be spliced back into the source code.
-
-\begin{lstlisting}
-cons :: a -> FingerTree a -> FingerTree a
-cons a EmptyT = Single a
-cons a (Deep (Four b c d e) m sf) =
-    Deep (Two a b) (cons (Node3 c d e) m) sf
--- other cases omitted
-
-$(mkTimingFn "T" ''cons)
--- generates this; variable names omitted:
--- consT :: a -> FingerTree a -> Int
--- consT _ EmptyT = 1
--- consT _ (Deep (Four _ c d e) m _) =
---    consT (Node3 c d e) m + 1
--- other cases omitted
-\end{lstlisting}
-
-\subsection{LiquidHaskell for verification}
-
-As shown in earlier sections, LiquidHaskell has some limitations at the moment with regards to proving more complex functions.
-First and foremost the support for type classes or at least the support for instances involving Haskell \texttt{newtype}s is not great.
-Currently, LiquidHaskell only supports measure classes, which need to be defined by recursion over a single data type similar to a normal measure.
-For this reason it was not possible to use it for the \texttt{Sized} type class from \textit{containers}.
-There are several open issues about the type class support on the LiquidHaskell bug tracker, so we expect this to get better in the future.
-
-Another pain point was the combinatorial explosion when generating the constraints for the termination of \texttt{append} (see section~\ref{sec:append}).
-It might be possible to be smarter when generating the constraints here and thus creating less work for the SMT solver.
-LiquidHaskell also provides an option to disable the termination checker, but enabling this option surprisingly did not improve compile time.
-In fact, several other options that are supposed to improve performance lik \texttt{{-}{-}diff} which enables incremental type checking or \texttt{{-}{-}fast} which disables inference of polymorphic refinement, did either not affect performance at all or (in the case of \texttt{{-}{-}fast}) made it even worse.
-
-\section{Conclusion}
-
-Despite the current limitations of LiquidHaskell preventing more complicated functions from being proved formally, there is still a lot of value in this approach.
-Proving complexity of core data structures immediately pays back in a huge gain in confidence across the many applications that depend on them.
-While at the moment, amortized complexity of \O{1} seems to be the most feasable to proof, we expect the tooling around LiquidHaskell to improve a lot in the future, be it by incremental changes like better support for type classes or by more complicated means like a deeper embedding into a future dependently typed Haskell core.
-
-Aside from the industry application, we think this approach also has a big didactic advantage.
-To learn more about the complexity of data structures, a learner does not need to learn a new language on top of new concepts, but instead can use the language they already know to apply their new knowledge.
-The recent improvements in the usability of LiquidHaskell --- turning the previously standalone executable into a GHC plugin~\cite{lh_plugin} --- play a big role here as it drastically reduces the barrier of entry.
-
-%%
-%% The acknowledgments section is defined using the "acks" environment
-%% (and NOT an unnumbered section). This ensures the proper
-%% identification of the section in the article metadata, and the
-%% consistent spelling of the heading.
-%\begin{acks}
-%\end{acks}
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{references}
-
-%%
-%% If your work has an appendix, this is the place to put it.
-\appendix
-
-\section{Full proof code for the stack}
-
-\begin{lstlisting}
-import Language.Haskell.Liquid.ProofCombinators
-
-data Stack a = Empty | Elem a (Stack a)
-
-{-@ reflect push @-}
-push :: a -> Stack a -> Stack a
-push = Elem
-
-{-@ multipop :: Nat -> Stack a -> Stack a @-}
-multipop :: Int -> Stack a -> Stack a
-multipop _ Empty = Empty
-multipop 0 s = s
-multipop n (Elem _ s) = multipop (n - 1) s
-
-{-@ reflect phi @-}
-{-@ phi :: Stack a -> Nat @-}
-phi :: Stack a -> Int
-phi Empty = 0
-phi (Elem _ s) = 1 + phi s
-
-{-@ reflect pushT @-}
-{-@ pushT :: a -> Stack a -> Nat @-}
-pushT :: a -> Stack a -> Int
-pushT _ _ = 1
-
-{-@ reflect multipopT @-}
-{-@ multipopT :: Nat -> Stack a -> Nat @-}
-multipopT :: Int -> Stack a -> Int
-multipopT _ Empty = 1
-multipopT 0 _ = 1
-multipopT n (Elem _ s) = 1 + multipopT (n - 1) s
-
-{-@ pushP :: x:a -> s:Stack a ->
-  { pushT x s + phi (push x s) - phi s <= 2 }
-@-}
-pushP :: a -> Stack a -> Proof
-pushP x s = pushT x s + phi (push x s) - phi s
--- Use definition of push and pushT
-  === 1 + phi (Elem x s) - phi s
--- Use definition of phi (Elem _ _)
-  === 1 + (1 + phi s) - phi s
--- phi s gets canceled, 1 + 1 <= 2
-  *** QED
-
-{-@ multipopP :: k:Nat -> s:Stack a ->
-  { multipopT k s + phi (multipop k s)
-      - phi s <= 2 }
-@-}
-multipopP :: Int -> Stack a -> Proof
-multipopP 0 s = multipopT 0 s
-    + phi (multipop 0 s) - phi s
--- Use definition of multipop/T 0 _
-  === 1 + phi s - phi s
--- phi s cancels, 1 <= 2
-  *** QED
-multipopP k Empty = multipopT k Empty
-    + phi (multipop k Empty) - phi Empty
--- Use definition of multipop/T _ Empty
-  === 1 + phi Empty - phi Empty
--- phi Empty cancels, 1 <= 2
-  *** QED
-multipopP k xs@(Elem x s) = multipopT k xs
-    + phi (multipop k xs) - phi xs
--- Use definition of multipopT for recusive case
-  === (1 + multipopT (k - 1) s)
-    + phi (multipop k xs) - phi xs
--- Use definition of multipop for recusive case
-  === (1 + multipopT (k - 1) s)
-    + phi (multipop (k - 1) s) - phi xs
--- Use definition of phi (Elem _ _)
-  === (1 + multipopT (k - 1) s)
-    + phi (multipop (k - 1) s) - (1 + phi s)
--- Remove parenthesis
-  === 1 + multipopT (k - 1) s
-    + phi (multipop (k - 1) s) - 1 - phi s
--- 1s cancel out
-  === multipopT (k - 1) s
-    + phi (multipop (k - 1) s) - phi s
--- use induction
-  ? multipopP (k - 1) s
-  *** QED
-\end{lstlisting}
-
-\section{Full proof code of the finger tree}\label{ap:fingertree}
-
-\begin{lstlisting}
-data FingerTree a = Empty
-  | Single a
-  | Deep (Digit a) (FingerTree (Node a)) (Digit a)
-
-data Node a = Node2 a a | Node3 a a a
-
-data Digit a = One a
-  | Two a a
-  | Three a a a
-  | Four a a a a
-
-{-@ reflect cons @-}
-cons :: a -> FingerTree a -> FingerTree a
-cons a EmptyT = Single a
-cons a (Single b) =
-  Deep (One a) EmptyT (One b)
-cons a (Deep (One b) m sf) =
-  Deep (Two a b) m sf
-cons a (Deep (Two b c) m sf) =
-  Deep (Three a b c) m sf
-cons a (Deep (Three b c d) m sf) =
-  Deep (Four a b c d) m sf
-cons a (Deep (Four b c d e) m sf) =
-  Deep (Two a b) (cons (Node3 c d e) m) sf
-
-{-@ reflect snoc @-}
-snoc :: FingerTree a -> a -> FingerTree a
-snoc EmptyT x = Single x
-snoc (Single a) x =
-  Deep (One a) EmptyT (One x)
-snoc (Deep pr m (One a)) x =
-  Deep pr m (Two a x)
-snoc (Deep pr m (Two a b)) x =
-  Deep pr m (Three a b x)
-snoc (Deep pr m (Three a b c)) x =
-  Deep pr m (Four a b c x)
-snoc (Deep pr m (Four a b c d)) x =
-  Deep pr (snoc m (Node3 a b c)) (Two d x)
-
-{-@ reflect dang @-}
-{-@ dang :: Digit a -> Nat @-}
-dang :: Digit a -> Int
-dang One{} = 1
-dang Two{} = 0
-dang Three{} = 0
-dang Four{} = 1
-
-{-@ reflect phi @-}
-{-@ phi :: FingerTree a -> Nat @-}
-phi :: FingerTree a -> Int
-phi EmptyT = 0
-phi Single{} = 0
-phi (Deep pr m sf) =
-  dang pr + phi m + dang sf
-
-{-@ reflect consT @-}
-{-@ consT :: a -> FingerTree a -> Nat @-}
-consT :: a -> FingerTree a -> Int
-consT _ EmptyT = 1
-consT _ Single{} = 1
-consT _ (Deep One{} _ _) = 1
-consT _ (Deep Two{} _ _) = 1
-consT _ (Deep Three{} _ _) = 1
-consT _ (Deep (Four _ c d e) m _) =
-  1 + consT (Node3 c d e) m
-
-{-@ reflect snocT @-}
-{-@ snocT :: FingerTree a -> a -> Nat @-}
-snocT :: FingerTree a -> a -> Int
-snocT EmptyT _ = 1
-snocT Single{} _ = 1
-snocT (Deep _ _ One{}) _ = 1
-snocT (Deep _ _ Two{}) _ = 1
-snocT (Deep _ _ Three{}) _ = 1
-snocT (Deep _ m (Four a b c _)) _ =
-    1 + snocT m (Node3 a b c)
-
-{-@ consP :: x:a -> t:FingerTree a ->
-  { consT x t + phi (cons x t) - phi t <= 3 }
-@-}
-consP :: a -> FingerTree a -> Proof
-consP x t@EmptyT =
-    consT x t + phi (cons x t) - phi t
--- Use definition of cons/T _ EmptyT
-  === 1 + phi (Single x) - phi EmptyT
--- Use definition of phi
-  === 1 + 0 - 0 -- trivial, 1 <= 3
-  *** QED
-consP x t@(Single a) =
-    consT x t + phi (cons x t) - phi t
--- Use definition of cons/T _ (Single a)
-  === 1 + phi (Deep (One x) EmptyT (One a))
-        - phi (Single a)
--- Use definition of phi
-  === 1 + dang (One x) + phi EmptyT
-        + dang (One a) - 0
--- Use definition of phi and dang
-  === 1 + 1 + 0 + 1 - 0 -- trivial, 3 <= 3
-  *** QED
-consP x t@(Deep (One a) m sf) =
-    consT x t + phi (cons x t) - phi t
--- Use definition of cons/T
-  === 1 + phi (Deep (Two x a) m sf) - phi t
--- Use definition of phi
-  === 1 + dang (Two x a) + phi m + dang sf
-    - (dang (One a) + phi m + dang sf)
--- Remove parenthesis, use definition of dang
-  === 1 + 0 + phi m + dang sf
-    - 1 - phi m - dang sf
--- everything cancels
-  === 0 -- trivial, 0 <= 3
-  *** QED
-consP x t@(Deep (Two a b) m sf) =
-    consT x t + phi (cons x t) - phi t
--- Use definition of cons/T
-  === 1 + phi (Deep (Three x a b) m sf) - phi t
--- Use definition of phi
-  === 1 + dang (Three x a b) + phi m + dang sf
-    - (dang (Two a b) + phi m + dang sf)
--- Remove parenthesis, use definition of dang
-  === 1 + 0 + phi m + dang sf
-    - 0 - phi m - dang sf
--- phi and dang cancels
-  === 1 -- trivial, 1 <= 3
-  *** QED
-consP x t@(Deep (Three a b c) m sf) =
-    consT x t + phi (cons x t) - phi t
--- Use definition of cons/T
-  === 1 + phi (Deep (Four x a b c) m sf) - phi t
--- Use definition of phi
-  === 1 + dang (Four x a b c) + phi m + dang sf
-    - (dang (Three a b c) + phi m + dang sf)
--- Remove parenthesis, use definition of dang
-  === 1 + 1 + phi m + dang sf
-    - 0 - phi m - dang sf
--- phi and dang cancels
-  === 2 -- trivial, 2 <= 3
-  *** QED
-consP x t@(Deep (Four a b c d) m sf) =
-    consT x t + phi (cons x t) - phi t
--- Use definition of consT
-  === (1 + consT (Node3 b c d) m)
-  + phi (cons x t) - phi t
--- Use definition of cons
-  === (1 + consT (Node3 b c d) m)
-  + phi (Deep (Two x a) (cons (Node3 b c d) m) sf)
-  - phi t
--- Use definition of phi
-  === (1 + consT (Node3 b c d) m)
-  + dang (Two x a) + phi (cons (Node3 b c d) m)
-  + dang sf
-  - (dang (Four a b c d) + phi m + dang sf)
--- Remove parenthesis
-  === 1 + consT (Node3 b c d) m
-  + dang (Two x a) + phi (cons (Node3 b c d) m)
-  + dang sf
-  - dang (Four a b c d) - phi m - dang sf
--- dang sf cancels, use definition of dang
-  === 1 + consT (Node3 b c d) m
-  + 0 + phi (cons (Node3 b c d) m) - 1 - phi m
--- 1s cancel
-  === consT (Node3 b c d) m
-  + phi (cons (Node3 b c d) m) - phi m
--- Use induction
-  ? consP (Node3 b c d) m
-  *** QED
 

### 2021-05-22 paper/main.tex
Fix last errors

@@
 Fundamental data structures are one of the most important parts of a programming language ecosystem and a lot of production code depends on them.
-In the Haskell ecosystem, this job is done by the \textit{containers} package which has over 5300 reverse dependencies\footnote{reverse dependecies are other packages that depend on \textit{containers}} on Hackage~\cite{hackage}, the main repository for Haskell packages.
+In the Haskell ecosystem, this job is done by the \textit{containers} package which has over 5300 reverse dependencies\footnote{reverse dependencies are other packages that depend on \textit{containers}} on Hackage~\cite{hackage}, the main repository for Haskell packages.
 A lot of code depends on the assumptions that the data structures are fast and adhere to computational complexities stated in their documentation.
@@
 
-Usually these proofs are done in an external theorem prover (e.g.
-Coq~\cite{complexity_coq} or Isabelle/HOL~\cite{complexity_isabelle}), requiring an explicit translation step from the production language to the language the theorem prover uses.
+Usually these proofs are done in an external theorem prover (compare e.g.~\cite{complexity_coq} or~\cite{complexity_isabelle}), requiring an explicit translation step from the production language to the language the theorem prover uses.
 This translation step has a few disadvantages.
@@
 A different approach comes in the form of LiquidHaskell~\cite{liquidhaskell}.
-It is a refinement type~\cite{refinement_types} checker plugin\footnote{before 2020, LiquidHaskell was implemented as a standalone executable, it now being a GHC plugin~\cite{lh_plugin} drastically improves usability} for GHC, the main Haskell compiler used in industry.
-It allows specifing additional constraints on types, e.g.
+It is a refinement type~\cite{refinement_types} checker plugin\footnote{before 2020, LiquidHaskell was implemented as a standalone executable, it now being a GHC plugin~\cite{lh_plugin} drastically improves usability} for GHC, the main Haskell compiler used in the industry.
+It allows specifying additional constraints on types, e.g.
 "integers greater than zero" instead of just "integers".
@@
 This avoids any translation steps and makes it easy to keep implementation and its respective proof in sync, as any failure to do so will result in a compiler error.
-Another benefit is that the reader or author of a proof does not need to learn an additional language, the production language is the proof language.
+Another benefit is that the reader or author of a proof does not need to learn an additional language, the production language is the same as the proof language.
 
@@
 Expensive operations on the other hand can use the stored time to pay for their actual cost, hopefully putting them in a cheaper complexity class.
-In our example a \texttt{push} has to pay one actual time unit and it will also pay one extra time unit into the bank account.
-The difference between amortized cost and actual cost is constant, so \texttt{push} is still in \O{1}.
+In our example a \texttt{push} has to pay one time unit as actual cost and it will also pay one extra time unit into the bank account.
+The difference between amortized cost and actual cost is constant factor, so \texttt{push} is still in \O{1}.
 Because for every element pushed we deposit one extra time unit into the bank, the account always contains as many time units as there are elements on the stack.
 As \texttt{multipop} can not remove more elements than there are on the stack, withdrawing $k$ time units to pay for the actual cost will never make the account become negative.
-Because withdrawing from the bank account is enough to pay for the complete actual cost of \texttt{multipop} regardless of the choice of $k$, its \textit{amortized} runtime is in \O{1}.
+Because withdrawing from the bank account is enough to pay for the complete actual cost of \texttt{multipop} regardless of the choice of $k$, its \textit{amortized runtime} is in \O{1}.
 
@@
 We then define the amortized cost of an operation as $c + \Phi(h') - \Phi(h)$ where $c$ is the actual time cost of the operation, $h$ is the data structure before the operation and $h'$ is the data structure afterwards.
-Looking at a chain of operations $1, 2, ..., n$, the sum of the amortized time is an overestimation of the actual cost by $\Phi(h_n)$ and thus an upper bound on the actual cost (see equation~\ref{eq:sum}, remember that by definition $\Phi(h_0) = 0$).
+Looking at a chain of operations $1, 2, ..., n$, the sum of the amortized time is an overestimation of the actual cost by $\Phi(h_n)$ and thus an upper bound on the actual cost (see equation~\ref{eq:sum}, remember that by definition $\Phi(h_0) = 0$ and $\Phi(h_n) \ge 0$).
 
@@
 For a limited class of functions called \textit{measures}, this can be done while retaining full automatic reasoning.
-A measure is a function that takes only one argument which is an algebraic data type, has one equation per constructor of this data type and which only uses arithmetic functions and other measures in their right hand sides.
+A measure is a function that takes only one argument which is an algebraic data type, has one equation per constructor of this data type and which only uses arithmetic functions and other measures in its right hand sides.
 Other function still can be used in refinements when annotated with \texttt{reflect}, but the SMT solver is not able to check those refinement automatically.
@@
 While the amortized cost is in the same complexity class as the actual cost, we need to make sure that our potential function does not cause those two to diverge.
-LiquidHaskell does not allow quantifiers in refinement, because for functions other than measures, instanciating quantifiers in the SMT solver has unpredictable performance~\cite{lh_quantifiers}.
+LiquidHaskell does not allow quantifiers in refinement, because for functions other than measures, instantiating quantifiers in the SMT solver has unpredictable performance~\cite{lh_quantifiers}.
 This means we have to guess an upper limit for our amortized runtime.
@@
 
-To prove the complexity of \texttt{multipop}, we will need to distinquish between the two base cases and the recursive case.
+To prove the complexity of \texttt{multipop}, we will need to distinguish between the two base cases and the recursive case.
 With LiquidHaskell this is just a normal pattern match.
@@
 To show that this approach scales to more complicated data structures that are widely used in the industry, we will prove the complexity of finger trees, the data structure behind \texttt{Data.Seq} from \textit{containers}~\cite{containers}.
-It was originally described by Hinze and Paterson~\cite{fingertrees}, the implementation from \textit{containers} closely follows this description.
+It was originally described by Hinze and Paterson~\cite{fingertrees} and the implementation from \textit{containers} closely follows this description.
 The following definition is taken directly from \textit{containers}, but slightly modified to remove the size parameter.
@@
 It makes the data structure type safe as it guarantees equal nesting of \texttt{Node} types on both sides of the spine.
-It is also what makes it difficult to verify in other languages, because for example the both Isabelle/HOL~\cite{isabelle} and the Lean~\cite{lean} theorem provers do not directly support polymorphic recursion.
+It is also what makes it difficult to verify in other languages, because for example the both the Isabelle/HOL~\cite{isabelle} and the Lean~\cite{lean} theorem provers do not directly support polymorphic recursion.
 In these languages, it is necessary to define the type in such a way that it would allow uneven nesting of \texttt{Node}s and then "carve out" the subset of valid finger trees.
@@
 So we can define the \textit{danger} of a digit as every state that will make the operation expensive.
-To also support \texttt{uncons}, the inverse operation, we not only mark a full digits as dangerous, but also one that has only one element inside (as this would make \texttt{uncons} go into the recursion).
+To also support \texttt{uncons}, the inverse operation, we not only mark a full digits as dangerous, but also those that have only one element inside (as this would make \texttt{uncons} go into the recursion).
 We then define the potential to be the sum of the danger in the tree.
@@
 
-The timing function for \texttt{cons} is similar to the timing functions for the stack just a copy of the original function where all non-recursive cases return $1$.
+The timing function for \texttt{cons} is similar to the timing functions for the stack just a copy of the original function where all non-recursive cases return $1$:
 
@@
 Because we do not have a simple mental model like with the stack, we will delay this choice until we have proven the base cases.
-We will also use \textit{as patterns} again to avoid repetition.
+We will also use \textit{as patterns} again to avoid repetition:
 
@@
 Because the \texttt{addDigitsX} functions do not pattern match on their tree arguments at all, LiquidHaskell can not find a termination metric automatically.
-For this, we need to track the arguments across functions calls because \texttt{appendTreeX} has a strictly decreasing argument.
+For this, we need to track the arguments across function calls because only \texttt{appendTreeX} has a strictly decreasing argument.
 We can do so by adding a ghost parameter to order the functions for LiquidHaskell.

### 2021-07-05 seminar_paper/main.tex
Add seminar paper and presentation

\documentclass[acmlarge,screen,authorversion=true,nonacm=true]{acmart}
%\documentclass[sigplan,screen,review,anonymous]{acmart}

\usepackage{xcolor}
\definecolor{codegreen}{rgb}{0,0.6,0}

\usepackage{cancel}

\usepackage{listings}
\lstset{%
    language=Haskell,
    columns=fullflexible,
    basicstyle=\ttfamily,
    commentstyle=\color{codegreen},
    showstringspaces=false
}

% shorthand for big O
\renewcommand\O[1]{$\mathcal{O}(#1)$}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2021}
\acmYear{2021}
\acmDOI{tbd}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Haskell '21]{the 14th ACM SIGPLAN International Haskell Symposium}{August 26--27, 2021}{Virtual Event}
%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%  June 03--05, 2018, Woodstock, NY}
%\acmPrice{15.00}
%\acmISBN{978-1-4503-XXXX-X/18/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Theorem Proving for All: Equational Reasoning in Liquid Haskell (Seminar paper)}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Jan van Brügge}
\orcid{0000-0003-1560-7326}
\affiliation{%
  \institution{Technical University of Munich}
  \streetaddress{Arcisstraße 21}
  \city{Munich}
  \state{Bavaria}
  \country{Germany}
  \postcode{80333}
}
\email{jan.van-bruegge@tum.de}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Trovato and Tobin, et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
When programming in a pure functional language like Haskell, it is possible to reason about the execution of a function by viewing its definitions as equations that can be used to repeatedly rewrite a given expression until a final result is computed. Usually such reasoning is done "out of band" with pen and paper. LiquidHaskell is a refinement type checker for Haskell that allows the user to do verified equational reasoning directly in Haskell. This seminar paper presents how such reasoning can be used to prove theorems and derive efficient implementations from simple specification. It is based on the functional pearl by Vazou et al.~\cite{tpfa}.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10011007.10010940.10010992.10010998.10010999</concept_id>
       <concept_desc>Software and its engineering~Software verification</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Software verification}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{LiquidHaskell, theorem proving, seminar paper}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}\label{sec:introduction}

One of the most praised advantages of pure functional languages like Haskell is the possibility to use equational reasoning to make sense of the execution of a function. For a new learner it is helpful to "execute" a given function call manually to understand how the code behaves. For example when given the recursive definition of a very simple factorial function:

\begin{lstlisting}
fact :: Int -> Int
fact 0 = 1
fact n = n * fact (n - 1)
\end{lstlisting}

To understand how this function works, a learner can repeatedly replace the calls to the \texttt{fact} function with its definition and simplify the result:

\begin{lstlisting}
let x = fact 3           -- use definition
      = 3 * fact (3 - 1) -- simplify
      = 3 * fact 2       -- use definition
      = 3 * (2 * fact (2 - 1)) -- simplify
      = 3 * (2 * fact 1) -- use definition
      = 3 * (2 * (1 * fact (1 - 1))) -- simplify
      = 3 * (2 * (1 * fact 0)) -- use definition
      = 3 * (2 * (1 * 1)) -- simplify
      = 3 * (2 * 1) -- simplify
      = 3 * 2 -- simplify
      = 6
\end{lstlisting}

Usually such reasoning is done "out of band", for example with pen and paper. LiquidHaskell~\cite{liquidhaskell} enables the user to do such reasoning and proof properties about Haskell code directly in Haskell itself.

Usually proofs of software are done in external theorem provers like Coq~\cite{coq}, Lean~\cite{lean} or Isabelle/HOL~\cite{isabelle}. But to prove properties about Haskell code, the user first has to translate the code into the language of the respective theorem prover. Such a translation step is not needed with LiquidHaskell. In addition, the user does not need to learn a completely new language with subtly different semantics (for example lazy vs. strict evaluation) to do their proofs.

In this paper we:
\begin{itemize}
\item{list and explain the core features of LiquidHaskell (section~\ref{sec:features})}
\item{present two case studies, one to derive an efficient implementation and one to prove the amortized complexity of a stack (section~\ref{sec:case_studies})}
\item{briefly present the SMT solver that is used to check the refinements (section~\ref{sec:hood})}
\item{list the limitations of LiquidHaskell at the time of writing (section~\ref{sec:limitations})}
\item{compare LiquidHaskell to other popular theorem provers (section~\ref{sec:comparison})}
\item{summarize the current user experience in a conclusion (section~\ref{sec:conclusion})}
\end{itemize}

\section{Core features of LiquidHaskell}\label{sec:features}

\subsection{Refinement types}\label{sec:refinement_types}

The core addition of LiquidHaskell to Haskell's type system are \textit{refinement types}~\cite{refinement_types}. A refinement type allows the user to specify subsets of types to be used during type checking. For example if \texttt{Integer} is the type of all integers, then the refinement type \texttt{\{ x:Integer | x >= 0 \}} specifies the subset of non-negative integers. This syntax is taken directly from LiquidHaskell. Refinement types are written inside of braces (\texttt{\{\}}) where the type on the left of the pipe character (\texttt{|}) can name the type (\texttt{x} in our case) and refer to it on the right. When used on the arguments and return types of functions, refinement types work as pre- and postconditions. This means that LiquidHaskell will check the postcondition (the refinement on the return type) under the assumption of the pre-conditions (the refinements on the function arguments). For example a function that returns the tail of a list for non-empty lists:

\begin{lstlisting}
{-@ tail :: { xs:[a] | len xs > 0 } -> { ys:[a] | len ys == len xs - 1 } @-}
tail :: [a] -> [a]
tail (_:xs) = xs
\end{lstlisting}

As seen above, LiquidHaskell annotations are written in special comments (\texttt{\{-@ @-\}}) so that the code still compiles even without LiquidHaskell. While from the normal Haskell type signature below the annotation the function would be partial (it does not have an equation for the empty list), LiquidHaskell makes use of the precondition and thus can deduce that the input list will never be empty (see more about totality in section~\ref{sec:totality}).

The refinement types can also be extracted into type synonyms to enable reuse of common refinements:

\begin{lstlisting}
{-@ type Nat = { x:Integer | x >= 0 } @-}
\end{lstlisting}

Another property that the specification of subsets on types causes is subtyping. Normally, Haskell has no subtyping like object oriented languages where for example every valid \texttt{Dog} object is also a valid \texttt{Animal} object. With LiquidHaskell, subtyping is defined as the subset relation. For example the earlier type of non-negative integers is a subtype of all integers, because the set of all non-negative integers is a subset of all integers. In practice, this means that the user can pass a value that is a non-negative integer to a function that expects any integer without any explicit type coercions.

\subsection{Totality and termination}\label{sec:totality}

By default all functions in LiquidHaskell need to be total. This means that a function needs to handle all possible inputs and it is not allowed to use the \texttt{error} function on a live code path. While checking totality, LiquidHaskell takes the preconditions into account (see the code example in section~\ref{sec:refinement_types}).

In addition to handling all inputs, a function also has to terminate. For this, LiquidHaskell has a termination checker that infers a suitable termination metric by default. If this is not possible, the user can also specify a termination metric manually. This, however, is rarely needed, because the common case of recursive calls with strictly smaller arguments is usually enough:

\begin{lstlisting}
{-@ length :: [a] -> Int @-}
length :: [a] -> Int
length [] = 0 -- no recursion, terminates trivially
length (x:xs) = 1 + length xs -- argument to recursive call is smaller
\end{lstlisting}

The termination metric allows LiquidHaskell to generate an induction principle for the function (well-founded induction), which is used to prove the postconditions of recursive functions, for example that the sum of a list of non-negative numbers still is non-negative:

\begin{lstlisting}
{-@ sum :: [Nat] -> Nat @-}
sum [] = 0 -- trivial, 0 >= 0
sum (x:xs) =
    x -- from precondition: x >= 0
    + sum xs -- induction, assume postcondition already holds for `sum xs`
\end{lstlisting}

See the case studies in sections~\ref{sec:efficient} and~\ref{sec:complexity} for examples of using these induction principles to prove properties about the code.

\subsection{Lightweight and deep reasoning}\label{sec:reasoning}

By default, the SMT solver powering LiquidHaskell can only reason about linear arithmetic, like in the \texttt{sum} example in section~\ref{sec:totality} above. If the user wants to use functions in the refinements itself, they need to \textit{reflect} them first to the logic level. For a limited class of functions called \textit{measures} this is possible while retaining the complete automatic reasoning of the SMT solver. This is called \textit{lightweight reasoning}. For all other functions, the user needs to provide the proofs themselves (\textit{deep reasoning}). A measure is a function that:
\begin{itemize}
\item{takes exactly one argument, which must be an algebraic data type}
\item{is defined by a single equation per constructor of the data type}
\item{only calls primitive (e.g. arithmetic) functions of the SMT solver and other measures in its body}
\item{is marked with the \texttt{\{-@ measure @-\}} pragma}
\end{itemize}

As an example, the \texttt{length} function from earlier is a measure:

\begin{lstlisting}
{-@ measure length @-}
{-@ length :: [a] -> Nat @-}
length :: [a] -> Int
length [] = 0 -- one equation for [] constructor
length (x:xs) = 1 + length xs -- one equation for (:) constructor
\end{lstlisting}

Because we retain the automated reasoning for \texttt{length} we can use it in refinement without specifying a proof manually, for example to define an \texttt{append} function:

\begin{lstlisting}
{-@ (++) :: xs:[a] -> ys:[a] -> { zs:[a] | length zs == length xs + length ys } @-}
(++) :: [a] -> [a] -> [a]
[] ++ ys = ys
(x:xs) ++ ys = x:(xs ++ ys)
\end{lstlisting}

To use non-measures in refinements (for example the \texttt{append} function above), the user has to mark them with the \texttt{\{-@ reflect @-\}} pragma and provide the proofs of the refinements themselves (see sections~\ref{sec:proof_combinators} and~\ref{sec:case_studies}).

\subsection{Proof combinators}\label{sec:proof_combinators}

To make formal proofs of Haskell code possible, LiquidHaskell also provides a set of proof combinators in the \texttt{Language.Haskell.Liquid.ProofCombinators} module. The most important one is the equality operator (\texttt{===}):

\begin{lstlisting}
infixl 3 ===
{-@ (===) :: x:a -> { y:a | y == x } -> { z:a | z == x && z == y } @-}
(===) :: a -> a -> a
_ === y = y
\end{lstlisting}
The refinement on the operator asserts that both sides are provably equal and that it returns a value that is itself equal to both inputs. The implementation ignores the first argument and just returns the second, so at runtime only the last step of the reasoning is kept.

To make proofs look nicer, LiquidHaskell defines a \texttt{Proof} type alias and a \texttt{QED} data type together with a triple star operator (\texttt{***}):
\begin{lstlisting}
type Proof = ()

infixl 3 ***
{-@ assume (***) :: a -> p:QED -> { if (isAdmit p) then false else true } @-}
(***) :: a -> QED -> Proof
_ *** _ = ()

data QED = Admit | QED

{-@ measure isAdmit :: QED -> Bool @-}
{-@ Admit :: {v:QED | isAdmit v } @-}
\end{lstlisting}

\texttt{Proof} is just an alias for the normal Haskell unit type. The \texttt{QED} type has two constructors, \texttt{Admit} if you want to abort your proof and \texttt{QED} to run it. The refinement uses syntactic sugar for proofs:

\begin{lstlisting}
{-@ {- ... -} -> { property } @-} is the same as {-@ {- ... -} -> { x:() | property } @-}
\end{lstlisting}

These combinators allow us to make the example from the introduction (section~\ref{sec:introduction}) into valid Haskell code that is checked by LiquidHaskell:

\begin{lstlisting}
{-@ x :: { fact 3 == 6 } @-}
x :: Proof
x = fact 3           -- use definition
    === 3 * fact (3 - 1) -- simplify
    === 3 * fact 2       -- use definition
    {- steps left out for clarity -}
    === 3 * 2 -- simplify
    === 6
    *** QED
\end{lstlisting}

\section{Case studies}\label{sec:case_studies}

\subsection{Deriving efficient implementations from simple specifications}\label{sec:efficient}

Often, the simplest implementation of a function is not the most efficient version, but more efficient implementations are harder to reason about. Take for example a naive \texttt{reverse} function that uses the \texttt{append} function defined earlier in section~\ref{sec:reasoning}:

\begin{lstlisting}
{-@ reverse :: xs:[a] -> { ys:[a] | length xs == length ys } @-}
reverse [] = []
reverse (x:xs) = reverse xs ++ [x]
\end{lstlisting}

Because for every element in the list it calls \texttt{append}, which in turn has to traverse the whole first argument every time, this naive version has \O{n^2} time complexity where $n$ is the length of the list. Using the combinators from LiquidHaskell for equational reasoning, we can derive a more efficient version that uses an explicit accumulator and provably has the same behavior. To be able to use the \texttt{append} and reverse function in the refinement we need to add \texttt{reflect} annotations as neither function is a measure:

\begin{lstlisting}
{-@ reflect (++) @-}
{-@ reflect reverse @-}

{-@ revApp :: xs:[a] -> ys:[a] -> { zs:[a] | revApp xs ys == reverse xs ++ ys } @-}
revApp :: [a] -> [a] -> [a]
revApp xs ys = reverse xs ++ ys
\end{lstlisting}

The first step in the actual implementation is to pattern match on the first argument and replace the variable with the matched result:

\begin{lstlisting}
{-@ revApp :: xs:[a] -> ys:[a] -> { zs:[a] | revApp xs ys == reverse xs ++ ys } @-}
revApp :: [a] -> [a] -> [a]
revApp [] ys = reverse [] ++ ys
revApp (x:xs) ys = reverse (x:xs) ++ ys
\end{lstlisting}

With that, we can use LiquidHaskell's equational reasoning combinator (\texttt{===}) to simplify both equations by replacing the calls to \texttt{reverse} with its definition:

\begin{lstlisting}
{-@ revApp :: xs:[a] -> ys:[a] -> { zs:[a] | revApp xs ys == reverse xs ++ ys } @-}
revApp :: [a] -> [a] -> [a]
revApp [] ys = reverse [] ++ ys
    === [] ++ ys
    === ys
revApp (x:xs) ys = reverse (x:xs) ++ ys
    === (reverse xs ++ [x]) ++ ys
\end{lstlisting}

At this point, the first equation is already fully simplified, but we can not simplify the second equation further. The problem is that we have not proven that append is associative yet, so we can not move the parenthesis at the moment. We can prove this fact separately using the \texttt{?} combinator to apply the induction hypothesis:

\begin{lstlisting}
{-@ apAssocP :: xs:[a] -> ys:[a] -> zs:[a] -> { (xs ++ ys) ++ zs == xs ++ (ys ++ zs) } @-}
apAssocP :: [a] -> [a] -> [a] -> Proof
apAssocP [] ys zs = ([] ++ ys) ++ zs
    === ys ++ zs
    === [] ++ (ys ++ zs)
    *** QED
apAssocP (x:xs) ys zs = ((x:xs) ++ ys) ++ zs
    === x:(xs ++ ys) ++ zs
    === x:((xs ++ ys) ++ zs)
    ? apAssocP xs ys zs
    === x:(xs ++ (ys ++ zs))
    === x:xs ++ (ys ++ zs)
    *** QED
\end{lstlisting}

With this auxiliary lemma, we can continue the definition of the optimized reverse function:

\begin{lstlisting}
{-@ revApp :: xs:[a] -> ys:[a] -> { zs:[a] | revApp xs ys == reverse xs ++ ys } @-}
revApp :: [a] -> [a] -> [a]
revApp [] ys = reverse [] ++ ys
    === [] ++ ys
    === ys
revApp (x:xs) ys = reverse (x:xs) ++ ys
    === (reverse xs ++ [x]) ++ ys
    ? apAssocP (reverse xs) [x] ys
    === reverse xs ++ ([x] ++ ys)
    === reverse xs ++ (x:[] ++ ys)
    === reverse xs ++ (x:([] ++ ys))
    === reverse xs ++ (x:ys)
\end{lstlisting}

At this point, the definition still uses the old and inefficient \texttt{reverse} function, but the definition has the form \texttt{reverse as ++ bs} just like in our specification, so we can apply this equality:

\begin{lstlisting}
{-@ revApp :: xs:[a] -> ys:[a] -> { zs:[a] | revApp xs ys == reverse xs ++ ys } @-}
revApp :: [a] -> [a] -> [a]
revApp [] ys = reverse [] ++ ys
    === [] ++ ys
    === ys
revApp (x:xs) ys = reverse (x:xs) ++ ys
    === (reverse xs ++ [x]) ++ ys
    ? apAssocP (reverse xs) [x] ys
    === reverse xs ++ ([x] ++ ys)
    === reverse xs ++ (x:[] ++ ys)
    === reverse xs ++ (x:([] ++ ys))
    === reverse xs ++ (x:ys)
    === revApp xs (x:ys)
\end{lstlisting}

Now, our \texttt{revApp} does not use any other functions and only traverses the list once, making it \O{n}. Using this more generalized \texttt{reverse} function that also appends the second argument, we can define a third version that only takes a single argument and behaves like the first version:

\begin{lstlisting}
{-@ reverse' :: xs:[a] ->
    { ys:[a] | ys == reverse xs } @-}
reverse' :: [a] -> [a]
reverse' xs = reverse xs
    === undefined -- step missing
    === reverse xs ++ []
    === revApp xs []
\end{lstlisting}

Using the refinement of \texttt{revApp} we can derive the last step from the step before, but we are still unable to complete the definition. LiquidHaskell will not allow to just add an empty list to the right. We have to prove the fact that the first step and this step are the same with a right identity proof:

\begin{lstlisting}
{-@ rightIdP :: xs:[a] -> { xs ++ [] == xs } @-}
rightIdP :: [a] -> Proof
rightIdP [] = [] ++ []
    === []
    *** QED
rightIdP (x:xs) = (x:xs) ++ []
    === x:(xs ++ [])
    ? rightIdP xs
    === x:xs
    *** QED
\end{lstlisting}

With this lemma we can complete the definition of the new reverse function:

\begin{lstlisting}
{-@ reverse' :: xs:[a] -> { ys:[a] | ys == reverse xs } @-}
reverse' :: [a] -> [a]
reverse' xs = reverse xs
    ? rightIdP (reverse xs)
    === reverse xs ++ []
    === revApp xs []
\end{lstlisting}

At this point we could remove all the reasoning steps, simplifying the function to just the result. However given how the \texttt{===} operator is defined to ignore its first argument and return the second argument, this will be optimized away by the compiler, so we can leave in the explicit equality proof of the two functions.

\subsection{Proving amortized complexity of data structures}\label{sec:complexity}

The previous case study improved the complexity of \texttt{reverse} from \O{n^2} to \O{n}, but focused on showing that both implementations behave the same way. Our second case study thus focuses on proving complexity itself. In particular, We show the proof of the proof of the amortized complexity of a stack with a \textit{multipop} operation. In particular, we proof the amortized complexity in a \textit{linear} setting, this means that intermediate data structures are not reused.

Sometimes when analyzing the complexity of operations on data structures, the worst case is too pessimistic. One of the simplest examples is a stack that has two operations: \texttt{push} adds one cell on top of the stack and is in \O{1}, \texttt{multipop(k)} removes the top $k$ elements from the stack. In the worst case $k$ is the height of the stack, so \texttt{multipop} is in \O{n}. But in order to be able to remove an element from the stack, it needs to have been pushed at some point. We can use this to "spread out" the cost of \texttt{multipop} over multiple \texttt{push} operations. Such a stack can be defined in Haskell as follows. Normally \texttt{multipop} would return the removed elements along with the rest of the stack, but this just unnecessarily clutters the proofs later:

\begin{lstlisting}
data Stack a = Empty | Elem a (Stack a)

{-@ reflect push @-}
push :: a -> Stack a -> Stack a
push x s = Elem x s

{-@ reflect multipop @-}
{-@ multipop :: Nat -> Stack a -> Stack a @-}
multipop :: Int -> Stack a -> Stack a
multipop _ Empty = Empty
multipop 0 s = s
multipop n (Elem _ s) = multipop (n - 1) s
\end{lstlisting}

To formally prove the complexity, we will use the \textit{physicist's method}. It requires us to define a potential function $\Phi$ ("phi") that takes a state of the data structure and returns some cost. This costs represents how much time units need to be saved up for expensive operations for a given state of the data structure. It has to be zero for an empty data structure and greater or equal to zero for all other states. The \textit{amortized cost} is then defined as $c + \Phi(h') - \Phi(h)$ where $c$ is the actual time cost of the operation, $h$ is the data structure before the operation and $h'$ is the data structure afterwards. Looking at a chain of operations $1, 2, ..., n$, the sum of the amortized time is an overestimation of the actual cost by $\Phi(h_n)$ and thus an upper bound on the actual cost (see equation~\ref{eq:sum}, remember that by definition $\Phi(h_0) = 0$ and $\Phi(h_i) \ge 0$).

\begin{gather}
\nonumber (c_1 + \cancel{\Phi(h_1)} - \Phi(h_0)) + (c_2 + \cancel{\Phi(h_2)} - \cancel{\Phi(h_1)}) + ... + (c_n + \Phi(h_n) - \cancel{\Phi(h_{n - 1})}) = \\
  c_1 + c_2 + ... + c_n + \Phi(h_n) - \Phi(h_0) \label{eq:sum}
\end{gather}

To get the actual cost of an operation we additionally have to define a timing function \texttt{multipopT} that returns $1$ for non-recursive cases:

\begin{lstlisting}
{-@ measure phi @-}
{-@ phi :: Stack a -> Nat @-}
phi :: Stack a -> Int
phi Empty = 0
phi (Elem _ s) = 1 + phi s

{-@ reflect multipopT @-}
{-@ multipopT :: Nat -> Stack a -> Nat @-}
multipopT :: Int -> Stack a -> Int
multipopT _ Empty = 1
multipopT 0 _ = 1
multipopT n (Elem _ s) = 1 + multipopT (n - 1) s
\end{lstlisting}

For the actual proof we will start with the two base cases of \texttt{multipop}. Our proof goal is to keep the amortized cost below a fixed constant. LiquidHaskell does not have an existential quantifier (see section~\ref{sec:hood}), so we have to guess a constant beforehand. For this proof we will choose $2$:

\begin{lstlisting}
{-@ multipopP :: k:Nat -> s:Stack a ->
    { multipopT k s + phi (multipop k s) - phi s <= 2 } @-}
multipopP :: Int -> Stack a -> Proof
multipopP 0 s = multipopT 0 s
    + phi (multipop 0 s) - phi s
-- Use definition of multipop/T 0 _
  === 1 + phi s - phi s
-- phi s cancels, 1 <= 2
  *** QED
multipopP k Empty = multipopT k Empty
    + phi (multipop k Empty) - phi Empty
-- Use definition of multipop/T _ Empty
  === 1 + phi Empty - phi Empty
-- phi Empty cancels, 1 <= 2
  *** QED
\end{lstlisting}

To avoid cluttering the proof of the recursive case, we use an \textit{as pattern} to alias \texttt{Elem x s} to \texttt{xs}. The first steps are again applying the definitions of the used functions and simplifying the result. In the end we use the \texttt{?} operator to add the induction hypothesis to the proof step and thus complete the proof:

\begin{lstlisting}
multipopP k xs@(Elem x s) = multipopT k xs
    + phi (multipop k xs) - phi xs
-- Use definition of multipopT for recursive case
  === (1 + multipopT (k - 1) s)
    + phi (multipop k xs) - phi xs
-- Use definition of multipop for recursive case
  === (1 + multipopT (k - 1) s)
    + phi (multipop (k - 1) s) - phi xs
-- Use definition of phi (Elem _ _)
  === (1 + multipopT (k - 1) s)
    + phi (multipop (k - 1) s) - (1 + phi s)
-- Remove parentheses
  === 1 + multipopT (k - 1) s
    + phi (multipop (k - 1) s) - 1 - phi s
-- 1s cancel out
  === multipopT (k - 1) s
    + phi (multipop (k - 1) s) - phi s
-- use induction hypothesis
  ? multipopP (k - 1) s
  *** QED
\end{lstlisting}

Because we were successful in bounding the amortized complexity with a constant, our amortized runtime of \texttt{multipop} is in \O{1}.

This technique can also be used for much more sophisticated data structures. In "Liquid Amortization - Proving amortized complexity with LiquidHaskell"~\cite{amortized} we prove the amortized complexity of finger trees with this method.

\section{LiquidHaskell under the hood}\label{sec:hood}

To check all the refinement types, LiquidHaskell uses a SMT (Satisfiability Modulo Theories) solver, by default the z3 solver~\cite{z3}. Contrary to SAT (satisfiability) solvers, which can only solve boolean logic, SMT solvers can additionally reason about builtin theories, for example linear arithmetic. So, similar to how SAT solvers would try to find an assignment of true or false to the variables in a formula like this:

\begin{equation}
(x_1 \vee \neg x_2) \wedge (\neg x_1 \vee x_2 \vee x_3) \wedge \neg x_1
\end{equation}

An SMT solver finds valid assignments to formulas like:

\begin{equation}
\forall x\ y\ z.\ 3x + 2y - z > 4
\end{equation}

Using the SMT-LIB 2 standard~\cite{smt-lib} the formula above can be encoded for the z3 solver like this (note that functions that take unit as argument are regarded as constants):

\begin{lstlisting}
(declare-fun x () Int)
(declare-fun y () Int)
(declare-fun z () Int)

(assert (> (- (+ (* 3 x) (* 2 y)) z) 4))
(check-sat)
(get-model)
\end{lstlisting}

The solver then finds a valid assignment, in this case $x = 2$, $y = 0$, $z = 0$. To check the refinements, LiquidHaskell encodes them in the SMT-lib standard as assertions and hands them over to z3.

With a trivial refinement type like this:
\begin{lstlisting}
{-@ simple :: { x:Int | x > 2 } -> { x > 0 } @-}
simple :: Int -> Proof
simple _ = ()
\end{lstlisting}

LiquidHaskell generates the following SMT-LIB 2 code, simplified here. Note that LiquidHaskell negates the actual goal and checks if the solver returns \texttt{unsat}:

\begin{lstlisting}
(declare-fun x () Int)

(assert (> x 2))
(push 1)
(assert (not (> x 0)))
(check-sat)
; SMT Says: Unsat
(pop 1)
(exit)
\end{lstlisting}

As the SMT solver only has builtin theories about linear arithmetic, it can only reason about equality for other types like \texttt{String}.

For arbitrary formulas, the runtime of the solver is unbounded (see~\cite{lh_quantifiers}). This would be a very poor user experience for a type checker, where small changes in the types could cause the type checker to run forever. For this reason, LiquidHaskell only uses forall quantifiers and those only at the outermost level. There are no existential quantifiers and no higher rank types. For the proofs, this means that sometimes we need to guess constants instead of quantifying over them. On the other hand, all refinements are guaranteed to terminate eventually. While the runtime can still be very high, the restrictions force the user to split refinement more often, so usually the proofs terminate quickly.

\section{Limitations}\label{sec:limitations}

\subsection{Combinatorial explosion of constraints}

For functions with deep pattern matches, LiquidHaskell has to generate all other possible patterns for the constraints which leads to a combinatorial explosion of constraints. For simple functions, like the ones described in section~\ref{sec:case_studies}, LiquidHaskell generates less than 100 constraints for the SMT solver. But, for example, to prove the termination of the concatenation of two finger trees, it has to generate over 3000. This makes type checking take several minutes for a rather small source file (see~\cite{amortized}).

\subsection{Error messages}

Particularly bad are also the error messages at the moment. If there is a type mismatch and the refinement is rejected, LiquidHaskell shows the expected type and the actual type with all their refinements and mentions that the two are not in a subtype relation. Usually the types are so convoluted that the only helpful thing in the message is the line number where the error happened.

Even worse are errors regarding totality. If the user forgets a pattern match that is not excluded by the preconditions, LiquidHaskell displays a hard coded error message mentioning \texttt{Addr\#}. This apparently stems from the recent conversion of LiquidHaskell to a GHC plugin (see section~\ref{sec:conclusion}).

\subsection{Missing support for type classes}

At the time of writing there are some limitations to the usefulness of LiquidHaskell. First and foremost, the missing support for type classes. Most of the standard Haskell type classes have laws in addition to their methods. These laws are only stated in the documentation, but never checked by a compiler. For example, the \texttt{Semigroup} type class specifies that the combine operation (\texttt{<>}) has to be associative:

\begin{lstlisting}
(a <> b) <> c == a <> (b <> c)
\end{lstlisting}

As seen in section~\ref{sec:efficient}, such properties are often needed during proofs. One could imagine a type class \texttt{VSemigroup} (for \textit{verified} semigroup) that looks like this:

\begin{lstlisting}
class Semigroup a => VSemigroup a where
    {-@ assocP :: x:a -> y:a -> z:a ->
        { (x <> y) <> z == x <> (y <> z) } @-}
    assocP :: a -> a -> a -> Proof
\end{lstlisting}

This general form is however not allowed. The only form of type classes LiquidHaskell accepts are so called \textit{measure classes}, that share the limitations of normal measures (see section~\ref{sec:reasoning}) and thus do not allow for methods with more than one argument.

\section{Comparison with other proof assistants}\label{sec:comparison}

\subsection{Reasoning about Haskell code}

The strongest advantage of LiquidHaskell over external theorem provers like Coq~\cite{coq}, Lean~\cite{lean} or Isabelle/HOL~\cite{isabelle} is that the proofs are done directly in Haskell itself. While there are tools to automatically translate Haskell to Coq (\texttt{hs-to-coq}~\cite{hs-to-coq}) or Isabelle/HOL (\texttt{Haskabelle}~\cite{haskabelle}, only for Isabelle2009), they do not support the full range of Haskell's expressiveness. For example Haskell supports polymorphic recursion in data types which is not directly representable in the language of Isabelle/HOL (see~\cite{amortized} for a practical use of polymorphic recursion).

\subsection{Equational reasoning}

One advantage of the limited automation of LiquidHaskell is that it forces the user to explicitly provide their reasoning steps as part of the proof code. Usually, theorem provers like Coq or Lean use so called \textit{apply style} where the user repeatedly applies \textit{tactics} to prove the goal. Without inspecting the proof state after each tactic, it is not really possible to follow more complicated proofs as a reader. For example, even the rather simple associativity proof of list append in Coq looks like this:

\begin{lstlisting}
Theorem app_assoc : forall l1 l2 l3 : natlist,
    (l1 ++ l2) ++ l3 = l1 ++ (l2 ++ l3).
Proof.
    intros l1 l2 l3.
    induction l1 as [| n l1' IHl1'].
    - reflexivity.
    - simpl. rewrite -> IHl1'. reflexivity.
Qed.
\end{lstlisting}

Only Isabelle/HOL provides a similar means to do true equational reasoning in the form of Isar~\cite{isar}. In Isar, the append proof looks very similar to the LiquidHaskell version, including the explicit use of the induction hypothesis:

\begin{lstlisting}
lemma app_assoc:
    "(xs ++ ys) ++ zs = xs ++ (ys ++ zs)"
proof(induction xs)
    case Nil
    have "([] ++ ys) ++ zs = ys ++ zs" by simp
    also have "... = [] ++ (ys ++ zs)" by simp
    finally show ?case.
next
    case (Cons x xs)
    have "(x:xs ++ ys) ++ zs = x:(xs ++ ys) ++ zs" by simp
    also have "... = x:xs ++ (ys ++ zs)" by (simp add: Cons.IH)
    also have"... = (x:xs) ++ (ys ++ zs)" by simp
    finally show ?case.
qed
\end{lstlisting}

\section{Conclusion}\label{sec:conclusion}

Despite these limitations, LiquidHaskell is a very powerful tool. Even if the user does not care about proofs, the pre- and postconditions for functions drastically improve type safety. By default it disallows partial functions, so the number of runtime errors that can arise is substantially lower.

Additionally the limitations are not fundamental and can be fixed. In fact, the user experience of LiquidHaskell was greatly improved recently by converting it from a standalone executable to a GHC plugin~\cite{lh_plugin}. This makes it easy to install in a project by adding it to the project configuration and it also directly integrates with the Haskell language server that supplies IDE-like features to a wide range of text editors.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\appendix

\end{document}
\endinput

### 2024-07-01 paper/main.tex
Add seminar paper and presentation
Final paper version

@@
-%\documentclass[sigplan,screen]{acmart}
 \documentclass[sigplan,screen,review,anonymous]{acmart}
@@
 %% complete the rights form.
-\setcopyright{acmlicensed}
-\copyrightyear{2021}
-\acmYear{2021}
+\setcopyright{none}
+%\copyrightyear{2024}
+\acmYear{2024}
 \acmDOI{tbd}
@@
 %% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[Haskell '21]{the 14th ACM SIGPLAN International Haskell Symposium}{August 26--27, 2021}{Virtual Event}
+\acmConference[Haskell '24]{the 17th ACM SIGPLAN International Haskell Symposium}{September 02--05, 2024}{Milan, Italy}
 %\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
@@
 \affiliation{%
-  \institution{Technical University of Munich}
-  \streetaddress{Arcisstraße 21}
-  \city{Munich}
-  \state{Bavaria}
-  \country{Germany}
-  \postcode{80333}
+  \institution{Heriot-Watt University}
+  %\streetaddress{Arcisstraße 21}
+  \city{Edinburgh}
+  %\state{Bavaria}
+  \country{United Kingdom}
+  %\postcode{80333}
 }
-\email{jan.van-bruegge@tum.de}
+\email{jsv2000@hw.ac.uk}
 
@@
 \begin{abstract}
-Proofs of complexity of production code are usually done in external theorem provers.
-This makes it necessary to translate the data structures to be verified from the production language into the language of the prover.
+Formal reasoning about the time complexity of algorithms and data structures is usually done in interactive theorem provers like Isabelle/HOL~\cite{isabelle}.
+This includes reasoning about \textit{amortized} time complexity which looks at the worst case performance over a \textit{series} of operations.
+However, most programs are not written within a theorem prover and thus use the data structures of the production language.
+To verify the correctness it is necessary to translate the data structures from the production language into the language of the prover.
 Such a translation step could introduce errors, for example due to a mismatch in features between the two languages.
+
 We show how to prove amortized complexity of data structures directly in Haskell using LiquidHaskell~\cite{liquidhaskell}.
-Besides skipping the translation step, our approach also provides a didactic advantage as learners do not have to learn an additional language for their proofs and can focus on the new concepts only.
-We do not assume prior knowledge of amortized complexity as we explain the concepts and apply them in our first case study, a simple stack with multipop.
-The practicality of our approach in a production setting is demonstrated by our second case study: the finger tree.
-It powers \texttt{Data.Seq} from the \textit{containers} package~\cite{containers} which is widely depended on.
-Finally we discuss the current limitations of LiquidHaskell in this area and how it might be possible to improve the situation.
+Besides skipping the translation step, our approach can also provide a didactic advantage.
+Learners do not have to learn an additional language for proofs and can focus on the new concepts only.
+For this paper, we do not assume prior knowledge of amortized complexity as we explain the concepts and apply them in our first case study, a simple stack with multipop.
+Moving to more complicated (and useful) data structures, we show that the same technique works for binomial heaps which can be used to implement a priority queue.
+We also prove amortized complexity bounds for Claessen's version of the finger tree~\cite{fingertrees_new}, a sequence-like data structure with constant-time cons/uncons on either end.
+Finally we discuss the current limitations of LiquidHaskell that made certain versions of the data structures not feasible.
 \end{abstract}
@@
 
-Formally proving certain properties of production code is becoming more and more common.
-For safety critical code that is not allowed to "go wrong" as well as code that is fundamental and reused very often like core libraries, bugs can have a widespread impact.
+Formally proving properties of production code is becoming more and more common.
+For safety critical code that is not allowed to "go wrong" as well as code that is fundamental and reused very often, like core libraries, bugs can have a widespread impact.
 Proving that the code behaves according to the specification can provide an extra layer of confidence.
 Fundamental data structures are one of the most important parts of a programming language ecosystem and a lot of production code depends on them.
-In the Haskell ecosystem, this job is done by the \textit{containers} package which has over 5300 reverse dependencies\footnote{reverse dependencies are other packages that depend on \textit{containers}} on Hackage~\cite{hackage}, the main repository for Haskell packages.
-A lot of code depends on the assumptions that the data structures are fast and adhere to computational complexities stated in their documentation.
-Thus, proving those assumptions correct benefits a huge number of people.
-
-Usually these proofs are done in an external theorem prover (compare e.g.~\cite{complexity_coq} or~\cite{complexity_isabelle}), requiring an explicit translation step from the production language to the language the theorem prover uses.
+In languages like Haskell this comes mostly in the form of functional, persistent data structures like the ones described in Okasaki's book on purely functional data structures~\cite{okasaki}.
+
+Usually formal proofs are done in an external theorem prover (compare e.g.~\cite{complexity_coq} or~\cite{complexity_isabelle}), requiring an explicit translation step from the production language to the language the theorem prover uses.
 This translation step has a few disadvantages.
 First, the semantics of the target language could differ from the semantics of the production language.
-This is especially true for Haskell with its pervasive laziness, compared to most theorem provers that operate on a strict language.
+This is especially true for Haskell which is lazy, while most theorem provers have a strict language semantics.
 Second, such a translation step could introduce subtle differences or bugs, either by accident when doing a manual translation or through a bug when an automated translation tool is used.
@@
 
-A different approach comes in the form of LiquidHaskell~\cite{liquidhaskell}.
-It is a refinement type~\cite{refinement_types} checker plugin\footnote{before 2020, LiquidHaskell was implemented as a standalone executable, it now being a GHC plugin~\cite{lh_plugin} drastically improves usability} for GHC, the main Haskell compiler used in the industry.
+Another possibility is using LiquidHaskell~\cite{liquidhaskell}.
+It is a refinement type~\cite{refinement_types} checker plugin\footnote{Before 2020, LiquidHaskell was implemented as a standalone executable; it now being a GHC plugin~\cite{lh_plugin} drastically improves usability} for GHC, the main Haskell compiler used in industry.
 It allows specifying additional constraints on types, e.g.
@@
 \begin{itemize}
-\item{Assume no prior knowledge of amortized complexity as we recap the theoretical background (section~\ref{sec:complexity})}
-\item{Show the reasoning infrastructure that is needed to prove amortized complexity using LiquidHaskell (sections~\ref{sec:liquidhaskell} and~\ref{sec:stack})}
-\item{Apply this technique to \textit{finger trees}, the data structure behind \texttt{Data.Seq} from \textit{containers} (see our second case study in section~\ref{sec:fingertree})}
-\item{Discuss the usefulness of LiquidHaskell for such proofs compared with other theorem provers (sections~\ref{sec:limitations} and~\ref{sec:conclusion})}
+\item{Assume no prior knowledge of amortized complexity as we recap the theoretical background (section~\ref{sec:complexity}).}
+\item{Show the reasoning infrastructure that is needed to prove amortized complexity using LiquidHaskell (section~\ref{sec:liquidhaskell}).}
+\item{Apply this technique to several case studies: a simple \textit{stack} (section~\ref{sec:stack}), \textit{binomial heaps} (section~\ref{sec:binomal_heap}) and \textit{finger trees} (section~\ref{sec:fingertree}).}
+\item{Discuss the limitations of LiquidHaskell for such proofs compared with proper theorem provers (section~\ref{sec:conclusion}).}
 \end{itemize}
@@
 
-It is common practise to check how long an operation on a data structure takes in relation to the number of elements in the data structure.
-As the performance can vary wildly for the same operation depending on the state of data structure, usually the worst case for each operation is used as a metric and specified using \textit{Landau notation} ("big O").
-For example, a linear search takes $1$ unit of time if the searched element is at the front, while it takes $n$ units of time if it is the last element.
-Thus we would say that a linear search is in the complexity class of \O{n} because it will not take longer than $c \cdot n$ units of time for some fixed constant $c$.
+It is common practise to investigate how many units of time an operation on a data structure takes in relation to the number of elements in the data structure.
+As the performance can vary wildly for the same operation depending on the state of the data structure, usually the worst case for each operation is used as a metric and specified using \textit{Landau notation} ("big O").
+For example, a linear search\footnote{walking through a list element by element, checking if the current element is the one that is searched for} takes $1$ unit of time if the searched element is at the front, while it takes $n$ units of time if it is the last element.
+Thus we would say that linear search is in the complexity class of \O{n} because it will not take longer than $c \cdot n$ units of time for some fixed constant $c$.
 
@@
 In fact, to be able to remove an element, it has had to be pushed to the stack at some point, so we can spread out the cost of \texttt{multipop} over those pushes.
-There are multiple techniques to prove amortized complexity, we present the two most popular ones.
+
+There are multiple techniques to prove amortized complexity; we present the two most popular ones.
+Note that both of these techniques only apply in a non-shared setting where intermediate results are not reused.
+If we were allowed to keep a reference to the original data structure after an operation, we could do the same expensive operation several times without "paying" for it.
+Some data structures can exploit laziness to have the same (good) amortized complexity also in a shared setting, however formally verifying this would require explicit reasoning about the lazy semantics of the language and thus massively complicate the analysis.
 
@@
 
-The banker's method works by having a bank account that can store time units.
+The banker's method uses a "bank" account that can store time units.
 Operations have to pay their actual runtime cost and may additionally deposit into or withdraw from the account.
@@
 The idea is that cheap operations pay extra time units, making their amortized cost higher than their actual cost.
-The goal is to keep the difference between the two costs to a constant factor because that means the operation still retains their (cheap) complexity class.
-Expensive operations on the other hand can use the stored time to pay for their actual cost, hopefully putting them in a cheaper complexity class.
+The goal is to keep the sum of the actual cost and the extra deposit in the same (good) complexity class as just the actual cost.
+Expensive operations on the other hand can use the stored time to pay for their actual cost, hopefully putting them in a better complexity class.
 In our example a \texttt{push} has to pay one time unit as actual cost and it will also pay one extra time unit into the bank account.
-The difference between amortized cost and actual cost is constant factor, so \texttt{push} is still in \O{1}.
+The sum of the amortized cost and actual cost still constant, so \texttt{push} stays in \O{1}.
 Because for every element pushed we deposit one extra time unit into the bank, the account always contains as many time units as there are elements on the stack.
 As \texttt{multipop} can not remove more elements than there are on the stack, withdrawing $k$ time units to pay for the actual cost will never make the account become negative.
-Because withdrawing from the bank account is enough to pay for the complete actual cost of \texttt{multipop} regardless of the choice of $k$, its \textit{amortized runtime} is in \O{1}.
+Because withdrawing from the bank account is enough to pay for the complete actual cost of \texttt{multipop} regardless of the choice of $k$, its \textit{amortized complexity} is in \O{1}.
 
@@
 A simpler method is the physicist's method, which does not need such an account.
-The idea is to define a potential $\Phi$ ("phi") for the data structure.
+The idea is to define a potential $\Phi$ for the data structure.
 This potential is defined to be $0$ for an empty data structure and $\ge 0$ for all other states.
@@
 
-For our stack we define the potential to be the height of the stack.
+For our stack we can define the potential to be the height of the stack.
 This fulfils the requirements as the empty stack has a potential of zero and all other states have a potential greater or equal to zero.
@@
 
-As described by Vazou et.
-al.~\cite{tpfa}, one can use LiquidHaskell to do complex equational reasoning that is machine checked.
-It uses refinement types~\cite{refinement_types} that add additional constraints on top of normal Haskell data types which are checked by an external SMT solver.
+Usually LiquidHaskell is used to add more precise pre- and postconditions to functions with the help of refinement types~\cite{refinement_types}.
+The refinements are additional constraints on top of normal Haskell data types.
+These constraints are then checked by an external SMT solver.
 By default LiquidHaskell uses the z3~\cite{z3} solver, but it also supports cvc4~\cite{cvc4}.
-For example, we can declare the type of integers greater or equal to zero using a type synonym:
-
-\begin{lstlisting}
-{-@ type Nat = { x:Int | x >= 0 } @-}
-\end{lstlisting}
-
-LiquidHaskell annotations are multiline Haskell comments with "at" signs, such that the code still compiles even without LiquidHaskell.
+For example, it allows us to specify that the caller of a division function has to ensure that the denominator is not zero.
+
+\begin{lstlisting}
+{-@ div' :: Int -> { x:Int | x /= 0 } -> Int @-}
+div' :: Int -> Int -> Int
+div' = div
+\end{lstlisting}
+
+LiquidHaskell annotations are multiline Haskell comments delineated by @ signs, so that the code still compiles even without LiquidHaskell.
 It also allows the user to name argument types, so they can be referred to later.
-At the time of writing one has to copy the Haskell type signature into a LiquidHaskell comment and add the refinements there.
-A future version that is integrated deeper into the Haskell compiler might not need this code duplication.
-
-In LiquidHaskell, all functions need to terminate.
-Usually a suitable termination metric is automatically deduced, but in some more complex cases, we can also manually supply this metric (see section~\ref{sec:append} for an example).
-With the termination, LiquidHaskell is able to generate an induction principle for the function that can be used in a proof.
-Because LiquidHaskell is just Haskell, this induction amounts to a recursive call to the proof and is also used when defining ordinary Haskell functions with refinement on their types.
-For example the fact that the length of a list is never negative:
+Alternatively there exists a Quasiquoter that would allow us to get rid of the duplication between the Haskell and the LiquidHaskell type signatures.
+
+In LiquidHaskell, all functions have to terminate.
+Usually a suitable termination metric is automatically deduced, but in some more complex cases, we can also manually supply this metric.
+From this termination metric, LiquidHaskell is able to generate an induction principle for the function that can be used in a proof.
+For example to prove that the length of a list is never negative, LiquidHaskell uses the induction hypothesis to assume that the length of some list \texttt{xs} is not negative and uses that to prove \texttt{length (x:xs)} is not negative:
 
@@
 {-@ measure length @-}
-{-@ length :: [a] -> Nat @-}
+{-@ length :: [a] -> { x:Int | x >= 0 } @-}
 length :: [a] -> Int
@@
 A measure is a function that takes only one argument which is an algebraic data type, has one equation per constructor of this data type and which only uses arithmetic functions and other measures in its right hand sides.
-Other function still can be used in refinements when annotated with \texttt{reflect}, but the SMT solver is not able to check those refinement automatically.
-The user always has to provide the proofs themselves.
-
-To do so, LiquidHaskell provides a set of proof combinators and a \texttt{Proof} type.
-This type is just a type alias for the normal Haskell unit type.
-The reasoning comes from the \texttt{===} operator that asserts that both sides of the equality are truly equal.
-For a simple example we prove that the length of \texttt{append} applied to two lists is the sum of the lengths of the two lists.
+Other functions still can be used in refinements when annotated with \texttt{reflect}, but the SMT solver is not able to check those refinements automatically.
+The user always has to provide the proofs themselves using LiquidHaskell's ability to do complex equational reasoning~\cite{tpfa} that is checked by a machine.
+
+To facilitate these proofs, LiquidHaskell provides a set of proof combinators and a \texttt{Proof} type.
+This type is just a type alias for the normal Haskell unit type that is refined by the property that needs to be proved.
+The reasoning comes from the \texttt{===} operator that asserts that both sides of the equality have the same refinement.
+We can bring other facts in scope using the \texttt{?} operator.
+A proof can be finished by the \texttt{***} operator that casts a chain of equalities to the proof type (type slightly simplified here).
+
+\begin{lstlisting}
+type Proof = ()
+
+trivial :: Proof
+trivial = ()
+
+infixl 3 ===
+{-@ (===) :: x:a -> y:{a | y == x}
+  -> {v:a | v == x && v == y} @-}
+(===) :: a -> a -> a
+_ === y  = y
+
+infixl 3 ?
+{-@ (?) :: forall a b
+  <pa :: a -> Bool, pb :: b -> Bool>.
+  a<pa> -> b<pb> -> a<pa> @-}
+(?) :: a -> b -> a 
+x ? _ = x
+
+data QED = QED
+infixl 3 ***
+{-@ assume (***) :: a -> p:QED -> { true } @-}
+(***) :: a -> QED -> Proof
+_ *** _ = ()
+\end{lstlisting}
+
+To illustrate we prove that the length of \texttt{append} applied to two lists is the sum of the lengths of the two lists.
 Because \texttt{append} takes more than one argument, it is not a measure and thus we need to do the proof manually.
 
-First, we need to do a case distinction on the first argument.
+For this proof, we need to do a case distinction on the first argument.
 This is done by pattern matching on it like in a normal Haskell function.
@@
 
-The triple star operator (\texttt{***}) and the \texttt{QED} type at the end are just there to cast our proof to Haskell's unit type and improve readability.
-
-\section{Case studies}
-
-\subsection{Stack with Multipop}\label{sec:stack}
+This proof is quite lengthy, because we have to manually unfold definitions.
+LiquidHaskell also has a feature called \textit{proof by logical evaluation (ple)}.
+With this feature turned on, definitions are automatically unfolded several times for the SMT solver.
+Of course doing that for every definition might result in a performance hit, so it is also possible to selectively enable this feature on a per definition basis by tagging it with the \texttt{automatic-instances} flag.
+With \textit{ple} enabled, the proof boils down to doing the case distinction and specifying the induction step:
+
+\begin{lstlisting}
+{-@ automatic-instances lengthP @-}
+{-@ lengthP :: xs:[a] -> ys:[a] -> {
+  length (append xs ys) == length xs + length ys
+} @-}
+lengthP :: [a] -> [a] -> Proof
+lengthP [] _ = trivial
+lengthP (_ : xs) ys = trivial ? lengthP xs ys
+
+\end{lstlisting}
+
+\section{Case study: Stack with Multipop}\label{sec:stack}
 
 To continue with the example from section~\ref{sec:complexity}, we prove that $multipop$ has an amortized complexity of \O{1}.
-This data structure and its operations are defined like the following in Haskell.
-We already add some refinements to assert that $k$ is not negative.
-Note that normally, \texttt{multipop} would return the removed elements in addition to the new stack, but as this is irrelevant to the proof and only makes it more verbose we omit this in our definition.
-
-\begin{lstlisting}
+This data structure and its operations is defined below.
+We already add a precondition to assert that $k$ is not negative with the help of a LiquidHaskell type synonym.
+
+\begin{lstlisting}
+{-@ type Nat = { x:Int | x >= 0 } @-}
+
 data Stack a = Empty | Elem a (Stack a)
@@
 {-@ reflect multipop @-}
-{-@ multipop :: Nat -> Stack a -> Stack a @-}
-multipop :: Int -> Stack a -> Stack a
-multipop _ Empty = Empty
-multipop 0 s = s
-multipop n (Elem _ s) = multipop (n - 1) s
+{-@
+  multipop :: Nat -> Stack a -> ([a], Stack a)
+@-}
+multipop :: Int -> Stack a -> ([a], Stack a)
+multipop _ Empty = ([], Empty)
+multipop 0 s = ([], s)
+multipop n (Elem x s) =
+  let (xs, s') = multipop (n - 1) s
+  in (x:xs, s')
 \end{lstlisting}
@@
 As seen in section~\ref{sec:physicist}, we use the height of the stack as our potential.
-Thus we need to define a function that calculates the height of a stack.
-To define the timing functions, we copy the code for the operation, replace constant time steps with $1$ and adjust the recursion:
+
+The timing functions tell us the cost of every operation.
+Their definition follow the recursive structure of the operations.
+We could use TemplateHaskell to automatically derive these timing functions from the original operations but for clarity we will define them manually throughout this work.
 
@@
 {-@ reflect pushT @-}
-{-@ pushT :: a -> Stack a -> Nat @-}
+{-@ pushT :: a -> Stack a
+      -> { x:Int | x >= 1 } @-}
 pushT :: a -> Stack a -> Int
-pushT _ _ = 1
+pushT _ _ = 1 -- no recursion in
+               -- original definition
 
 {-@ reflect multipopT @-}
-{-@ multipopT :: Nat -> Stack a -> Nat @-}
+{-@ multipopT :: Nat -> Stack a
+     -> { x: Int | x >= 1 } @-}
 multipopT :: Int -> Stack a -> Int
@@
 
-First we will prove the complexity of \texttt{push}.
+First, we will prove the complexity of \texttt{push}.
 While the amortized cost is in the same complexity class as the actual cost, we need to make sure that our potential function does not cause those two to diverge.
-LiquidHaskell does not allow quantifiers in refinement, because for functions other than measures, instantiating quantifiers in the SMT solver has unpredictable performance~\cite{lh_quantifiers}.
+LiquidHaskell does not allow quantifiers in refinements, because for functions other than measures, instantiating quantifiers in the SMT solver has unpredictable performance~\cite{lh_quantifiers}.
 This means we have to guess an upper limit for our amortized runtime.
-As seen earlier in section~\ref{sec:banker}, we can pay two time units for a push to amortize the pops, so we will use $2$ as our guess:
+In section~\ref{sec:banker}, we paid two time units for a push to amortize the pops, so we will use $2$ as our guess:
 
@@
 With LiquidHaskell this is just a normal pattern match.
-Leaving the recursive case undefined for now, proving the base cases is very similar to \texttt{push}:
-
-\begin{lstlisting}
+Leaving the recursive case undefined for now, proving the base cases is very similar to \texttt{push} so we use \textit{ple} to automate the proofs:
+
+\begin{lstlisting}
+{-@ automatic-instances multipopP @-}
 {-@ multipopP :: k:Nat -> s:Stack a ->
-  { multipopT k s + phi (multipop k s)
+  { multipopT k s + phi (snd (multipop k s))
       - phi s <= 2 }
@@
 multipopP :: Int -> Stack a -> Proof
-multipopP 0 s = multipopT 0 s
-    + phi (multipop 0 s) - phi s
--- Use definition of multipop/T 0 _
-  === 1 + phi s - phi s
--- phi s cancels, 1 <= 2
-  *** QED
-multipopP k Empty = multipopT k Empty
-    + phi (multipop k Empty) - phi Empty
--- Use definition of multipop/T _ Empty
-  === 1 + phi Empty - phi Empty
--- phi Empty cancels, 1 <= 2
-  *** QED
+multipopP 0 _ = trivial
+multipopP _ Empty = trivial
 \end{lstlisting}
@@
 The first steps are again applying the definitions of the used functions and simplifying the result.
-In the end we use the \texttt{?} operator to add the induction hypothesis to the proof step and thus complete the proof:
+In the end we use the \texttt{?} operator to add the induction hypothesis to the proof step and thus complete the proof.
+We again use \textit{ple} here to unfold the definition of \texttt{snd} and \texttt{multipop} at once.
 
@@
   === (1 + multipopT (k - 1) s)
-    + phi (multipop k xs) - phi xs
+    + phi (snd (multipop k xs)) - phi xs
 -- Use definition of multipop for recursive case
   === (1 + multipopT (k - 1) s)
-    + phi (multipop (k - 1) s) - phi xs
+    + phi (snd (multipop (k - 1) s)) - phi xs
 -- Use definition of phi (Elem _ _)
   === (1 + multipopT (k - 1) s)
-    + phi (multipop (k - 1) s) - (1 + phi s)
+    + phi (snd (multipop (k - 1) s)) - (1 + phi s)
 -- Remove parentheses
   === 1 + multipopT (k - 1) s
-    + phi (multipop (k - 1) s) - 1 - phi s
+    + phi (snd (multipop (k - 1) s)) - 1 - phi s
 -- 1s cancel out
   === multipopT (k - 1) s
-    + phi (multipop (k - 1) s) - phi s
+    + phi (snd (multipop (k - 1) s)) - phi s
 -- use induction hypothesis
@@
 
-\subsection{Finger Tree}\label{sec:fingertree}
-
-To show that this approach scales to more complicated data structures that are widely used in the industry, we will prove the complexity of finger trees, the data structure behind \texttt{Data.Seq} from \textit{containers}~\cite{containers}.
-It was originally described by Hinze and Paterson~\cite{fingertrees} and the implementation from \textit{containers} closely follows this description.
-The following definition is taken directly from \textit{containers}, but slightly modified to remove the size parameter.
-LiquidHaskell caused problems in the proofs when using type classes because at the time of writing it is not possible to replace a call to a polymorphic function with the implementation in its instance.
-None of the function we use in the proofs use this size parameter, so we omit it.
-We will discuss such limitations more in depth in section~\ref{sec:limitations}.
-
-\begin{lstlisting}
-data FingerTree a = Empty
-  | Single a
-  | Deep (Digit a) (FingerTree (Node a)) (Digit a)
-
-data Node a = Node2 a a | Node3 a a a
-
-data Digit a = One a
-  | Two a a
-  | Three a a a
-  | Four a a a a
-\end{lstlisting}
-
-The finger tree has between one and four \textit{digits} on either side to enable cheap \texttt{cons} and \texttt{snoc}.
-As shown by Claessen~\cite{fingertrees_new}, this would also work with a maximum of just three digits, but we will follow the definition from \textit{containers}.
+While this stack data structure might look contrived, it is basically the same as Okasaki's queues~\cite{okasaki}.
+A queue is one front and one back list which is in reverse order.
+This allows us to remove elements from the front list and add elements to the back list in \O{1}.
+When the front list is empty, the back list is rotated and used as the new front list, an operation which is basically the same as our \texttt{multipop} here.
+
+\section{Case study: Binomial Heaps}\label{sec:binomal_heap}
+
+We now move on from ordinary queues to priority queues.
+In functional languages the latter are usually implemented as some kind of (min) heap.
+A binomial heap is a forest of binomial trees where there exists at most one tree of every rank.
+A binomial tree of rank zero is just the root.
+A binomial tree of rank $k$ consists of its root and one binomial tree for every rank from $k - 1$ to zero as its children.
+Two binomial trees of rank $k$ can be merged by attaching the tree with the bigger root as first child of the other tree resulting in a binomial tree of rank $k + 1$ (see figure~\ref{fig:heap_merge}).
+
+\begin{figure}
+\includegraphics[width=\columnwidth]{./Binomial_heap_merge1.pdf}
+\caption{Merge of two binomial trees of rank 2, adapted from Wikipedia~\cite{heap_merge}}\label{fig:heap_merge}
+\end{figure}
+
+To insert a new element into a binomial heap, we insert a new tree of rank zero into the forest.
+If the forest already contains a tree of rank zero the \texttt{merge} operation is used to get a tree of rank one.
+If there is already a tree of rank one in the forest, they get merged again, and so on.
+Once a free spot is found, insertion is finished.
+
+We can define the type of trees and of forest, as well as the operations like this in Haskell:
+
+\noindent
+\begin{minipage}{\linewidth}
+\begin{lstlisting}
+data Tree a = MkTree a [Tree a]
+
+{-@ reflect mergeTree @-}
+mergeTree :: Ord a =>
+  Tree a -> Tree a -> Tree a
+mergeTree l@(MkTree lr lc) r@(MkTree rr rc)
+  | lr <= rr = MkTree lr (r : lc)
+  | otherwise = MkTree rr (l : rc)
+
+data Heap a = FEnd -- empty forest
+  | F0 (Heap a) -- no tree at this position
+  | F1 (Tree a) (Heap a)
+
+{-@ reflect insertTree @-}
+insertTree :: Ord a =>
+  Tree a -> Heap a -> Heap a
+insertTree t FEnd = F1 t FEnd
+insertTree t (F0 f) = F1 t f
+insertTree t (F1 t' f) =
+  F0 (insertTree (mergeTree t t') f)
+
+\end{lstlisting}
+\end{minipage}
+
+Ideally we would use more advanced type level features of Haskell to ensure correctness of our implementation.
+For example we could ensure that the children of the root of a binomial tree have the correct ranks in descending order.
+However at the time of writing some language extensions necessary for this like \texttt{DataKinds} are not well supported by LiquidHaskell and cause problems in the proofs (see appendix~\ref{app:datakinds} for the definition of such a correct-by-construction binomial heap).
+
+For the amortized analysis of the binomial heap, \texttt{merge} only does one comparison and one allocation no matter which rank the trees are of so it is in \O{1}.
+For the insertion we only work on the roots of which there are at most $log_2 \: n$ many.\footnote{A tree of rank $k$ consists of two trees of rank $k - 1$, so the number of elements in the tree always doubles} So the operation is in \O{log \: n}.
+
+As our potential $\Phi$ we will use the number of trees in the forest.
+The intuition here is that the worst case happens when there is a tree at every position and we need to walk through the whole forest to find a free spot.
+This also fulfils the requirement that the empty data structure has a potential of zero.
+Again for clarity we will define the timing functions manually to show how they are derived from the original operations.
+As the merge operation is constant we include its cost directly in the insert operation for brevity.
+The proof is very similar to the \texttt{multipop} example from section~\ref{sec:stack}, so we will use \textit{ple} to automate most of it.
+
+\begin{lstlisting}
+{-@ reflect phi @-}
+{-@ phi :: Heap a -> Nat @-}
+phi :: Heap a -> Int
+phi FEnd = 0
+phi (F0 rest) = phi rest
+phi (F1 rest) = 1 + phi rest
+
+{-@ reflect insertT @-}
+{-@ insertT :: Ord a =>
+  Tree a -> Heap a -> { x:Int | x >= 1 } @-}
+insertT :: Ord a =>
+  Tree a -> Heap a -> Int
+insertT _ FEnd = 1
+insertT _ (F0 _) = 1
+insertT t (F1 t' f) = 1 +
+  insertT (mergeTree t t') f
+
+{-@ automatic-instances insertTreeP @-}
+{-@ insertTreeP :: t:_ -> f:_ ->
+  { insertT t f + pot (insertTree t f)
+    - pot f <= 2 }
+@-}
+insertTreeP :: Ord a => Tree a -> Heap a -> Proof
+insertTreeP t (F1 t' f') = trivial
+  ? insertTreeP (mergeTree t t') f'
+insertTreeP _ _ = trivial
+\end{lstlisting}
+
+\section{Case study: Finger Trees}\label{sec:fingertree}
+
+To show that our approach scales to more complicated data structures that are widely used in industry, we will prove the complexity of finger trees, the data structure behind \texttt{Data.Seq} from \textit{containers}~\cite{containers}.
+It was originally described by Hinze and Paterson~\cite{fingertrees}, but we will follow the simplified version of Claessen~\cite{fingertrees_new} here.
+The simplified version does not need nested pattern matches to implement concatenation of finger trees, which would otherwise lead to a combinatorial explosion in LiquidHaskell constraints.
+
+\begin{lstlisting}
+data Seq a = Nil
+  | Unit a
+  | More (Digit a) (Seq (Tuple a)) (Digit a)
+
+data Digit a = One a | Two a a | Three a a a
+
+data Tuple a = Pair a a | Triple a a a
+\end{lstlisting}
+
+The finger tree has between one and three elements on either side to enable cheap \texttt{cons} and \texttt{snoc}.
 The most interesting property of the data structure is the polymorphic recursion.
-It makes the data structure type safe as it guarantees equal nesting of \texttt{Node} types on both sides of the spine.
-It is also what makes it difficult to verify in other languages, because for example the both the Isabelle/HOL~\cite{isabelle} and the Lean~\cite{lean} theorem provers do not directly support polymorphic recursion.
-In these languages, it is necessary to define the type in such a way that it would allow uneven nesting of \texttt{Node}s and then "carve out" the subset of valid finger trees.
+It ensures an equal depth of nesting of \texttt{Tuple} types on both sides of the spine.
+It is also what might make it difficult to verify in other languages, as for example Isabelle/HOL~\cite{isabelle} does not directly support polymorphic recursion.
+In this case, it would be necessary to define the type such that it allows uneven nesting of \texttt{Tuple}s and then "carve out" the subset of valid finger trees.
 Other provers like Coq~\cite{coq} support polymorphic recursion, allowing a more direct one to one translation.
 
-\subsubsection{Cons and Snoc}
+\subsection{Cons and Snoc}
 
 Because the finger tree is a symmetric tree, \texttt{cons} and \texttt{snoc} are very similar.
-For this reason we show an explicit proof only for \texttt{cons} and provide the proof for \texttt{snoc} in the appendix.
-First we again copy the definition of \texttt{cons} from the \textit{containers} package, adjusting it slightly to remove the size information that would require the type class.
-In order to later use the function in a proof we add a \texttt{reflect} pragma.
+For this reason we show an explicit definition and proof only for \texttt{cons} and provide the definition and proof for \texttt{snoc} in the appendix.
+The base cases are straightforward and just use the flexibility of the digits to add the element to the front.
+The interesting case is when the front is already full and we need to recurse.
 
@@
 {-@ reflect cons @-}
-cons :: a -> FingerTree a -> FingerTree a
-cons a EmptyT = Single a
-cons a (Single b) =
-  Deep (One a) EmptyT (One b)
-cons a (Deep (One b) m sf) =
-  Deep (Two a b) m sf
-cons a (Deep (Two b c) m sf) =
-  Deep (Three a b c) m sf
-cons a (Deep (Three b c d) m sf) =
-  Deep (Four a b c d) m sf
-cons a (Deep (Four b c d e) m sf) =
-  Deep (Two a b) (cons (Node3 c d e) m) sf
-\end{lstlisting}
-
-The only complex case is if we already have four digits at the front, as in all other cases we can simply grow the digit in size.
-In the complex case, we put the new data in front and push the "overflow" into the recursion.
+cons :: a -> Seq a -> Seq a
+cons x Nil = Unit x
+cons x (Unit y) = More (One x) Nil (One y)
+cons x (More (One y) q u) = More (Two x y) q u
+cons x (More (Two y z) q u) =
+  More (Three x y z) q u
+cons x (More (Three y z w) q u) =
+  More (Two x y) (cons (Pair z w) q) u
+\end{lstlisting}
+
 The next step is to find a valid potential function $\Phi$.
@@
 So we can define the \textit{danger} of a digit as every state that will make the operation expensive.
-To also support \texttt{uncons}, the inverse operation, we not only mark a full digits as dangerous, but also those that have only one element inside (as this would make \texttt{uncons} go into the recursion).
+To also support \texttt{uncons} -- the inverse operation -- we not only mark a full digit as dangerous, but also those that have only one element inside (as this would make \texttt{uncons} go into the recursion).
 We then define the potential to be the sum of the danger in the tree.
-
-\begin{lstlisting}
-{-@ reflect dang @-}
-{-@ dang :: Digit a -> Nat @-}
-dang :: Digit a -> Int
-dang One{} = 1
-dang Two{} = 0
-dang Three{} = 0
-dang Four{} = 1
+The timing functions are again directly derived from the \texttt{cons} operation itself.
+The proof itself can be mostly automated with \textit{ple}.
+
+\begin{lstlisting}
+{-@ reflect danger @-}
+{-@ danger :: Digit a -> Nat @-}
+danger :: Digit a -> Int
+danger One{} = 1
+danger Two{} = 0
+danger Three{} = 1
 
 {-@ reflect phi @-}
-{-@ phi :: FingerTree a -> Nat @-}
-phi :: FingerTree a -> Int
-phi EmptyT = 0
-phi Single{} = 0
-phi (Deep pr m sf) =
-  dang pr + phi m + dang sf
-\end{lstlisting}
-
-The timing function for \texttt{cons} is similar to the timing functions for the stack just a copy of the original function where all non-recursive cases return $1$:
-
-\begin{lstlisting}
+{-@ phi :: Seq a -> Nat @-}
+phi :: Seq a -> Int
+phi Nil = 0
+phi Unit{} = 0
+phi (More pr m sf) = danger pr + phi m + danger sf
+
 {-@ reflect consT @-}
-{-@ consT :: a -> FingerTree a -> Nat @-}
-consT :: a -> FingerTree a -> Int
-consT _ EmptyT = 1
-consT _ Single{} = 1
-consT _ (Deep One{} _ _) = 1
-consT _ (Deep Two{} _ _) = 1
-consT _ (Deep Three{} _ _) = 1
-consT _ (Deep (Four _ c d e) m _) =
-  1 + consT (Node3 c d e) m
-\end{lstlisting}
-
-With all the required preliminaries defined, we start proving the amortized complexity.
-Just like in the stack example (see section~\ref{sec:stack}) we need to guess a constant that is an upper bound of our amortized run time.
-Because we do not have a simple mental model like with the stack, we will delay this choice until we have proven the base cases.
-We will also use \textit{as patterns} again to avoid repetition:
-
-\begin{lstlisting}
-{-@ consP :: x:a -> t:FingerTree a ->
-  { consT x t + phi (cons x t) - phi t <= c }
+{-@ consT :: a -> Seq a -> { x:Int | x >= 1 } @-}
+consT :: a -> Seq a -> Int
+consT _ Nil = 1
+consT _ (Unit _) = 1
+consT _ (More (One _) _ _) = 1
+consT _ (More (Two _ _) _ _) = 1
+consT _ (More (Three _ z w) q _) =
+  1 + consT (Pair z w) q
+
+{-@ automatic-instances consP @-}
+{-@ consP :: x:a -> t:Seq a ->
+    { consT x t + phi (cons x t) - phi t <= 3 }
 @-}
-consP :: a -> FingerTree a -> Proof
-consP x t@EmptyT =
-    consT x t + phi (cons x t) - phi t
--- Use definition of cons/T _ EmptyT
-  === 1 + phi (Single x) - phi EmptyT
--- Use definition of phi
-  === 1 + 0 - 0
--- 1 must be <= c, so c >= 1
+consP :: a -> Seq a -> Proof
+consP _ (More (Three _ z w) q _) = trivial
+  ? consP (Pair z w) q
+consP _ _ = trivial
+\end{lstlisting}
+
+\subsection{Append}\label{sec:append}
+
+Even though \texttt{append} has the same amortized complexity as its normal complexity, we need to show that it never increases the potential more than a logarithmic amount.
+Claessen~\cite{fingertrees_new} uses a helper function \texttt{glue} that concatenates two finger trees with at most three extra elements in the middle.
+Several other functions also require very specific lower and upper bounds on the length of their arguments.
+This can be directly expressed using LiquidHaskell.
+
+\begin{lstlisting}
+{-@ reflect glue @-}
+{-@ glue :: Seq a -> { xs:[_] | len xs <= 3 }
+    -> Seq a -> Seq a @-}
+glue :: Seq a -> [a] -> Seq a -> Seq a
+glue Nil as q2 = foldr cons q2 as
+glue q1 as Nil = foldl snoc q1 as
+glue (Unit x) as q2 = foldr cons q2 (x : as)
+glue q1 as (Unit x) = snoc (foldl snoc q1 as) x
+glue (More u1 q1 v1) as (More u2 q2 v2) =
+  More u1 (glue q1 (
+    toTuples (toList v1 ++ as ++ toList u2)
+  ) q2) v2
+
+{-@ reflect append @-}
+append :: Seq a -> Seq a -> Seq a
+append q1 q2 = glue q1 [] q2
+
+{-@ reflect toList @-}
+{-@ toList :: Some a ->
+  { xs:[a] | len xs >= 1 && len xs <= 3 } @-}
+toList :: Some a -> [a]
+toList (One x) = [x]
+toList (Two x y) = [x, y]
+toList (Three x y z) = [x, y, z]
+
+{-@ reflect toTuples @-}
+{-@ toTuples ::
+  { xs:[_] | len xs >= 2 && len xs <= 9 }
+  -> { ys:[_] | len ys >= 1 && len ys <= 3 } @-}
+toTuples :: [a] -> [Tuple a]
+toTuples = toTuples'
+
+{-@ reflect toTuples' @-}
+{-@ toTuples' ::
+  { xs:[_] | len xs != 1 && len xs <= 9 }
+  -> { ys:[_] | if len xs == 0 then
+        len ys == 0
+      else if len xs <= 3 then
+        len ys == 1
+      else if len xs <= 6 then
+        len ys == 2
+      else len ys == 3 } @-}
+toTuples' :: [a] -> [Tuple a]
+toTuples' [] = []
+toTuples' [x, y] = [Pair x y]
+toTuples' [x, y, z, w] = [Pair x y, Pair z w]
+toTuples' (x:y:z:xs) =
+  Triple x y z : toTuples' xs
+\end{lstlisting}
+
+We also need to reimplement \texttt{foldl}, \texttt{foldr} and list concatenation because their definitions in the standard library do not come with \texttt{reflect} annotations so they cannot be used in a LiquidHaskell proof.
+
+\begin{lstlisting}
+{-@ reflect foldl @-}
+foldl :: (b -> a -> b) -> b -> [a] -> b 
+foldl _ x [] = x
+foldl f a (x:xs) = foldl f (f a x) xs
+
+{-@ reflect foldr @-}
+foldr :: (a -> b -> b) -> b -> [a] -> b
+foldr _ x [] = x
+foldr f a (x:xs) = f x (foldr f a xs)
+
+{-@ reflect ++ @-}
+{-@ (++) :: x:[_] -> y:[_] ->
+  { z:[_] | len z == len x + len y } @-}
+(++) :: [a] -> [a] -> [a]
+[] ++ ys = ys
+(x:xs) ++ ys = x : (xs ++ ys)
+\end{lstlisting}
+
+For the amortization proof we of course have to use the same potential as for \texttt{cons}/\texttt{snoc}, and the timing functions are directly derived from the \texttt{glue} operation itself and all other functions called by it.
+Note that the timing functions of the higher-order \texttt{foldl} and \texttt{foldr} functions take a timing function as input.
+
+\begin{lstlisting}
+{-@ reflect foldrT @-}
+{-@ foldrT :: (a -> b -> b) ->
+  (a -> b -> { x:Int | x >= 1 }) -> b ->
+  [a] -> { x:Int | x >= 1 } @-}
+foldrT :: (a -> b -> b) -> (a -> b -> Int)
+  -> b -> [a] -> Int
+foldrT _ _ _ [] = 1
+foldrT f fT b (x:xs) = fT x (foldr f b xs)
+  + foldrT f fT b xs
+
+{-@ reflect foldlT @-}
+{-@ foldlT :: (b -> a -> b) ->
+  (b -> a -> { x:Int | x >= 1 }) -> b ->
+  [a] -> { x:Int | x >= 1 } @-}
+foldlT :: (b -> a -> b) -> (b -> a -> Int)
+  -> b -> [a] -> Int
+foldlT _ _ _ [] = 1
+foldlT f fT a (x:xs) = fT a x
+  + foldlT f fT (f a x) xs
+
+{-@ reflect glueT @-}
+{-@ glueT :: Seq a -> { as:[a] | len as <= 3 }
+  -> Seq a -> { x:Int | x >= 1 } @-}
+glueT :: Seq a -> [a] -> Seq a -> Int
+glueT Nil as q2 = 1 + foldrT cons consT q2 as
+glueT q1 as Nil = 1 + foldlT snoc snocT q1 as
+glueT (Unit x) as q2 =
+  1 + foldrT cons consT q2 (x : as)
+glueT q1 as (Unit x) =
+  1 + snocT (foldl snoc q1 as) x
+    + foldlT snoc snocT q1 as
+glueT (More _ q1 v1) as (More u2 q2 _) =
+  1 + glueT q1 (
+      toTuples (toList v1 ++ as ++ toList u2)
+    ) q2
+\end{lstlisting}
+
+We also need a way to specify the logarithmic complexity, so we will define a logarithm function as well as a way to calculate the number of elements in a finger tree.
+
+\begin{lstlisting}
+{-@ reflect log2 @-}
+{-@ log2 :: { x:Int | x >= 1 } -> Nat @-}
+log2 :: Int -> Int
+log2 1 = 0
+log2 n = 1 + log2 (n `div` 2)
+
+{-@ reflect tuplesToList @-}
+{-@ tuplesToList :: x:[_] -> { y:[_] |
+    len y <= 3 * len x && len y >= 2 * len x
+  } @-}
+tuplesToList :: [Tuple a] -> [a]
+tuplesToList [] = []
+tuplesToList (Pair a b:xs) = a:b:tuplesToList xs
+tuplesToList (Triple a b c:xs) =
+  a:b:c:tuplesToList xs
+
+{-@ reflect seqToList @-}
+seqToList :: Seq a -> [a]
+seqToList Nil = []
+seqToList (Unit x) = [x]
+seqToList (More u q v) = toList u ++
+  tuplesToList (seqToList q) ++ toList v
+\end{lstlisting}
+
+For the base cases of \texttt{glue} it is helpful to prove that folding \texttt{cons}/\texttt{snoc} over a list takes as many time steps as the amortized complexity of \texttt{cons}/\texttt{snoc} times the number of elements in the list.
+
+\begin{lstlisting}
+{-@ automatic-instances foldrTCons @-}
+{-@ foldrTCons :: q:Seq a -> as:[a] ->
+  { foldrT cons consT q as + pot (foldr cons q as)
+    - pot q <= 3 * len as + 1 } @-}
+foldrTCons :: Seq a -> [a] -> Proof
+foldrTCons _ [] = trivial
+foldrTCons q as@(x:xs) =
+  foldrT cons consT q as + pot (foldr cons q as)
+    - pot q
+  === consT x (foldr cons q xs)
+      + foldrT cons consT q xs
+      + pot (cons x (foldr cons q xs)) - pot q
+      + pot (foldr cons q xs)
+      - pot (foldr cons q xs)
+  ? consAmortized x (foldr cons q xs)
+  =<= 3 + foldrT cons consT q xs
+    + pot (foldr cons q xs) - pot q
+  ? foldrTCons q xs
+  =<= 3 + 3 * length xs + 1
   *** QED
-consP x t@(Single a) =
-    consT x t + phi (cons x t) - phi t
--- Use definition of cons/T _ (Single a)
-  === 1 + phi (Deep (One x) EmptyT (One a))
-        - phi (Single a)
--- Use definition of phi
-  === 1 + dang (One x) + phi EmptyT
-        + dang (One a) - 0
--- Use definition of phi and dang
-  === 1 + 1 + 0 + 1 - 0
--- 3 must be <= c, so c >= 3
+
+{-@ automatic-instances foldlTSnoc @-}
+{-@ foldlTSnoc :: q:Seq a -> as:[a] ->
+  { foldlT snoc snocT q as + pot (foldl snoc q as)
+    - pot q <= 3 * len as + 1 } @-}
+foldlTSnoc :: Seq a -> [a] -> Proof
+foldlTSnoc _ [] = trivial
+foldlTSnoc q as@(x:xs) =
+  foldlT snoc snocT q as + pot (foldl snoc q as)
+    - pot q
+  === snocT q x + foldlT snoc snocT (snoc q x) xs
+      + pot (foldl snoc (snoc q x) xs) - pot q
+      + pot (snoc q x) - pot (snoc q x)
+  ? snocAmortized q x
+  =<= 3 + foldlT snoc snocT (snoc q x) xs
+    + pot (foldl snoc (snoc q x) xs)
+    - pot (snoc q x)
+  ? foldlTSnoc (snoc q x) xs
   *** QED
@@
 
-We omit the proof cases for one to three digits, as they work exactly as the other two base cases.
-The constant however will not change in the other base cases, so $3$ really is our upper bound to use here.
-The interesting case is when there are four digits.
-In this case we need to use induction to prove our goal:
-
-\begin{lstlisting}
-consP x t@(Deep (Four a b c d) m sf) =
-    consT x t + phi (cons x t) - phi t
--- Use definition of consT
-  === (1 + consT (Node3 b c d) m)
-  + phi (cons x t) - phi t
--- Use definition of cons
-  === (1 + consT (Node3 b c d) m)
-  + phi (Deep (Two x a) (cons (Node3 b c d) m) sf)
-  - phi t
--- Use definition of phi
-  === (1 + consT (Node3 b c d) m)
-  + dang (Two x a) + phi (cons (Node3 b c d) m)
-  + dang sf
-  - (dang (Four a b c d) + phi m + dang sf)
--- Remove parenthesis
-  === 1 + consT (Node3 b c d) m
-  + dang (Two x a) + phi (cons (Node3 b c d) m)
-  + dang sf
-  - dang (Four a b c d) - phi m - dang sf
--- dang sf cancels, use definition of dang
-  === 1 + consT (Node3 b c d) m
-  + 0 + phi (cons (Node3 b c d) m) - 1 - phi m
--- 1s cancel
-  === consT (Node3 b c d) m
-  + phi (cons (Node3 b c d) m) - phi m
--- Use induction
-  ? consP (Node3 b c d) m
+Now for the actual proof we will only show one of the base cases as the others follow the same schema.
+All the base cases have a constant upper bound on their complexity.
+For the recursive case we also need some small helper facts about the logarithm such as monotonicity.
+
+\begin{lstlisting}
+{-@ log2Mono :: { x:Int | 1 <= x } ->
+  { y:Int | x <= y } -> { log2 x <= log2 y } @-}
+log2Mono :: Int -> Int -> Proof
+log2Mono 1 m = log2 1 <= log2 m *** QED
+log2Mono n m =
+  log2 n <= log2 m
+  === 1 + log2 (n `div` 2) <= 1 + log2 (m `div` 2)
+  === log2 (n `div` 2) <= log2 (m `div` 2)
+  ? log2Mono (n `div` 2) (m `div` 2)
   *** QED
-\end{lstlisting}
-
-\subsubsection{Append}\label{sec:append}
-
-Similar to the \texttt{push} operation on the stack, even though \texttt{append} has the same amortized complexity as its normal complexity, we need to show that it never increases the potential more than a logarithmic amount.
-Contrary to the simplified definition of Claessen~\cite{fingertrees_new}, the \textit{containers} package uses two sets of mutually recursive functions to implement \texttt{append}.
-One set of functions called \texttt{appendTreeX} handles the base cases where one of the two trees is either empty, or consists of a singular element by using \texttt{cons} and \texttt{snoc}.
-In the recursive case it calls the \texttt{addDigitsX} functions that group arguments as \texttt{Node}s and in turn call the \texttt{appendTreeX} functions.
-Because the \texttt{addDigitsX} functions do not pattern match on their tree arguments at all, LiquidHaskell can not find a termination metric automatically.
-For this, we need to track the arguments across function calls because only \texttt{appendTreeX} has a strictly decreasing argument.
-We can do so by adding a ghost parameter to order the functions for LiquidHaskell.
-This order is defined to be smaller for all \texttt{appendTreeX} functions than the \texttt{addDigitsX} functions.
-Additionally we need to set the case expansion to a high number such that all pattern matches are evaluated, else the termination proof would fail.
-The following snippet provides an example with most cases and function omitted for readability.
-
-\begin{lstlisting}
-{-@ LIQUID --max-case-expand=1000 @-}
-
-{-@ appendTree1 :: a:FingerTree (Node a) ->
-  Node a -> FingerTree (Node a) ->
-  FingerTree (Node a) / [size a, 1] @-}
-{-@ reflect appendTree1 @-}
-appendTree1 :: FingerTree (Node a) -> Node a ->
-    FingerTree (Node a) -> FingerTree (Node a)
-appendTree1 EmptyT !a xs = a `consTree` xs
-appendTree1 xs !a EmptyT = xs `snocTree` a
-appendTree1 (Single x) !a xs =
-    x `consTree` a `consTree` xs
-appendTree1 xs !a (Single x) =
-    xs `snocTree` a `snocTree` x
-appendTree1 (Deep pr1 m1 sf1) a (Deep pr2 m2 sf2) =
-    Deep pr1 (addDigits1 m1 sf1 a pr2 m2) sf2
-
-{-@ addDigits1 :: a:FingerTree (Node (Node a)) ->
-  Digit (Node a) -> Node a -> Digit (Node a) ->
-  FingerTree (Node (Node a)) ->
-  FingerTree (Node (Node a)) / [size a, 5] @-}
-{-@ reflect addDigits1 @-}
-addDigits1 :: FingerTree (Node (Node a)) ->
-  Digit (Node a) -> Node a -> Digit (Node a) ->
-  FingerTree (Node (Node a)) ->
-  FingerTree (Node (Node a))
-addDigits1 m1 (One a) b (One c) m2 =
-  appendTree1 m1 (Node3 a b c) m2
-addDigits1 m1 (One a) b (Two c d) m2 =
-  appendTree2 m1 (Node2 a b) (Node2 c d) m2
--- further cases omitted
-\end{lstlisting}
-
-This case expansion results in an explosion of constraints for the SMT solver to check.
-While the whole proof for \texttt{cons} including the defined functions generates fewer than 100 constraints, this termination proof generates over 3000.
-This leads to extremely high compile times which is further complicated by the need to define a timing function.
-The function needs to follow this mutually recursive structure and generates an additional 8000 constraints, such that loading the file takes several minutes.
-Another problem is the logarithmic runtime itself.
-SMT solvers are great with linear arithmetic, but are significantly slower when tasked with multiplication.
-Division is usually left as an unevaluated function.
-While it is possible to transform the proof goal to convert logarithms into powers, it still requires assumptions about basic arithmetic rules on the logic level.
-
-For those reasons, using LiquidHaskell to proof the complexity with the physicist's method is not feasible (see section~\ref{sec:limitations} for a discussion of those limitations).
-Claessen~\cite{fingertrees_new} does not formally prove this in his functional pearl as well, but provides a different reasoning to why \texttt{append} is logarithmic.
-The potential only depends on the spine of the tree, so a new layer in the tree only adds at most a constant amount to the potential ($2$ if both digits are dangerous) making it grow linearly.
-Because the nodes in the digits always at least double (in the case of two \texttt{Node2}) the overall size of the tree grows exponentially.
-Thus the potential grows logarithmic in terms of the size of the tree.
-
-\section{Limitations of our approach}\label{sec:limitations}
-
-\subsection{Definition of the timing function}
-
-The analysis could lead to wrong results is if the timing function does not directly mirror the original function.
-When the implementation changes and the timing function does not, it might be possible for the proofs to still pass despite being invalid.
-While this is unlikely, given how verbose proofs with LiquidHaskell are, it is not negligible.
-
-If it is possible to change the implementation, it would be possible to factor out the recursion and share it between the implementation and the timing function.
-This however would clutter the implementation and most likely decrease performance as it is often the case with recursion schemes.
-
-A better solution is to derive the timing function using meta programming in the form of \textit{TemplateHaskell}.
-Given the name of the function in the implementation, TemplateHaskell would allow to create a new function with a configurable suffix.
-This can be done by checking if the right hand sides of the original function contain recursive calls to the given name.
-If not, replace the right hand side with $1$, otherwise change the recursion to use the name with the suffix and add $1$ at the end.
-The resulting timing function is then be spliced back into the source code:
-
-\begin{lstlisting}
-cons :: a -> FingerTree a -> FingerTree a
-cons a EmptyT = Single a
-cons a (Deep (Four b c d e) m sf) =
-    Deep (Two a b) (cons (Node3 c d e) m) sf
--- other cases omitted
-
-$(mkTimingFn "T" ''cons)
--- generates this; variable names omitted:
--- consT :: a -> FingerTree a -> Int
--- consT _ EmptyT = 1
--- consT _ (Deep (Four _ c d e) m _) =
---    consT (Node3 c d e) m + 1
--- other cases omitted
-\end{lstlisting}
-
-\subsection{LiquidHaskell for verification}
-
-As shown in earlier sections, at the time of writing LiquidHaskell has some limitations with regards to proving more complex functions.
-First and foremost the support for type classes and Haskell \texttt{newtype}s is insufficient.
-Currently, LiquidHaskell only supports measure classes, which need to be defined by recursion over a single data type similar to a normal measure.
-For this reason it is not possible to use them for the size parameter in finger trees (see section~\ref{sec:fingertree}).
-There are several open tickets concerning the type class support on the LiquidHaskell bug tracker, so we expect this to improve in the future.
-
-Another road block was the combinatorial explosion when generating the constraints for the termination of \texttt{append} (see section~\ref{sec:append}).
-It might be possible to improve the constraint generation and thus create less work for the SMT solver.
-LiquidHaskell also provides an option to disable the termination checker, but with this option the compile time surprisingly did not change.
-Similarly, several other options that are supposed to improve performance like \texttt{{-}{-}diff}, which enables incremental type checking or \texttt{{-}{-}fast}, which disables inference of polymorphic refinement, did either not affect performance at all or (in the case of \texttt{{-}{-}fast}) made it even worse.
+
+{-@ divCancel :: x:Int ->
+  { div (2 * x) 2 == x } @-}
+divCancel :: Int -> Proof
+divCancel _ = trivial
+
+{-@ glueAmortized :: q1:Seq a
+  -> { as:[a] | len as <= 3 }
+  -> q2:Seq a -> {
+      glueT q1 as q2 + pot (glue q1 as q2)
+      - pot q1 - pot q2
+    <= log2 (max (len (seqToList q1)
+      + len (seqToList q2)) 2) + 14
+  } @-}
+glueAmortized :: Seq a -> [a] -> Seq a -> Proof
+glueAmortized q1@Nil as q2 =
+  glueT q1 as q2 + pot (glue q1 as q2)
+    - pot q1 - pot q2
+  === 1 + foldrT cons consT q2 as
+    + pot (foldr cons q2 as) - pot q1 - pot q2
+  ? foldrTCons q2 as
+  =<= 1 + 3 * length as + 1 - pot q1
+  =<= log2 (max (length (seqToList q1)
+    + length (seqToList q2)) 2) + 14
+  *** QED
+glueAmortized qq1@(More u1 q1 v1) as
+  qq2@(More u2 q2 v2) =
+  glueT qq1 as qq2 + pot (glue qq1 as qq2)
+  - pot qq1 - pot qq2
+  === 1 + glueT q1 (toTuples
+        (toList v1 ++ as ++ toList u2)) q2
+      + pot (More u1 (glue q1 (toTuples
+          (toList v1 ++ as ++ toList u2)) q2) v2)
+      - danger u1 - pot q1 - danger v1 - danger u2
+      - pot q2 - danger v2
+  === 1 + glueT q1 (toTuples
+          (toList v1 ++ as ++ toList u2)) q2
+      + pot (glue q1 (toTuples
+          (toList v1 ++ as ++ toList u2)) q2)
+      - pot q1 - danger v1 - danger u2 - pot q2
+-- apply induction
+  ? glueAmortized q1
+      (toTuples (toList v1 ++ as ++ toList u2)) q2
+  =<= log2 (max (length (seqToList q1)
+        + length (seqToList q2)) 2) + 15
+-- move (+1) into the logarithm
+  ? divCancel (max (length (seqToList q1)
+      + length (seqToList q2)) 2)
+  === 1 + log2 (2 * max (length (seqToList q1)
+      + length (seqToList q2)) 2 `div` 2) + 14
+  === log2 (2 * max (length (seqToList q1)
+      + length (seqToList q2)) 2) + 14
+-- simplify
+  ? log2Mono (2 * max (length (seqToList q1)
+      + length (seqToList q2)) 2)
+    (4 + 2 * length (seqToList q1)
+      + 2 * length (seqToList q2))
+  =<= log2 (4 + 2 * length (seqToList q1)
+      + 2 * length (seqToList q2)) + 14
+  ? log2Mono (4 + 2 * length (seqToList q1)
+    + 2 * length (seqToList q2))
+      (4 + length (tuplesToList (seqToList q1))
+        + length (tuplesToList (seqToList q2)))
+  =<= log2 (4 + length (tuplesToList (seqToList q1))
+      + length (tuplesToList (seqToList q2))) + 14
+  ? log2Mono (4 + length (tuplesToList (seqToList q1))
+    + length (tuplesToList (seqToList q2))) 
+      (max (length (toList u1 ++ tuplesToList
+          (seqToList q1) ++ toList v1)
+        + length (toList u2 ++ tuplesToList
+            (seqToList q2) ++ toList v2)) 2)
+  =<= log2 (max (length (toList u1 ++ tuplesToList
+        (seqToList q1) ++ toList v1)
+      + length (toList u2 ++ tuplesToList
+        (seqToList q2) ++ toList v2)) 2) + 14
+  === log2 (max (length (seqToList qq1)
+      + length (seqToList qq2)) 2) + 14
+  *** QED
+\end{lstlisting}
 
@@
 
-Despite the current limitations of LiquidHaskell preventing more complex functions from being formally proven, it drastically improves the readability of proofs compared to other theorem provers.
-The explicit equational reasoning of LiquidHaskell makes following the proof a lot easier than in tactic based provers like Coq and Lean.
-While Isabelle/HOL also features equational reasoning with Isar~\cite{isar}, it does not support polymorphic recursion which makes the definition of our second example a lot harder.
-
-While at the moment amortized complexity of \O{1} seems to be the most feasable to prove, we expect the tooling around LiquidHaskell to improve a lot in the future, be it by incremental changes like better support for type classes, or by more complicated means like a deeper embedding into a future dependently typed Haskell core.
-The recent improvements in the usability of LiquidHaskell --- that were achieved by turning the previously standalone executable into a GHC plugin~\cite{lh_plugin} --- are already a big step into this future by drastically reducing the barrier of entry.
+In this paper we successfully applied LiquidHaskell to prove the amortized complexity of increasingly complex data structures.
+Despite some limitations of LiquidHaskell, like incomplete support for \texttt{DataKinds}, proving amortized complexity directly in Haskell is quite feasible.
+However some more complex proofs might require a lot of code for rather simple proof steps (compare the use of monotonicity in the previous section).
+In such cases, the extra automation built into modern theorem provers would certainly improve the user experience.
+
+On the other hand, LiquidHaskell's explicit equational reasoning is easier to follow for a new learner than opaque proof scripts written in a tactic-based theorem prover like Coq or Lean.
+While Isabelle/HOL also features equational reasoning with Isar~\cite{isar}, it does not support polymorphic recursion which makes the definition of our third case study a lot harder.
+
+For future work, it would be interesting to investigate how complex an adaptation of our technique for reasoning about lazy semantics would be.
+This would allow to expand the complexity results to a shared setting where reuse of intermediate results is allowed.
+It might be possible to adapt Yao et al.'s work on complexity for lazy programs~\cite{forking_paths}, but LiquidHaskell's lack of quantifiers will make this challenging.
 
@@
 
-\section{Full proof code for the stack}
-
-\begin{lstlisting}
-import Language.Haskell.Liquid.ProofCombinators
-
-data Stack a = Empty | Elem a (Stack a)
-
-{-@ reflect push @-}
-push :: a -> Stack a -> Stack a
-push = Elem
-
-{-@ multipop :: Nat -> Stack a -> Stack a @-}
-multipop :: Int -> Stack a -> Stack a
-multipop _ Empty = Empty
-multipop 0 s = s
-multipop n (Elem _ s) = multipop (n - 1) s
-
-{-@ reflect phi @-}
-{-@ phi :: Stack a -> Nat @-}
-phi :: Stack a -> Int
-phi Empty = 0
-phi (Elem _ s) = 1 + phi s
-
-{-@ reflect pushT @-}
-{-@ pushT :: a -> Stack a -> Nat @-}
-pushT :: a -> Stack a -> Int
-pushT _ _ = 1
-
-{-@ reflect multipopT @-}
-{-@ multipopT :: Nat -> Stack a -> Nat @-}
-multipopT :: Int -> Stack a -> Int
-multipopT _ Empty = 1
-multipopT 0 _ = 1
-multipopT n (Elem _ s) = 1 + multipopT (n - 1) s
-
-{-@ pushP :: x:a -> s:Stack a ->
-  { pushT x s + phi (push x s) - phi s <= 2 }
-@-}
-pushP :: a -> Stack a -> Proof
-pushP x s = pushT x s + phi (push x s) - phi s
--- Use definition of push and pushT
-  === 1 + phi (Elem x s) - phi s
--- Use definition of phi (Elem _ _)
-  === 1 + (1 + phi s) - phi s
--- phi s gets canceled, 1 + 1 <= 2
-  *** QED
-
-{-@ multipopP :: k:Nat -> s:Stack a ->
-  { multipopT k s + phi (multipop k s)
-      - phi s <= 2 }
-@-}
-multipopP :: Int -> Stack a -> Proof
-multipopP 0 s = multipopT 0 s
-    + phi (multipop 0 s) - phi s
--- Use definition of multipop/T 0 _
-  === 1 + phi s - phi s
--- phi s cancels, 1 <= 2
-  *** QED
-multipopP k Empty = multipopT k Empty
-    + phi (multipop k Empty) - phi Empty
--- Use definition of multipop/T _ Empty
-  === 1 + phi Empty - phi Empty
--- phi Empty cancels, 1 <= 2
-  *** QED
-multipopP k xs@(Elem x s) = multipopT k xs
-    + phi (multipop k xs) - phi xs
--- Use definition of multipopT for recusive case
-  === (1 + multipopT (k - 1) s)
-    + phi (multipop k xs) - phi xs
--- Use definition of multipop for recusive case
-  === (1 + multipopT (k - 1) s)
-    + phi (multipop (k - 1) s) - phi xs
--- Use definition of phi (Elem _ _)
-  === (1 + multipopT (k - 1) s)
-    + phi (multipop (k - 1) s) - (1 + phi s)
--- Remove parenthesis
-  === 1 + multipopT (k - 1) s
-    + phi (multipop (k - 1) s) - 1 - phi s
--- 1s cancel out
-  === multipopT (k - 1) s
-    + phi (multipop (k - 1) s) - phi s
--- use induction
-  ? multipopP (k - 1) s
-  *** QED
-\end{lstlisting}
-
-\section{Full proof code of the finger tree}\label{ap:fingertree}
-
-\begin{lstlisting}
-data FingerTree a = Empty
-  | Single a
-  | Deep (Digit a) (FingerTree (Node a)) (Digit a)
-
-data Node a = Node2 a a | Node3 a a a
-
-data Digit a = One a
-  | Two a a
-  | Three a a a
-  | Four a a a a
-
-{-@ reflect cons @-}
-cons :: a -> FingerTree a -> FingerTree a
-cons a EmptyT = Single a
-cons a (Single b) =
-  Deep (One a) EmptyT (One b)
-cons a (Deep (One b) m sf) =
-  Deep (Two a b) m sf
-cons a (Deep (Two b c) m sf) =
-  Deep (Three a b c) m sf
-cons a (Deep (Three b c d) m sf) =
-  Deep (Four a b c d) m sf
-cons a (Deep (Four b c d e) m sf) =
-  Deep (Two a b) (cons (Node3 c d e) m) sf
-
+\section{Correct-by-construction binomial heap}\label{app:datakinds}
+
+\begin{lstlisting}
+import Data.Kind (Type)
+
+data Nat = Zero | Succ Nat
+
+type Tree :: Nat -> Type -> Type
+data Tree k a where
+  MkTree :: a -> DecrList Tree k a -> Tree k a
+
+type DecrList :: (Nat -> Type -> Type)
+  -> Nat -> Type -> Type
+data DecrList t k a where
+  DNil :: DecrList t Zero a
+  DCons :: t k a -> DecrList t k a
+    -> DecrList t (Succ k) a
+
+mergeTree :: Ord a => Tree k a -> Tree k a
+  -> Tree (Succ k) a
+mergeTree l@(MkTree lr lc) r@(MkTree rr rc)
+  | lr <= rr = MkTree lr (DCons r lc)
+  | otherwise = MkTree rr (DCons l rc)
+
+data Binary
+  = B0 Binary
+  | B1 Binary
+  | BEnd
+
+type BInc :: Binary -> Binary
+type family BInc b where
+  BInc BEnd = B1 BEnd
+  BInc (B0 rest) = B1 rest
+  BInc (B1 rest) = B0 (BInc rest)
+
+type Forest :: Nat -> Binary -> Type -> Type
+data Forest k b a where
+  FEnd :: Forest k BEnd a
+  F0 :: Forest (Succ k) b a -> Forest k (B0 b) a
+  F1 :: Tree k a -> Forest (Succ k) b a
+    -> Forest k (B1 b) a
+
+type Heap :: Binary -> Type -> Type
+newtype Heap b a = MkHeap (Forest Zero b a)
+
+insert :: Ord a => Tree k a -> Forest k b a
+  -> Forest k (BInc b) a
+insert t FEnd = F1 t FEnd
+insert t (F0 f) = F1 t f
+insert t (F1 t' f) = F0 (insert (mergeTree t t') f)
+\end{lstlisting}
+
+\section{Definition and proof for the snoc operation on a finger tree}
+
+\begin{lstlisting}
 {-@ reflect snoc @-}
-snoc :: FingerTree a -> a -> FingerTree a
-snoc EmptyT x = Single x
-snoc (Single a) x =
-  Deep (One a) EmptyT (One x)
-snoc (Deep pr m (One a)) x =
-  Deep pr m (Two a x)
-snoc (Deep pr m (Two a b)) x =
-  Deep pr m (Three a b x)
-snoc (Deep pr m (Three a b c)) x =
-  Deep pr m (Four a b c x)
-snoc (Deep pr m (Four a b c d)) x =
-  Deep pr (snoc m (Node3 a b c)) (Two d x)
-
-{-@ reflect dang @-}
-{-@ dang :: Digit a -> Nat @-}
-dang :: Digit a -> Int
-dang One{} = 1
-dang Two{} = 0
-dang Three{} = 0
-dang Four{} = 1
-
-{-@ reflect phi @-}
-{-@ phi :: FingerTree a -> Nat @-}
-phi :: FingerTree a -> Int
-phi EmptyT = 0
-phi Single{} = 0
-phi (Deep pr m sf) =
-  dang pr + phi m + dang sf
-
-{-@ reflect consT @-}
-{-@ consT :: a -> FingerTree a -> Nat @-}
-consT :: a -> FingerTree a -> Int
-consT _ EmptyT = 1
-consT _ Single{} = 1
-consT _ (Deep One{} _ _) = 1
-consT _ (Deep Two{} _ _) = 1
-consT _ (Deep Three{} _ _) = 1
-consT _ (Deep (Four _ c d e) m _) =
-  1 + consT (Node3 c d e) m
+snoc :: Seq a -> a -> Seq a
+snoc Nil x = Unit x
+snoc (Unit x) y = More (One x) Nil (One y)
+snoc (More u q (One x)) y = More u q (Two x y)
+snoc (More u q (Two x y)) z = More u q (Three x y z)
+snoc (More u q (Three x y z)) w =
+  More u (snoc q (Pair x y)) (Two z w)
 
 {-@ reflect snocT @-}
-{-@ snocT :: FingerTree a -> a -> Nat @-}
-snocT :: FingerTree a -> a -> Int
-snocT EmptyT _ = 1
-snocT Single{} _ = 1
-snocT (Deep _ _ One{}) _ = 1
-snocT (Deep _ _ Two{}) _ = 1
-snocT (Deep _ _ Three{}) _ = 1
-snocT (Deep _ m (Four a b c _)) _ =
-    1 + snocT m (Node3 a b c)
-
-{-@ consP :: x:a -> t:FingerTree a ->
-  { consT x t + phi (cons x t) - phi t <= 3 }
-@-}
-consP :: a -> FingerTree a -> Proof
-consP x t@EmptyT =
-    consT x t + phi (cons x t) - phi t
--- Use definition of cons/T _ EmptyT
-  === 1 + phi (Single x) - phi EmptyT
--- Use definition of phi
-  === 1 + 0 - 0 -- trivial, 1 <= 3
-  *** QED
-consP x t@(Single a) =
-    consT x t + phi (cons x t) - phi t
--- Use definition of cons/T _ (Single a)
-  === 1 + phi (Deep (One x) EmptyT (One a))
-        - phi (Single a)
--- Use definition of phi
-  === 1 + dang (One x) + phi EmptyT
-        + dang (One a) - 0
--- Use definition of phi and dang
-  === 1 + 1 + 0 + 1 - 0 -- trivial, 3 <= 3
-  *** QED
-consP x t@(Deep (One a) m sf) =
-    consT x t + phi (cons x t) - phi t
--- Use definition of cons/T
-  === 1 + phi (Deep (Two x a) m sf) - phi t
--- Use definition of phi
-  === 1 + dang (Two x a) + phi m + dang sf
-    - (dang (One a) + phi m + dang sf)
--- Remove parenthesis, use definition of dang
-  === 1 + 0 + phi m + dang sf
-    - 1 - phi m - dang sf
--- everything cancels
-  === 0 -- trivial, 0 <= 3
-  *** QED
-consP x t@(Deep (Two a b) m sf) =
-    consT x t + phi (cons x t) - phi t
--- Use definition of cons/T
-  === 1 + phi (Deep (Three x a b) m sf) - phi t
--- Use definition of phi
-  === 1 + dang (Three x a b) + phi m + dang sf
-    - (dang (Two a b) + phi m + dang sf)
--- Remove parenthesis, use definition of dang
-  === 1 + 0 + phi m + dang sf
-    - 0 - phi m - dang sf
--- phi and dang cancels
-  === 1 -- trivial, 1 <= 3
-  *** QED
-consP x t@(Deep (Three a b c) m sf) =
-    consT x t + phi (cons x t) - phi t
--- Use definition of cons/T
-  === 1 + phi (Deep (Four x a b c) m sf) - phi t
--- Use definition of phi
-  === 1 + dang (Four x a b c) + phi m + dang sf
-    - (dang (Three a b c) + phi m + dang sf)
--- Remove parenthesis, use definition of dang
-  === 1 + 1 + phi m + dang sf
-    - 0 - phi m - dang sf
--- phi and dang cancels
-  === 2 -- trivial, 2 <= 3
-  *** QED
-consP x t@(Deep (Four a b c d) m sf) =
-    consT x t + phi (cons x t) - phi t
--- Use definition of consT
-  === (1 + consT (Node3 b c d) m)
-  + phi (cons x t) - phi t
--- Use definition of cons
-  === (1 + consT (Node3 b c d) m)
-  + phi (Deep (Two x a) (cons (Node3 b c d) m) sf)
-  - phi t
--- Use definition of phi
-  === (1 + consT (Node3 b c d) m)
-  + dang (Two x a) + phi (cons (Node3 b c d) m)
-  + dang sf
-  - (dang (Four a b c d) + phi m + dang sf)
--- Remove parenthesis
-  === 1 + consT (Node3 b c d) m
-  + dang (Two x a) + phi (cons (Node3 b c d) m)
-  + dang sf
-  - dang (Four a b c d) - phi m - dang sf
--- dang sf cancels, use definition of dang
-  === 1 + consT (Node3 b c d) m
-  + 0 + phi (cons (Node3 b c d) m) - 1 - phi m
--- 1s cancel
-  === consT (Node3 b c d) m
-  + phi (cons (Node3 b c d) m) - phi m
--- Use induction
-  ? consP (Node3 b c d) m
-  *** QED
-
-{-@ snocP :: t:FingerTree a -> x:a ->
-  { snocT t x + phi (snoc t x) - phi t <= 3 }
-@-}
-snocP :: a -> FingerTree a -> Proof
-snocP t@EmptyT x =
-    snocT t x + phi (snoc t x) - phi t
--- Use definition of snoc/T EmptyT _
-  === 1 + phi (Single x) - phi EmptyT
--- Use definition of phi
-  === 1 + 0 - 0 -- trivial, 1 <= 3
-  *** QED
-snocP t@(Single a) x =
-    snocT t x + phi (snoc t x) - phi t
--- Use definition of snoc/T (Single a) _
-  === 1 + phi (Deep (One a) EmptyT (One x))
-        - phi (Single a)
--- Use definition of phi
-  === 1 + dang (One a) + phi EmptyT
-        + dang (One x) - 0
--- Use definition of phi and dang
-  === 1 + 1 + 0 + 1 - 0 -- trivial, 3 <= 3
-  *** QED
-snocP t@(Deep pr m (One a)) x =
-    snocT t x + phi (snoc t x) - phi t
--- Use definition of snoc/T
-  === 1 + phi (Deep pr m (Two a x)) - phi t
--- Use definition of phi
-  === 1 + dang pr + phi m + dang (Two a x)
-    - (dang pr + phi m + dang (One a))
--- Remove parenthesis, use definition of dang
-  === 1 + dang pr + phi m + 0
-    - dang pr - phi m - 1
--- everything cancels
-  === 0 -- trivial, 0 <= 3
-  *** QED
-snocP t@(Deep pr m (Two a b)) x =
-    snocT t x + phi (snoc t x) - phi t
--- Use definition of snoc/T
-  === 1 + phi (Deep pr m (Three a b x)) - phi t
--- Use definition of phi
-  === 1 + dang pr + phi m + dang (Three a b x)
-    - (dang pr + phi m + dang (Two a b))
--- Remove parenthesis, use definition of dang
-  === 1 + dang pr + phi m + 0
-    - dang pr - phi m - 0
--- phi and dang cancels
-  === 1 -- trivial, 1 <= 3
-  *** QED
-snocP t@(Deep pr m (Three a b c)) x =
-    snocT t x + phi (snoc t x) - phi t
--- Use definition of snoc/T
-  === 1 + phi (Deep pr m (Four a b c x)) - phi t
--- Use definition of phi
-  === 1 + dang pr + phi m + dang (Four a b c x)
-    - (dang pr + phi m + dang (Three a b c))
--- Remove parenthesis, use definition of dang
-  === 1 + dang pr + phi m + 1
-    - dang pr - phi m - 0
--- phi and dang cancels
-  === 2 -- trivial, 2 <= 3
-  *** QED
-snocP t@(Deep pr m (Four a b c d)) x =
-    snocT t x + phi (snoc t x) - phi t
--- Use definition of snocT
-  === (1 + snocT m (Node3 a b c))
-  + phi (snoc t x) - phi t
--- Use definition of snoc
-  === (1 + snocT m (Node3 a b c))
-  + phi (Deep pr (snoc m (Node3 a b c)) (Two x a))
-  - phi t
--- Use definition of phi
-  === (1 + snocT m (Node3 a b c))
-  + dang pr + phi (snoc m (Node3 a b c))
-  + dang (Two x a)
-  - (dang pr + phi m + dang (Four a b c d))
--- Remove parenthesis
-  === 1 + snocT m (Node3 a b c)
-  + dang pr + phi (snoc m (Node3 a b c))
-  + dang (Two x a)
-  - dang pr - phi m - dang (Four a b c d))
--- dang pr cancels, use definition of dang
-  === 1 + snocT m (Node3 a b c)
-  + phi (snoc m (Node3 a b c)) + 0 - phi m - 1
--- 1s cancel
-  === snocT m (Node3 a b c)
-  + phi (snoc m (Node3 a b c)) - phi m
--- Use induction
-  ? snocP m (Node3 a b c)
-  *** QED
+{-@ snocT :: Seq a -> a -> { x:Int | x >= 1 } @-}
+snocT :: Seq a -> a -> Int
+snocT Nil _ = 1
+snocT (Unit _) _ = 1
+snocT (More _ _ (One _)) _ = 1
+snocT (More _ _ (Two _ _)) _ = 1
+snocT (More _ q (Three x y _)) _ =
+  1 + snocT q (Pair x y)
+
+{-@ automatic-instances snocAmortized @-}
+{-@ snocAmortized :: q:Seq a -> x:a ->
+  { snocT q x + pot (snoc q x) - pot q <= 3 } @-}
+snocAmortized :: Seq a -> a -> Proof
+snocAmortized (More _ q (Three x y _)) _ =
+  trivial ? snocAmortized q (Pair x y)
+snocAmortized _ _ = trivial
 

### 2024-07-01 code/src/Heap.hs

@@ -51,7 +51,7 @@ type family BAdd (x :: Binary) (y :: Binary) :: Binary where
   BAdd (B0 x) (B1 y) = B1 (BAdd x y)
   BAdd (B1 x) (B1 y) = B0 (BInc (BAdd x y))
 
-type Forest :: Nat -> Binary -> Type -> Type
+--type Forest :: Nat -> Binary -> Type -> Type
 {-data Forest k b a where
   FEnd :: Forest k BEnd a
   F0 :: Forest (Succ k) b a -> Forest k (B0 b) a
@@ -106,7 +106,8 @@ pot (F1 _ rest) = 1 + pot rest
 insertTreeAmortized :: Ord a => Tree k a -> Forest k b a -> Proof
 insertTreeAmortized t f@FEnd =
   insertTreeT t f + pot (insertTree t f) - pot f
-    === 1 + pot (F1 t FEnd) - 0
+    === 1 + pot (insertTree t f) - 0
+    === 1 + pot (F1 t FEnd) -- fails
     === undefined
     *** QED
 {-    === 1 + pot (F1 t FEnd) - 0

### 2024-07-01 code/src/Heap2.hs

@@ -4,7 +4,7 @@ module Heap2 where
 
 import Prelude hiding (replicate)
 import Data.List (uncons)
-import Language.Haskell.Liquid.ProofCombinators (Proof, QED (..), (***), (===), (?))
+import Language.Haskell.Liquid.ProofCombinators (Proof, (?), trivial)
 
 data Tree a = MkTree a [Tree a]
 
@@ -56,14 +56,8 @@ pot (F1 _ rest) = 1 + pot rest
 {-@ automatic-instances insertTreeAmortized @-}
 {-@ insertTreeAmortized :: t:_ -> f:_ -> { insertTreeT t f + pot (insertTree t f) - pot f <= 2 } @-}
 insertTreeAmortized :: Ord a => Tree a -> Heap a -> Proof
-insertTreeAmortized t f@FEnd = insertTreeT t f + pot (insertTree t f) - pot f *** QED
-insertTreeAmortized t f@(F0 _) = insertTreeT t f + pot (insertTree t f) - pot f *** QED
-insertTreeAmortized t f@(F1 t' f') =
-  insertTreeT t f + pot (insertTree t f) - pot f
-  === 1 + insertTreeT (mergeTree t t') f' + pot (F0 (insertTree (mergeTree t t') f')) - 1 - pot f'
-  === insertTreeT (mergeTree t t') f' + pot (insertTree (mergeTree t t') f') - pot f'
-  ? insertTreeAmortized (mergeTree t t') f'
-  *** QED
+insertTreeAmortized t (F1 t' f') = trivial ? insertTreeAmortized (mergeTree t t') f'
+insertTreeAmortized _ _ = trivial
 
 -- Appendix: Pretty print trees
 mapToList :: (Tree a -> b) -> [Tree a] -> [b]

### 2024-07-01 code/src/Main.hs

@@ -1,13 +1,26 @@
+{-@ LIQUID "--ple-local" @-}
+
 module Main where
 
-import Language.Haskell.Liquid.ProofCombinators (Proof, QED (..), (***), (===), (?))
+import Language.Haskell.Liquid.ProofCombinators (Proof, QED (..), (***), (===), (?), trivial)
 import Stack ()
-import Prelude hiding (reverse)
+import Prelude hiding (reverse, length)
+
+{-@ measure length @-}
+length :: [a] -> Int
+length [] = 0
+length (_:xs) = 1 + length xs
 
 {-@ append :: xs:[a] -> ys:[a] -> { zs:[a] | len zs == len xs + len ys } @-}
 append :: [a] -> [a] -> [a]
 append [] ys = ys
 append (x : xs) ys = x : append xs ys
+
+{-@ automatic-instances appendLengthP @-}
+{-@ appendLengthP :: xs:[a] -> ys:[a] -> { length (append xs ys) == length xs + length ys } @-}
+appendLengthP :: [a] -> [a] -> Proof
+appendLengthP [] _ = trivial
+appendLengthP (_ : xs) ys = trivial ? appendLengthP xs ys
 
 {-@ appendAssocP :: xs:[a] -> ys:[a] -> zs:[a] ->
       { append (append xs ys) zs == append xs (append ys zs) } @-}

### 2024-07-01 code/src/Seq2.hs

@@ -7,7 +7,7 @@ module Seq2 where
 
 import Prelude hiding (foldl, foldr, (++), max)
 
-import Language.Haskell.Liquid.ProofCombinators (Proof, QED (..), (***), (=<=), (===), (?))
+import Language.Haskell.Liquid.ProofCombinators (Proof, QED (..), (***), (=<=), (===), (?), trivial)
 
 data Seq a = Nil
   | Unit a
@@ -153,16 +153,8 @@ consAmortized x qq@(More (Three y z w) q u) =
 {-@ automatic-instances snocAmortized @-}
 {-@ snocAmortized :: q:Seq a -> x:a -> { snocT q x + pot (snoc q x) - pot q <= 3 } @-}
 snocAmortized :: Seq a -> a -> Proof
-snocAmortized qq@(More u q (Three x y z)) w =
-  snocT qq w + pot (snoc qq w) - pot qq
-  === 1 + snocT q (Pair x y) + pot (More u (snoc q (Pair x y)) (Two z w)) - dang u - pot q - 1
-  === snocT q (Pair x y) + pot (snoc q (Pair x y)) - pot q
-  ? snocAmortized q (Pair x y)
-  *** QED
-snocAmortized qq w =
-  snocT qq w + pot (snoc qq w) - pot qq
-  =<= 3
-  *** QED
+snocAmortized (More _ q (Three x y _)) _ = trivial ? snocAmortized q (Pair x y)
+snocAmortized _ _ = trivial
 
 {-@ reflect foldrT @-}
 {-@ foldrT :: (a -> b -> b) -> (a -> b -> { x:Int | x >= 1 }) -> b -> [a] -> { x:Int | x >= 1 } @-}
@@ -214,12 +206,10 @@ log2Mono n m =
   ? log2Mono (n `div` 2) (m `div` 2)
   *** QED
 
+{-@ automatic-instances foldrTCons @-}
 {-@ foldrTCons :: q:Seq a -> as:[a] -> { foldrT cons consT q as + pot (foldr cons q as) - pot q <= 3 * len as + 1 } @-}
 foldrTCons :: Seq a -> [a] -> Proof
-foldrTCons q as@[] =
-  foldrT cons consT q as + pot (foldr cons q as) - pot q
-  === 1 + pot q - pot q
-  *** QED
+foldrTCons _ [] = trivial
 foldrTCons q as@(x:xs) =
[…]

### 2024-07-01 code/src/Stack.hs

@@ -1,6 +1,8 @@
+{-@ LIQUID "--ple-local" @-}
+
 module Stack (Stack, push, multipop) where
 
-import Language.Haskell.Liquid.ProofCombinators (Proof, QED (..), (***), (===), (?))
+import Language.Haskell.Liquid.ProofCombinators (Proof, QED (..), (***), (===), (?), trivial)
 import Prelude hiding (min)
 
 data Stack a = Nil | Cons a (Stack a)
@@ -73,20 +75,11 @@ multipop2 0 xs = ([], xs)
 multipop2 _ Nil = ([], Nil)
 multipop2 n (Cons x xs) = case multipop2 (n - 1) xs of (l, s) -> (x : l, s)
 
+{-@ automatic-instances ammortized_pop2 @-}
 {-@ ammortized_pop2 :: { n:Int | n >= 0 } -> st:Stack a -> { t_multipop n st + phi (snd (multipop2 n st)) - phi st == c2 } @-}
 ammortized_pop2 :: Int -> Stack a -> Proof
-ammortized_pop2 0 st =
-  t_multipop 0 st + phi (snd (multipop2 0 st)) - phi st
-    === 0 + phi (snd ([], st)) - phi st
-    === 0 + phi st - phi st
-    === c2
-    *** QED
-ammortized_pop2 n Nil =
-  t_multipop n Nil + phi (snd (multipop2 n Nil)) - phi Nil
-    === 0 + phi (snd ([], Nil)) - phi Nil
-    === 0 + 0 - 0
-    === c2
-    *** QED
+ammortized_pop2 0 _ = trivial
+ammortized_pop2 _ Nil = trivial
 ammortized_pop2 n (Cons x xs) =
   t_multipop n (Cons x xs) + phi (snd (multipop2 n (Cons x xs))) - phi (Cons x xs)
     === t_multipop n (Cons x xs) + phi (snd (case multipop2 (n - 1) xs of (l, st') -> (x : l, st'))) - phi (Cons x xs)

### 2024-07-18 paper/main.tex
Refine paper with minor comments

@@
-\documentclass[sigplan,screen,review,anonymous]{acmart}
+\documentclass[sigplan,screen]{acmart}
 
@@
     basicstyle=\ttfamily,
-    commentstyle=\color{codegreen}
+    commentstyle=\color{codegreen},
+    deletekeywords={Int, otherwise, div, length, snd, foldl, foldr, max, insert}
 }
@@
 
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{none}
-%\copyrightyear{2024}
+\setcopyright{rightsretained}
+\acmDOI{10.1145/3677999.3678282}
 \acmYear{2024}
-\acmDOI{tbd}
-
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[Haskell '24]{the 17th ACM SIGPLAN International Haskell Symposium}{September 02--05, 2024}{Milan, Italy}
-%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%  June 03--05, 2018, Woodstock, NY}
-%\acmPrice{15.00}
-%\acmISBN{978-1-4503-XXXX-X/18/06}
+\copyrightyear{2024}
+\acmISBN{979-8-4007-1102-2/24/09}
+\acmConference[Haskell '24]{Proceedings of the 17th ACM SIGPLAN International Haskell Symposium}{September 6--7, 2024}{Milan, Italy}
+\acmBooktitle{Proceedings of the 17th ACM SIGPLAN International Haskell Symposium (Haskell '24), September 6--7, 2024, Milan, Italy}
+\acmSubmissionID{icfpws24haskellmain-p94-p}
+\received{2024-06-03}
+\received[accepted]{2024-07-05}
 
@@
 
-%%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
-\title{Liquid Amortization - Proving amortized complexity with LiquidHaskell (Functional Pearl)}
-
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
+\title{Liquid Amortization: Proving Amortized Complexity with LiquidHaskell (Functional Pearl)}
+
 \author{Jan van Brügge}
@@
   \institution{Heriot-Watt University}
-  %\streetaddress{Arcisstraße 21}
   \city{Edinburgh}
-  %\state{Bavaria}
   \country{United Kingdom}
-  %\postcode{80333}
 }
@@
 As the performance can vary wildly for the same operation depending on the state of the data structure, usually the worst case for each operation is used as a metric and specified using \textit{Landau notation} ("big O").
-For example, a linear search\footnote{walking through a list element by element, checking if the current element is the one that is searched for} takes $1$ unit of time if the searched element is at the front, while it takes $n$ units of time if it is the last element.
+For example, a linear search\footnote{walking through a list element by element, checking if the current element is the one that is searched for} takes $1$ unit of time if the searched element is at the front, while it takes $n$ units of time if it is the last element (where $n$ is the number of elements in the list).
 Thus we would say that linear search is in the complexity class of \O{n} because it will not take longer than $c \cdot n$ units of time for some fixed constant $c$.
@@
 Because for every element pushed we deposit one extra time unit into the bank, the account always contains as many time units as there are elements on the stack.
-As \texttt{multipop} can not remove more elements than there are on the stack, withdrawing $k$ time units to pay for the actual cost will never make the account become negative.
+As \texttt{multipop} cannot remove more elements than there are on the stack, withdrawing $k$ time units to pay for the actual cost will never make the account become negative.
 Because withdrawing from the bank account is enough to pay for the complete actual cost of \texttt{multipop} regardless of the choice of $k$, its \textit{amortized complexity} is in \O{1}.
@@
 
-Usually LiquidHaskell is used to add more precise pre- and postconditions to functions with the help of refinement types~\cite{refinement_types}.
+Usually LiquidHaskell is used to add pre- and postconditions to functions with the help of refinement types~\cite{refinement_types}.
 The refinements are additional constraints on top of normal Haskell data types.
@@
 
+Refinements can use equality (written as \texttt{==}) and comparison without the need for the \texttt{Eq} or \texttt{Ord} typeclasses.
+The reason is that in refinments these operators are translated to the builtin operators of the SMT theory and thus cannot be customized.
 To use functions in the refinements, one has to \textit{reflect}~\cite{reflection} them to the logic level.
@@
 Other functions still can be used in refinements when annotated with \texttt{reflect}, but the SMT solver is not able to check those refinements automatically.
-The user always has to provide the proofs themselves using LiquidHaskell's ability to do complex equational reasoning~\cite{tpfa} that is checked by a machine.
+The user always has to provide the proofs themselves using LiquidHaskell's ability to do complex equational machine-checked reasoning~\cite{tpfa}.
 
@@
 We can bring other facts in scope using the \texttt{?} operator.
+The angle brackets used in its refinement type are used to explicity quantify over two refinements.
+This will make the facts "visible" in the type signature and thus make them available to the SMT solver.
 A proof can be finished by the \texttt{***} operator that casts a chain of equalities to the proof type (type slightly simplified here).
@@
   === 1 + phi (Elem x s) - phi s
--- Use definition of phi (Elem _ _)
+-- Use definition of phi (Elem _ s)
   === 1 + (1 + phi s) - phi s
@@
 
-While this stack data structure might look contrived, it is basically the same as Okasaki's queues~\cite{okasaki}.
-A queue is one front and one back list which is in reverse order.
+While this stack data structure might look contrived, it is basically half of Okasaki's queues~\cite{okasaki}.
+A queue is one front and one back list (or stack) where the back list is in reverse order.
 This allows us to remove elements from the front list and add elements to the back list in \O{1}.
-When the front list is empty, the back list is rotated and used as the new front list, an operation which is basically the same as our \texttt{multipop} here.
+When the front list is empty, the back list is rotated and used as the new front list, an operation which is similar to \texttt{multipop} here.
 
@@
 
-We now move on from ordinary queues to priority queues.
-In functional languages the latter are usually implemented as some kind of (min) heap.
+In functional languages priority queues are usually implemented as some kind of (min) heap.
 A binomial heap is a forest of binomial trees where there exists at most one tree of every rank.
@@
 
-data Heap a = FEnd -- empty forest
-  | F0 (Heap a) -- no tree at this position
-  | F1 (Tree a) (Heap a)
+data Forest a = FEnd -- empty forest
+  | F0 (Forest a) -- no tree at this position
+  | F1 (Tree a) (Forest a)
+type Heap a = Forest a
 
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[escapechar=!]
 {-@ reflect phi @-}
@@
 {-@ reflect insertT @-}
-{-@ insertT :: Ord a =>
+{-@ insertT :: Ord a !\color{codegreen}{=>}!
   Tree a -> Heap a -> { x:Int | x >= 1 } @-}
@@
 {-@ automatic-instances insertTreeP @-}
-{-@ insertTreeP :: t:_ -> f:_ ->
+{-@ insertTreeP :: t:Tree a -> f:Heap a ->
   { insertT t f + pot (insertTree t f)
@@
 {-@ reflect toList @-}
-{-@ toList :: Some a ->
+{-@ toList :: Digit a ->
   { xs:[a] | len xs >= 1 && len xs <= 3 } @-}
-toList :: Some a -> [a]
+toList :: Digit a -> [a]
 toList (One x) = [x]
@@
 
+\subsection{Comparison with Claessen's formalization}
+
+For their formalization, Claessen used HipSpec~\cite{hipspec}, a tool to translate Haskell to first order logic.
+However we were not able to directly compare these proofs to ours as the tool has not seen any updates in the last 5 years and the last supported GHC version is 7.8.
+LiquidHaskell on the other hand has seen some industry adoption and has kept up to date with new GHC versions (version 9.8 at the time of writing).
+
+Aside from these technical limitations, they only formalized the comparatively easy \texttt{cons} and \texttt{snoc} complexities and leave out the much more interesting logarithmic complexity of append.
+
 \section{Conclusion}\label{sec:conclusion}
@@
 
-On the other hand, LiquidHaskell's explicit equational reasoning is easier to follow for a new learner than opaque proof scripts written in a tactic-based theorem prover like Coq or Lean.
+On the other hand, LiquidHaskell's explicit equational reasoning is easier to follow for a new learner than opaque proof scripts written in a tactic-based theorem prover like Coq or Lean\footnote{Lean does support equational reasoning via the \texttt{calc} environment, however most proofs are done using tactic scripts}.
 While Isabelle/HOL also features equational reasoning with Isar~\cite{isar}, it does not support polymorphic recursion which makes the definition of our third case study a lot harder.
@@
 This would allow to expand the complexity results to a shared setting where reuse of intermediate results is allowed.
-It might be possible to adapt Yao et al.'s work on complexity for lazy programs~\cite{forking_paths}, but LiquidHaskell's lack of quantifiers will make this challenging.
+It might be possible to adapt \citet{forking_paths}'s or \citet{lazyfunction}'s works on complexity for lazy programs, but LiquidHaskell's lack of quantifiers will make this challenging.
 

### 2024-07-18 paper/main.tex
Add more review comments

@@
 Usually formal proofs are done in an external theorem prover (compare e.g.~\cite{complexity_coq} or~\cite{complexity_isabelle}), requiring an explicit translation step from the production language to the language the theorem prover uses.
-This translation step has a few disadvantages.
-First, the semantics of the target language could differ from the semantics of the production language.
-This is especially true for Haskell which is lazy, while most theorem provers have a strict language semantics.
-Second, such a translation step could introduce subtle differences or bugs, either by accident when doing a manual translation or through a bug when an automated translation tool is used.
-Lastly, the target language may not support certain features present in the source language (e.g.
+Such a translation step could introduce subtle differences or bugs if this translation is done manually.
+Additionally, the target language may not support certain features present in the source language (e.g.
 polymorphic recursion in data types, see section~\ref{sec:fingertree}).
@@
 With a handful of proof combinators defined by LiquidHaskell, it is possible to use Haskell directly as a theorem prover, enabling intuitive equational reasoning within the language~\cite{tpfa}.
-This avoids any translation steps and makes it easy to keep implementation and its respective proof in sync, as any failure to do so will result in a compiler error.
+This makes it easy to keep the implementation and its respective proof in sync, as any failure to do so will result in a compiler error.
+LiquidHaskell errors are also integrated with the normal Haskell IDE tooling, so the user can see proof failures directly in their editing environment.
 Another benefit is that the reader or author of a proof does not need to learn an additional language, the production language is the same as the proof language.
@@
 Some data structures can exploit laziness to have the same (good) amortized complexity also in a shared setting, however formally verifying this would require explicit reasoning about the lazy semantics of the language and thus massively complicate the analysis.
+Note that in a non-shared setting the amortized complexity bounds apply regardless of whether the language is lazy or strict.
+Laziness may cause the data structure to do less work -- and thus have a better complexity -- but as an upper bound the amortized complexity is still valid.
 
@@
     + phi (multipop k xs) - phi xs
--- Use definition of multipopT for recursive case
+-- Use definition of multipopT n (Elem _ s)
   === (1 + multipopT (k - 1) s)
     + phi (snd (multipop k xs)) - phi xs
--- Use definition of multipop for recursive case
+-- Use definition of multipop n (Elem _ s)
   === (1 + multipopT (k - 1) s)
     + phi (snd (multipop (k - 1) s)) - phi xs
--- Use definition of phi (Elem _ _)
+-- Use definition of phi (Elem _ s)
   === (1 + multipopT (k - 1) s)
-    + phi (snd (multipop (k - 1) s)) - (1 + phi s)
+    + phi (snd (multipop (k - 1) s))
+    - (1 + phi s)
 -- Remove parentheses
@@
 {-@ insertTreeP :: t:Tree a -> f:Heap a ->
-  { insertT t f + pot (insertTree t f)
-    - pot f <= 2 }
+  { insertT t f + phi (insertTree t f)
+    - phi f <= 2 }
 @-}
@@
 phi Unit{} = 0
-phi (More pr m sf) = danger pr + phi m + danger sf
+phi (More pr m sf) =
+  danger pr + phi m + danger sf
 
@@
 {-@ foldlTSnoc :: q:Seq a -> as:[a] ->
-  { foldlT snoc snocT q as + pot (foldl snoc q as)
+  { foldlT snoc snocT q as
+    + pot (foldl snoc q as)
     - pot q <= 3 * len as + 1 } @-}
@@
     - pot q
-  === snocT q x + foldlT snoc snocT (snoc q x) xs
+  === snocT q x
+      + foldlT snoc snocT (snoc q x) xs
       + pot (foldl snoc (snoc q x) xs) - pot q
@@
 For the recursive case we also need some small helper facts about the logarithm such as monotonicity.
+We also use a \texttt{where} binding to factor out common code and make the proof more readable.
 
@@
 log2Mono 1 m = log2 1 <= log2 m *** QED
-log2Mono n m =
-  log2 n <= log2 m
-  === 1 + log2 (n `div` 2) <= 1 + log2 (m `div` 2)
+log2Mono n m = log2 n <= log2 m
+  === 1 + log2 (n `div` 2)
+    <= 1 + log2 (m `div` 2)
   === log2 (n `div` 2) <= log2 (m `div` 2)
@@
     (toTuples (toList v1 ++ as ++ toList u2)) q2
-  =<= log2 (max (length (seqToList q1)
-        + length (seqToList q2)) 2) + 15
+  =<= log2 (max n 2) + 15
 -- move (+1) into the logarithm
-  ? divCancel (max (length (seqToList q1)
-      + length (seqToList q2)) 2)
-  === 1 + log2 (2 * max (length (seqToList q1)
-      + length (seqToList q2)) 2 `div` 2) + 14
-  === log2 (2 * max (length (seqToList q1)
-      + length (seqToList q2)) 2) + 14
--- simplify
-  ? log2Mono (2 * max (length (seqToList q1)
-      + length (seqToList q2)) 2)
-    (4 + 2 * length (seqToList q1)
-      + 2 * length (seqToList q2))
-  =<= log2 (4 + 2 * length (seqToList q1)
-      + 2 * length (seqToList q2)) + 14
-  ? log2Mono (4 + 2 * length (seqToList q1)
-    + 2 * length (seqToList q2))
-      (4 + length (tuplesToList (seqToList q1))
-        + length (tuplesToList (seqToList q2)))
-  =<= log2 (4 + length (tuplesToList (seqToList q1))
-      + length (tuplesToList (seqToList q2))) + 14
-  ? log2Mono (4 + length (tuplesToList (seqToList q1))
-    + length (tuplesToList (seqToList q2))) 
-      (max (length (toList u1 ++ tuplesToList
-          (seqToList q1) ++ toList v1)
-        + length (toList u2 ++ tuplesToList
-            (seqToList q2) ++ toList v2)) 2)
-  =<= log2 (max (length (toList u1 ++ tuplesToList
-        (seqToList q1) ++ toList v1)
-      + length (toList u2 ++ tuplesToList
-        (seqToList q2) ++ toList v2)) 2) + 14
+  ? divCancel (max n 2)
+  === 1 + log2 (2 * max n 2 `div` 2) + 14
+  === log2 (2 * max n 2) + 14
+
+-- simplify using monotonicity of log2
+  ? log2Mono (2 * max n 2) (4 + 2 * n1 + 2 * n2)
+  =<= log2 (4 + 2 * n1 + 2 * n2) + 14
+  ? log2Mono (4 + 2 * n1 + 2 * n2) (4 + n1' + n2')
+  =<= log2 (4 + n1' + n2') + 14
+  ? log2Mono (4 + n1' + n2')
+    (max (length (toList u1
+      ++ tuplesToList (seqToList q1) ++ toList v1
+    ) + length (toList u2
+      ++ tuplesToList (seqToList q2) ++ toList v2
+    )) 2)
+  =<= log2 (max (length (toList u1
+      ++ tuplesToList (seqToList q1) ++ toList v1)
+    + length (toList u2
+      ++ tuplesToList (seqToList q2) ++ toList v2)
+    ) 2) + 14
   === log2 (max (length (seqToList qq1)
@@
   *** QED
+  where
+    n1 = length (seqToList q1)
+    n2 = length (seqToList q2)
+    n = n1 + n2
+    n1' = length (tuplesToList (seqToList q1))
+    n2' = length (tuplesToList (seqToList q2))
 \end{lstlisting}
@@
 snoc (More u q (One x)) y = More u q (Two x y)
-snoc (More u q (Two x y)) z = More u q (Three x y z)
+snoc (More u q (Two x y)) z =
+  More u q (Three x y z)
 snoc (More u q (Three x y z)) w =

### 2024-07-18 code/src/Seq2.hs

@@ -241,7 +241,6 @@ max a b = if a >= b then a else b
 divCancel :: Int -> Proof
 divCancel _ = ()
 
-{-
 {-@ glueAmortized :: q1:Seq a -> { as:[a] | len as <= 3 } -> q2:Seq a -> { glueT q1 as q2 + pot (glue q1 as q2) - pot q1 - pot q2 <= log2 (max (len (seqToList q1) + len (seqToList q2)) 2) + 14 } @-}
 glueAmortized :: Seq a -> [a] -> Seq a -> Proof
 glueAmortized q1@Nil as q2 =
@@ -282,21 +281,24 @@ glueAmortized qq1@(More u1 q1 v1) as qq2@(More u2 q2 v2) =
   === 1 + glueT q1 (toTuples (toList v1 ++ as ++ toList u2)) q2
       + pot (glue q1 (toTuples (toList v1 ++ as ++ toList u2)) q2)
       - pot q1 - dang v1 - dang u2 - pot q2
-  ? glueAmortized q1 (toTuples (toList v1 ++ as ++ toList u2)) q2
-  =<= log2 (max (length (seqToList q1) + length (seqToList q2)) 2) + 15
-  ? divCancel (max (length (seqToList q1) + length (seqToList q2)) 2)
-  === 1 + log2 (2 * max (length (seqToList q1) + length (seqToList q2)) 2 `div` 2) + 14
-  === log2 (2 * max (length (seqToList q1) + length (seqToList q2)) 2) + 14
-  ? log2Mono (2 * max (length (seqToList q1) + length (seqToList q2)) 2) (4 + 2 * length (seqToList q1) + 2 * length (seqToList q2))
-  =<= log2 (4 + 2 * length (seqToList q1) + 2 * length (seqToList q2)) + 14
-  ? log2Mono (4 + 2 * length (seqToList q1) + 2 * length (seqToList q2))
-      (4 + length (tuplesToList (seqToList q1)) + length (tuplesToList (seqToList q2)))
-  =<= log2 (4 + length (tuplesToList (seqToList q1)) + length (tuplesToList (seqToList q2))) + 14
-  ? log2Mono (4 + length (tuplesToList (seqToList q1)) + length (tuplesToList (seqToList q2))) 
+    ? glueAmortized q1 (toTuples (toList v1 ++ as ++ toList u2)) q2
+  =<= log2 (max n 2) + 15 ? divCancel (max n 2)
[…]

### 2024-07-18 paper/main.tex
Fix last typesetting issues

@@
   =<= log2 (4 + 2 * n1 + 2 * n2) + 14
-  ? log2Mono (4 + 2 * n1 + 2 * n2) (4 + n1' + n2')
-  =<= log2 (4 + n1' + n2') + 14
-  ? log2Mono (4 + n1' + n2')
+  ? log2Mono (4 + 2 * n1 + 2 * n2) (4 + m1 + m2)
+  =<= log2 (4 + m1 + m2) + 14
+  ? log2Mono (4 + m1 + m2)
     (max (length (toList u1
@@
   =<= log2 (max (length (toList u1
-      ++ tuplesToList (seqToList q1) ++ toList v1)
-    + length (toList u2
-      ++ tuplesToList (seqToList q2) ++ toList v2)
-    ) 2) + 14
+      ++ tuplesToList (seqToList q1) ++ toList v1
+    ) + length (toList u2
+      ++ tuplesToList (seqToList q2) ++ toList v2
+    )) 2) + 14
   === log2 (max (length (seqToList qq1)
@@
     n = n1 + n2
-    n1' = length (tuplesToList (seqToList q1))
-    n2' = length (tuplesToList (seqToList q2))
+    m1 = length (tuplesToList (seqToList q1))
+    m2 = length (tuplesToList (seqToList q2))
 \end{lstlisting}
@@
 insert t (F0 f) = F1 t f
-insert t (F1 t' f) = F0 (insert (mergeTree t t') f)
-\end{lstlisting}
+insert t (F1 t' f) =
+  F0 (insert (mergeTree t t') f)
+\end{lstlisting}
+
+\balance
 
